### 2014-07-20 aps4p5/drafts/draft_140719.tex
added draft

%\documentclass[prd, twocolumn, nofootinbib, floatfix]{revtex4}
%\documentclass[useAMS,usenatbib]{mn2e}
\documentclass[useAMS,usenatbib]{aastex}

\usepackage{amsmath}
\usepackage{amsbsy}

\topmargin0.0cm
\textheight8.5in

\input epsf
\usepackage{amsmath,amssymb,subfigure}
\usepackage{graphicx}
\usepackage{epsfig}
\usepackage{color}
%\usepackage{ulem}
%\usepackage{epstopdf}

\renewcommand{\topfraction}{0.95}
\renewcommand{\bottomfraction}{0.95}

\newcommand{\be}{\begin{equation}}
\newcommand{\ee}{\end{equation}}
\newcommand{\gscr}{{\mathcal{G}}}
\newcommand{\vscr}{{\mathcal{V}}}

\newcommand{\highlight}[1]{\textcolor{blue}{#1}}

\newcommand{\beq}{\begin{equation}}
\newcommand{\eeq}{\end{equation}}
\newcommand{\beqa}{\begin{eqnarray}}
\newcommand{\eeqa}{\end{eqnarray}}
\newcommand{\om}{\Omega_m}
\newcommand{\omw}{\Omega_w}
\newcommand{\lam}{\Lambda} 
\newcommand{\dcc}{\Delta_m\chi^2} 
\newcommand{\gs}{\gtrsim} 
\newcommand{\ls}{\mathrel{\raise0.27ex\hbox{$<$}\kern-0.70em \lower0.71ex\hbox{{
$\scriptstyle \sim$}}}}
\newcommand{\like}{\chi^2}
\newcommand{\APS}{APS }

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\title
{An Efficient Parameter Space Search as an Alternative to Markov Chain Monte Carlo} 
\author
{Scott F.\ Daniel$^1$, Andrew J.\ Connolly$^1$, and Jeff Schneider$^2$\\
$^1$Department of Astronomy, University of Washington, Seattle, WA\\ 
$^2$Machine Learning Department, Carnegie Mellon University, Pittsburgh, PA
}
\begin{document}

%\pagerange{\pageref{firstpage}--\pageref{lastpage}}

\label{firstpage}

\date{\today}

\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract} 
We consider the problem of inferring constraints on a high-dimensional parameter
space with a computationally expensive likelihood function.  
Markov chain Monte Carlo (MCMC) methods offer significant improvements in
efficiency over grid-based searches and are easy to implement in a wide range of
cases.  However, MCMCs offer 
few guarantees that all of the interesting regions of 
parameter space are explored.
We propose a
machine learning algorithm that improves upon the performance of MCMC
by intelligently targeting likelihood evaluations so as to quickly and
accurately characterize the likelihood surface in both low- and high-likelihood
regions.  
We compare our algorithm to MCMC on toy examples and the 7-year WMAP
cosmic microwave background data release.  
Our algorithm finds comparable parameter
constraints to MCMC comparably quickly and with greater
certainty that all of the interesting regions of parameter space have been
explored.
\end{abstract}

\section{Introduction}
\label{sec:intro}

We consider the following problem.  A researcher wants to constrain a theory
described by some number $N_p$ of tunable parameters, which we will represent by
the $N_p$-dimensional vector $\vec{\theta}$.  The available data measure some
function $f$ which is an implicit function of $\vec{\theta}$.  The researcher
is able to calculate some statistic (e.g. $\chi^2$) quantifying the fit of a 
model $f(\vec{\theta})$ to the data and wishes to find a confidence limit
defined by some rule, e.g $\chi^2\le\chi^2_\text{lim}$.  This is a general
phrasing of a Frequentist confidence limit as discussed in Appendix
\ref{sec:frequentism}.  We discuss alternatives for determining
$\chi^2_\text{lim}$ in Section \ref{sec:chi} below.

If the calculation going from $\vec{\theta}\rightarrow\chi^2$ is rapid, this
problem is trivial.  One generates a fine grid of points in the 
$N_p$-dimensional parameter space, evaluates $\chi^2$ at each of these points, 
and defines the confidence limit as only those points which satisfy the rule.  
Often, however, the calcuation $\vec{\theta}\rightarrow\chi^2$ is not rapid, or $N_p$
is large enough that such a grid-based method will take a prohibitively long
time.  In that case, one must make intelligent choices regarding which points in
parameter space are actually interesting enough to warrant
 $\vec{\theta}\rightarrow\chi^2$ evalutation.

Markov Chain Monte Carlo (MCMC) methods propose to find the desired
limits by drawing random samples from the parameter space which are drawn
according to a probability density (the Bayesian posterior) which can be
interpreted as the probability that some value of $\vec{\theta}$ represents the
true value of $\vec{\theta}$ realized by the physical process underlying
the data.  
With enough of these samples, one can integrate this
probability density and find regions which contain some set amount of the total
probability.  In this case, ``credible limits'' are phrased as target values
for the probability integral, i.e. 
$0.68$ for ``1-$\sigma$'' limits, $0.95$ for ``2-$\sigma$'' limits
and so on \cite{mcmc}.

We propose an alternative solution, using methods from machine learning
(to wit, Gaussian processes, though the algorithm admits many possible drivers; see
Section \ref{sec:gp})
to find the desired confidence limit by directly searching the
parameter space of $\vec{\theta}$ for points which lie on the boundary of the
desired confidence limit $\chi^2\le\chi^2_\text{lim}$.  
We refer to this algorithm as Active Parameter Searching (APS).
We present the algorithm in detail in Sections \ref{sec:algorithm} and
\ref{sec:gp}.
As it searches, APS uses
the knowledge it has already learned about $\vec{\theta}\rightarrow\chi^2$ 
to improve its search according to a metric that rewards the
algorithm both for finding points on the confidence limit and for sampling
points from previously unexplored regions of parameter space.  We find that this
behavior makes APS more robust against multi-modal $\chi^2$ functions (see
Section \ref{sec:toy}) without
sacrificing the speed of convergence 
associated with traditional MCMC methods (see Section \ref{sec:wmap}).
We show that APS is capable of yielding both Frequentist confidence
limits (which we define in Appendix \ref{sec:frequentism}) as well as more traditional
Bayesian credible limits on the space of $\vec{\theta}$.

\section{The APS Algorithm}
\label{sec:algorithm}

The APS algorithm was originally presented by Bryan (2007) as

\begin{itemize}
\item(1A) Generate some initial number $N_s$ of randomly-distributed points in 
parameter space.  Find the values of $\chi^2$ corresponding to these points.
\\
\item(2A) Generate some number $N_c$ of candidate points.  Use the points
already sampled by APS to predict the value of $\chi^2$ at each of these points.
 We will signify this prediction by $\mu$.  Calculate some value $\sigma$
 quantifying confidence in $\mu$ as a prediction of $\chi^2$.  Low values of
 $\sigma$ should correspond to confident predictions.
 \\
\item(3A) Choose the candidate point which maximizes the statistic
\begin{equation}
\label{eq:sstat}
S=\sigma-|\mu-\chi^2_\text{lim}|
\end{equation}
and evaluate $\vec{\theta}\rightarrow\chi^2$ at that point.  Add this point and
its $\chi^2$ value to the list of sampled points begun in (1A). 
Return to (2A) and repeat until convergence.
\\
\end{itemize}
The confidence limit reported by APS is the list of all points found
for which $\chi^2\le\chi^2_\text{lim}$.  
These points, when plotted in one- or two-dimensional slices of parameter
space, ought to sketch out a region which contains the true value of $\vec{\theta}$
with the desired confidence (as defined in Appendix \ref{sec:frequentism}).
This is in contrast to the credible
limits of MCMC, which represent an integral over the sampled points containing some
fraction of the total posterior probability.

The $|\mu-\chi^2_\text{lim}|$ term in equation (\ref{eq:sstat}) provides an
incentive for APS to choose points that it believes will lie on the boundary of
the confidence limit.  
The $\sigma$ term provides an incentive for APS to explore unknown regions of
parameter space, making the algorithm robust against multimodal $\chi^2$
functions.

Bryan {\it et al.} (2007) show that this algorithm is robust against multi-modal
likelihood functions, using it to identify a disjoint region of high likelihood in
cosmological parameter space as constrained by the 1-year WMAP CMB anisotropy power
spectrum.  This is the principal advantage to selecting points for evaluation according
to the statistic (\ref{eq:sstat}).  Unfortunately, it is possible that the focus on exploring
unknown regions of parameter space might slow down convergence in the case of uni-modal
likelihood functions.  Therefore, we propose the following modifications to the algorithm
to ensure that it adequately characterizes known regions of high likelihood while
simultaneously searching for new regions to explore.

\subsubsection{Characterization of known high likelihood regions}
\label{sec:focus}

As the algorithm searches, it will keep track of a list of all of the discovered points
which are believed to be local minima of $\chi^2$ (how it finds these will be described below
in Section \ref{sec:simplex}).  After each cycle of the main algorithm (steps 1A-3A above),
our version of APS attempts ``focused searches'' about these local minima.

For each local minimum $\vec{M}$, APS performs the following:
\begin{itemize}
\item (1B) If fewer than $N_p$ focused searches have been proposed around this minimum, randomly
choose a point from a small sphere surrounding $\vec{M}$ in parameter space and 
evaluate $\chi^2$ at that point.  
\\
\item(2B) If the point does not satisfy $\chi^2=\chi^2_\text{lim}$ (which is highly probable), 
use bisection
along the line connecting the proposed point to $\vec{M}$ to find a point which does
satisfy $\chi^2=\chi^2_\text{lim}$, i.e. find a point on that line with
$\chi^2<\chi^2_\text{lim}$ -- often $\vec{M}$ itself -- and a point along that line
with $\chi^2>\chi^2_\text{lim}$ and iteratively step half the distance between them until
arriving at the desired $\chi^2_\text{lim}$ \cite{minuit}.  The points discovered this way with
$\chi^2=\chi^2_\text{lim}$ are stored as ``boundary points'' about $\vec{M}$.  Each local
minimum will have its own list of boundary points.
\\
\item(3B) If $N_p$ boundary points have been found for $\vec{M}$, skip steps (1B) and (2B).
Instead, examine the boundary points associated with $\vec{M}$ and identify the points
corresponding to both the maximum and minimum values of each parameter $\theta_i$.  These
points will be referred to as $\vec{B}^{(j)}$.  There will be $2\times N_p$ of them, one for
the maximum and one for the minimum in each of our $N_p$ parameters.  
\\
\item(4B) For each point
$\vec{B}^{(j)}$, propose some number $N_v$ of new points (we have chosen
$N_v=20$ for the sake of speed) $\vec{d}^{(k)} = \vec{B}^{(j)}+\vec{v}$ where $\vec{v}$ is a
random, small vector in parameter space constrained to be perpendicular to
$\vec{B}^{(j)}-\vec{M}$.  There will now be $2\times N_p \times N_v$ points $\vec{d}^{(k)}$
($N_v$ such points for each of the $2\times N_p$ extremal boundary points $\vec{B}^{(j)}$).
\\
\item(5B) At each point $\vec{d}^{(k)}$, evaluate equation (\ref{eq:sstat}).  Evaluate $\chi^2$
at the point $\vec{d}^{(k)}$ which maximizes $S$. If
$\chi^2 \neq \chi^2_\text{lim}$, bisection is again used and a new boundary point is found.
\end{itemize}

We are indebted to Eric Linder of Lawrence Berkeley National Lab who first suggested to us
both the use of bisection to isolate $\chi^2_\text{lim}$ and the idea of stepping
intentionally along the $\chi^2=\chi^2_\text{lim}$ surface by stepping perpendicularly to the
direction $\vec{B}^{(j)}-\vec{M}$.

To reiterate, our version of APS alternates between performing steps (1A-3A) and steps
(1B-5B).  If there are $N_m$ distinct known local minima in $\chi^2$, or algoritm will perform
$N_m$ iterations of steps (1A-3A), and then perform steps (1B-5B) once on each local minimum.

\subsubsection{Extending the boundary}
\label{sec:unitSphere}

The modification suggested in section \ref{sec:focus} is designed to explore 
$\chi^2=\chi^2_\text{lim}$ surface by starting from its extreme corners and stepping
perpendicularly to its ``radii'' (the vector $\vec{B}^{(j)}-\vec{M}$ would be a radius of
$\chi^2=\chi^2_\text{lim}$ described a sphere).

something about APS wide helping to discover new corners of the surface as well

\subsubsection{Function minimization}
\label{sec:simplex}

\subsection{Determining $\chi^2_\text{lim}$}
\label{sec:chi}

As discussed above, APS draws confidence limits by mapping
iso-$\chi^2$ contours of $\chi^2=\chi^2_\text{lim}$ in parameter space.
Thus, the user is presented with the problem of determining an
appropriate $\chi^2_\text{lim}$.  APS admits two possible solutions.
The user can fiat a definite value of $\chi^2_\text{lim}$
based on the characteristics of the data at hand,
or $\chi^2_\text{lim}$ can be set adaptively, based on APS'
knowledge of $\vec{\theta}\rightarrow\chi^2$.

The first possibility, a user-defined $\chi^2_\text{lim}$ is based
on the theory of Frequentist confidence intervals, which we briefly discuss
here and consider in much greater detail in Appendix \ref{sec:frequentism}.
If a data set is comprised of $N_d$ Gaussian-distributed data points
with a known covariance structure, then 
\begin{equation}
\chi^2\equiv\sum_{i,j}^{N_d}(d_i-f_i(\vec{\theta}))C^{-1}_{i,j}
(d_j-f_j(\vec{\theta}))
\end{equation}
will be distributed according to a $\chi^2$-distribution with $N_d$ degrees
of freedom.  This distribution can be integrated to determine the value
of $\chi^2$ containing $(1-\alpha)\%$ of the total probability.  This limiting
value will be the user-defined $\chi^2_\text{lim}$.

If the user does not feel comfortable settig $\chi^2_\text{lim}$
a priori, APS $\chi^2_\text{lim}$ can learn $\chi^2_\text{lim}$ by
defining it as $\chi^2_\text{min}+\Delta\chi^2$.  Here, $\chi^2_\text{min}$
is the minimum value of $\chi^2$ discovered by APS.  $\Delta \chi^2$ is set
based on the properties of the $N_p$-dimensional parameter space being explored.
To wit, Wilks (1938) proves that $\Delta\chi^2$ for an $N_p$-dimensional
parameter space will be distributed according to a $\chi^2$-distribution
with $N_p$ degrees of freedom.  One can perform the same integral discussed
above and set $\Delta\chi^2$ to that value which encloses the desired confidence
limit.  This is the well-known Likelihood Ratio Test \cite{np}.
In this case, the user is relying on APS to eventually find the true
$\chi^2_\text{min}$ of the likelihood surface being explored.  We show in
Section \ref{sec:wmap} that, with the inclusion of the gradient descent search
discussed above, this is a safe assumption to make.

\subsection{User-specified Parameters of APS}
\label{sec:user}

We make code implementing APS for an arbitrary $\vec{\theta}\rightarrow\chi^2$
function available at \verb|https://github.com/uwssg/APS/|.  
Below, we present a list of user-specified parameters 
required to run the algorithm.  Documentation is also provided with
the code.

\begin{itemize}
\item $N_s$ -- the number of randomly selected points selected in step (1A).
\\

\item $\chi^2_\text{lim}$ -- the target threshold value of $\chi^2$.  Alternatively,
the user can specify
\\
\item $\Delta \chi^2$ -- a threshold such that APS sets
$\chi^2_\text{lim}=\chi^2_\text{min}+\Delta\chi^2$ where $\chi^2_\text{min}$ is the
minimum value of $\chi^2$ discovered by APS at any given time.  This is the framework
adopted in this paper.
\\

\item $N_c$ -- the number of candidate points considered in setp (2A).  Recall
that the code will only evaluate $\vec{\theta}\rightarrow\chi^2$ on the
candidate point which maximizes equation (\ref{eq:sstat}).
\\

\item $N_g$ -- in order to speed up step (2A), the code
as presented only uses the $N_g$ nearest neighbors (in parameter space) of
each candidate point as inputs to prediction $\mu(\vec{\theta})$.
\\

\item $R_g$ -- (\verb|grat| in the code) the ratio determining whether a point
will be considered a candidate for initiating a gradient descent search. 
Gradient descent will only be considered for points for which
$$\chi^2-\chi^2_\text{lim}\le R_g\times \chi^2_\text{lim}$$
\\

\item $N_w$ -- the maximum number of gradient descent searches that the code
will allow to go on in parallel with the ``usual'' APS search at one time.  A
gradient descent search is terminated once it has either
found a point which satisfies $\chi^2\le\chi^2_\text{lim}$ or 
has reduced its step
size such that it is effectively no longer moving through parameter space.  
If all $N_w$ gradient descent slots are already full when another gradient descent
candidate is found, the code will add the new
gradient descent only if its present value of $\chi^2$ is closer to
$\chi^2_\text{lim}$ than the farthest active gradient descent.  In that case,
the farthest active gradient descent will be terminated.
\\

\item $\{\text{min}_i,\text{max}_i\}$ -- 
the minimum and maximum values of the $N_p$ parameters over which you want APS
to search.
\end{itemize}

\section{Gaussian Processes}
\label{sec:gp}

Step (2A) of \APS requires that a prediction $\mu$ be made regarding the value
of $\chi^2$ at points as yet unsampled by the algorithm.  An uncertainty $\sigma$
must also be assigned to this prediction.  APS writ large is agnostic regarding
the method used to find $\mu$ and $\sigma$.  We choose to use a Gaussian process.

Gaussian processes take noisy, sparse, measurements of an unknown function (in our
case, $\vec{\theta}\rightarrow\chi^2$ at
the points in parameter space already sampled by APS) and use them to
make inferences about unmeasured values of the function by assuming that the
function is a random process.  Gaussian processes have already found use in physics
and astronomy constructing non-parametric models of the cosmic expansion
\cite{ericgp}, interpolating the point spread function across 
astronomical images
\cite{psf}, and interpolating models of the non-linear matter
power spectrum from N-body simulations \cite{habib}.
They are useful in the current context because, not only do they
give robust predictions $\mu$ for the unmeasured values of the
function being modeled, but they also
provide well-motivated prediction uncertainties $\sigma$, a requirement of
equation (\ref{eq:sstat}).  We present below an introduction to the
formalism of Gaussian processes drawn heavily
from Chapter 2 and Appendix A 
of \cite{gp}.

Gaussian processes use the sampled data $\{\vec{\theta}^{(i)},\chi^{2,(i)}\}$, 
where $i$ indexes over all of the sampled points, to predict
$\mu=\chi^{2,(q)}$ at the candidate point 
$\vec{\theta}^{(q)}$ by assuming that the function
$\vec{\theta}\rightarrow\chi^2$ represents a sample drawn from a random process 
distributed
across parameter space.  At each point in parameter space, 
$\chi^2$ is assumed to be
distributed according to a normal distribution with mean 
$\bar{\chi}^2$
and variance dictated by the covariance function 
$C_{ij}(\vec{\theta}^{(i)},\vec{\theta}^{(j)})$.
If we have $N_d$ measurements of $\chi^2$, then we assume
they are distributed according to
\begin{equation}
\label{eq:likelihood}
P(\vec{\chi}^2)=
\frac{\exp\bigg[-\frac{1}{2}K^{-1}\sum_{i,j}^{N_d}
(\chi^{2,(i)}-\bar{\chi}^2)C^{-1}_{ij}
(\chi^{2,(j)}-\bar{\chi}^2)\bigg]}{\sqrt{2\pi}^{N_d}\text{det}|KC|^{N_d/2}}
\end{equation}
$C_{ij}$ is a function assumed by the user encoding how variations in
$\chi^2$ at one point in parameter space affect variations in 
$\chi^2$ at other
points in parameter space.  $K$ is a parameter controlling the
normalization of $C_{ij}$.
The diagonal elements $C_{ii}=1+\sigma^2_{ii}$ where 
$\sigma^2_{ii}$ is the intrinsic variance in the value
of $\chi^2$ at a single point $\vec{\theta}^{(i)}$.  
Rasmussen and Williams (2006) treat 
the special case
$\bar{\chi}^2=0$ and find (see their equations 2.19, 2.25 and 2.26)
\begin{eqnarray}
\mu&=&\sum_{i,j} C_{qi}C^{-1}_{ij}\chi^{2,(j)}\nonumber\\
\sigma^2&=&C_{qq}-\sum_{i,j}C_{qi}C^{-1}_{ij}C_{jq}\label{sig}
\end{eqnarray}
where the sums over $i$ and $j$ are sums over the sampled points 
in parameter space.  $C_{iq}$ relates the sampled point $i$ to the candidate
point $q$.
We do not wish to assume that the mean value of $\chi^2$ is zero everywhere.
Therefore, we modify the equation for $\mu$ to give
\begin{equation}
\label{mu}
\mu=\bar{\chi}^2+\sum_{i,j} C_{qi}C^{-1}_{ij}
(\chi^{2,(j)}-\bar{\chi}^2)
\end{equation}
where $\bar{\chi}^2$ is the algebraic mean of the 
sampled $\chi^{2,(i)}$.
Note the similarity to a multi-dimensional Taylor series expansion with the
covariance matrix playing the role of the derivatives.
Equation (\ref{sig}) differs from equation (6) in \cite{bryan} because
they used the semivariance
$$\gamma_{ij}=\text{var}[\chi^2(\vec{\theta}^{(i)})
-\chi^2(\vec{\theta}^{(j)})]$$
in place of the covariance $C_{ij}$.  In practice, the two assumptions result in
equivalently valid $\mu$ and $\sigma$.

The form of the covariance function 
$C_{ij}(\vec{\theta}^{(i)},\vec{\theta}^{(j)})$ must be
assumed.  One possibility, 
taken from equation 4.9 of Rasmussen and Williams (2006) is
\begin{equation}
\label{eq:covraw}
C_{ij}=\exp\left[-\frac{1}{2}D^2_{ij}\right]
\end{equation}
where $D_{ij}$ is the normalized distance in parameter space between the points 
$\vec{\theta}^{(i)}$
and $\vec{\theta}^{(j)}$.
\begin{equation}
D^2_{ij}\equiv\sum_n^{N_p}\left(\frac{\theta^{(i)}_n-\theta^{(j)}_n}{\text{max}_n-\text{min}_n}\right)^2
\end{equation}
We discuss the normalization $K$ below.
The exponential form of $C_{ij}$ quantifies the assumption 
that distant points should
not be very correlated.  We will use this covariogram in
our tests in Section \ref{sec:toy} and \ref{sec:wmap}.

Another possibility (equation 4.29 of Rasmussen and Williams) is
\begin{eqnarray}
C_{ij}=\phantom{\sin
\bigg[\frac{2(1+\tilde{\vec{\theta}}^{(i)}\cdot\tilde{\vec{\theta}}^{(j)})}
{\sqrt{(1+2(1+\tilde{\vec{\theta}}^{(i)}\cdot\tilde{\vec{\theta}}^{(i)}))
(1+2(1+\tilde{\vec{\theta}}^{(j)}\cdot\tilde{\vec{\theta}}^{(j)}))}}\bigg]}
\label{eq:cov_nn}\\
\frac{2}{\pi}\sin^{-1}
\bigg[\frac{2(1+\tilde{\vec{\theta}}^{(i)}\cdot\tilde{\vec{\theta}}^{(j)})}
{\sqrt{(1+2(1+\tilde{\vec{\theta}}^{(i)}\cdot\tilde{\vec{\theta}}^{(i)}))
(1+2(1+\tilde{\vec{\theta}}^{(j)}\cdot\tilde{\vec{\theta}}^{(j)}))}}\bigg]
\nonumber
\end{eqnarray}
corresponding to the covariance function of a neural network with a single
hidden layer.  In this function $\tilde{\vec{\theta}}$ is the vector of
parameters $\vec{\theta}$ recentered and renormalized relative to the span
$(\text{max}_n-\text{min}_n)$ in each dimension.  We also consider this
covariogram in Section \ref{sec:wmap} and find no appreciable difference
in performance when compared to the use of covariogram (\ref{eq:covraw}).

The normalization constant $K$ in equation (\ref{eq:likelihood}) --known as the
``Kriging parameter'' for the geophysicist who pioneered the overall method -- 
also
must be assumed.  Note that $K$ also multiplies the diagonal elements
$C_{ii}$.  Determining the value of $K$ 
is somewhat problematic because, examining equation
(\ref{mu}), one sees that the factors of $K$ and $K^{-1}$
cancel out of
the prediction $\mu$, so that the assumed value of $K$ has no effect on the accuracy
of the prediction.  
If the opposite had been true, one could heuristically set $K$ to
maximize the accuracy of $\mu$.  
Instead, we set $K$ to the value that maximizes the likelihood
of the input data, i.e. the $N_g$ nearest neighbors in parameter space
used to perform the Gaussian Process inference.

Recall from equation (\ref{eq:likelihood}) that we are treating our
input data as though they were samples drawn from a probability distribution.
To maximize the probability of the observed data (i.e. maximizing the value
of $P(\vec{\chi}^2)$) we must set $K$ equal to
\begin{equation}
\label{eq:kriging}
K=\frac{\sum_{i,j}^{N_g}(\chi^{2,(i)}-\bar{\chi}^2)
C^{-1}_{ij}(\chi^{2,(j)}-\bar{\chi}^2)}
{N_g}
\end{equation}
We adopt this assumption throughout this work.

Figure \ref{fig:gp} applies the Gaussian process of equations (\ref{sig}), 
(\ref{mu}), and (\ref{eq:kriging}) with covariogram (\ref{eq:covraw})
to a toy one-dimensional function.  Inspection shows many desirable
behaviors in $\mu$ and $\sigma$.  
As $\theta^{(q)}$ approaches the sampled points $\theta^{(i)}$, $\mu$
approaches $\chi^{2,(i)}$ and $\sigma$ approaches zero.  
Closer to a sampled point,
the Gaussian process knows more about the true behavior of the function. 
Far from the $\theta^{(i)}$, $\sigma$ is larger, and the $S$ statistic in
equation (\ref{eq:sstat}) will induce the \APS
algorithm to examine the true value of $\chi^2$.

\begin{figure}
\includegraphics[scale=0.3]{gp_example_1307.eps}
\caption{
A one-dimensional example of prediction using Gaussian processes.  
The black curve is the function
being considered.  The crosses are the points at which it has been sampled.  The
green curve is the resulting prediction and the red curves represent the 
1- and
2-$\sigma$ uncertainty bounds.  We set $K$ according to 
equation (\ref{eq:kriging}) wth $N_g=5$ for the 5 sampled points.
}
\label{fig:gp}
\end{figure}

\section{A Toy Example}
\label{sec:toy}

We will now test the ability of APS to learn the confidence limits of an
artificial $\chi^2$ function and compare it to that of MCMC driven by the
traditional Metropolis-Hastings algorithm \cite{mcmc,cosmomc}.

The algorithm (1A)-(3A) in Section \ref{sec:algorithm} 
was originally presented and tested against the 1-year data
release from the WMAP satellite in Bryan {\it et al}. (2007).  They found that
the algorithm detected a second locus of low $\chi^2$
that had gone undetected by previous, MCMC-based analyses of the data.  
This second fit to the data has disappeared as the signal-to-noise
ratio of the data has improved (as we will see in Section \ref{sec:wmap}),
therefore, we will use for our test a toy $\chi^2$ function
with multiple locii of low $\chi^2$.  We assume an $N_p=6$ dimensional
$\vec{\theta}$ and a $\chi^2$ that goes as
\begin{equation}
\chi^2=1300+\frac{d_1^2}{2}+\frac{d_2^2}{2}-153\exp[-2d_1^2]-100\exp[-d_2^2]
\label{eq:toychisq}
\end{equation}
where $d_1$ is the Euclidean parameter-space distance from 
$\vec{\theta}=\{3,0,0,0,0,0\}$ and
$d_2$ is the Euclidean parameter-space distance from 
$\vec{\theta}=\{-3,0,0,0,0,0\}$.  
Figure \ref{fig:toychisq} plots this $\chi^2$ as a function of $\theta_0$ with
all other parameters $\theta_i=0$.
This $\chi^2$ function has three features that
make it tricky for MCMC algorithms to handle.  It has two locii of low $\chi^2$
which are disjoint.  Those locii are narrow compared to the full
$N_p$-dimensional extent of the function (we choose to explore
$-10\le\theta_i\le10$).  There is a false minimum in $\chi^2$ near
$\vec{\theta}=\{0,0,0,0,0,0\}$.

\begin{figure}
\includegraphics[scale=0.3]{toychi_plot.eps}
\caption{
A one-dimensional slice of the toy $\chi^2$ function equation
(\ref{eq:toychisq}).  The other five parameters are set to zero.
The red line indicates the value $\chi^2=1281$.
}
\label{fig:toychisq}
\end{figure}

To test the performance of APS versus MCMC exploring equation
(\ref{eq:toychisq}), we run 200 instances of each algorithm on the function.
We run APS including the gradient descent modification introduced at the end of
section \ref{sec:algorithm}.  
Each instantiation of APS is begun with parameters (see Section \ref{sec:user})
\begin{eqnarray}
\Delta\chi^2&=&12.59\nonumber\\
R_g&=&0.1\nonumber\\
N_w&=&50\nonumber\\
N_c&=&250\nonumber\\
N_g&=&15\nonumber
\end{eqnarray}

Each instantiation of MCMC is comprised of four
independent chains, each of which is allowed to sample up to 250,000 points
in the parameter space.  Initially, the chains propose steps randomly
in the cardinal directions of the parameter space.  The steps in each direction
are drawn from a Gaussian 
distribution with standard deviation $2.0$.
After 4,000 steps have been taken, the MCMC approximates the covariance matrix of
the steps taken so far and adjusts its proposal algorithm so that it proposes steps
along the eigen vectors of this matrix, with characteristic step size equal to
the $(2.38)/\sqrt{N_p}$ times the
square root of the corresponding eigen values ($N_p=6$ is the
dimensionality of our parameter space) \cite{gelman}.  
The code reassesses this proposal distribution after 
every subsequent 2,000 steps.

Figure \ref{fig:toyneg} shows how many instantiations
of each algorithm successfully found 
the low $\chi^2$ region centered on $\theta_0=-3$
as a function of the number of
sampled points it took them to do so.  
Figure \ref{fig:toypos} shows the same
for the $\theta_0=3$  minimum.  The solid black curve represents
APS.  The dashed red curve
represents MCMC.  
For the purposes of these Figures, ``finding'' a minimimum means evaluating
$\vec{\theta}\rightarrow\chi^2$ at a point where $\chi^2\le1281$ near either
$\vec{\theta}=\{-3,0,\dots\}$ or $\vec{\theta}=\{3,0,\dots\}$.
Though APS and MCMC find the minimum at $\{-3,0,\dots\}$
(which is wider in parameter space)
with comparable rapidity, MCMC only finds the $\{3,0,\dots\}$
minimum  50\% of the time.  The other 50\% of the time, 
MCMC is complacent having
found the other minimum and calibrates itself to thoroughly exploring
that region of low $\chi^2$.
Because APS selects points based on how interesting they are
likely to be -- i.e. equation (\ref{eq:sstat}) -- APS finds both regions of low
$\chi^2$ 100\% of the time.
We also considered the performance of a uniform random sampling of our parameter
space.  In 1 million samplings, only 9 of the
200 instantiations found the $\{-3,0,\dots\}$ minimum.  
Only 5 instantiations found the $\{3,0,\dots\}$ minimum.

\begin{figure}
\subfigure[]{
\includegraphics[scale=0.3]{udder_found_n.eps}
\label{fig:toyneg}
}
\subfigure[]{
\includegraphics[scale=0.3]{udder_found_p.eps}
\label{fig:toypos}
}
\caption{
We run MCMC and APS on a 6-dimensional parmeter space with a $\chi^2$ function
given by equation (\ref{eq:toychisq}).
The horizontal axis represents the number of points sampled by each algorithm.
The vertical axis represents the number of instantiations of each algorithm that
had found the high likelihood region at $\{-3,0,0,0,0,0\}$ (Figure
\ref{fig:toyneg}) and $\{3,0,0,0,0,0\}$ (Figure \ref{fig:toypos}) in that
many iterations.
}
\label{fig:toy}
\end{figure}

\section{A Real Example: WMAP}
\label{sec:wmap}

Section \ref{sec:toy} demonstrates the robustness of APS against multi-modal
$\chi^2$ functions.  In this section, we demonstrate 
that, on well-behaved likelihood functions, APS' rate of convergence is
comparable to that of MCMC.  
We also show that APS can be used on actual astrophysical
problems by using for our $\chi^2$ function the likelihood function on the
7-year data release of the WMAP CMB satellite \cite{wmap7,wmap7likelihood}.
For simplicity, we consider only the anisotropy
power spectrum of the temperature-temperature correlation function.

The parameter space we explore is the six-dimensional space of the spatially
flat concordance $\Lambda$CDM cosmology:
\begin{itemize}
\item$\Omega_bh^2$ -- the density of baryons in the Universe
\\
\item$\Omega_\text{CDM}h^2$ -- the density of cold dark matter in the Universe
\\
\item$\tau$ -- the optical depth to the surface of last scattering.  $\tau$ cannot be
constrained without polarizationd data, which we do not consider here.  However, it functions
as a nuisance parameter to be marginalized over.  The fact that APS can handle it as well as
MCMC is an important test of APS' feasibility.
\\
\item$h$ -- the Hubble parameter in units of
$100~\text{km}~\text{s}^{-1}~\text{Mpc}^{-1}$
\\
\item$n_s$ -- the spectral index controlling the distribution of scalar density
perturbations in the Universe
\\
\item$\ln[10^{10}A_s]$ -- the amplitude of the initial scalar density
perturbations in the Universe
\end{itemize}
These are the parameters taken by the publicly available MCMC code CosmoMC,
which we use to run our baseline MCMC chains \cite{cosmomc}.  The
$\vec{\theta}\rightarrow\chi^2$ function in this case involves converting these
paramters into the power spectrum of anisotropy $C_\ell$s on the sky and
comparing those $C_\ell$s to the actual measurements registered by the WMAP
satellite.  Using the (very fast) Boltzmann code CAMB to calculate the $C_\ell$s
\cite{camb} takes 1.3 seconds on a 2.5 GHz processor.  
Evaluating the $\chi^2$ using the likelihood code provided by the WMAP team
\cite{wmap7likelihood} takes an additional 2 seconds.  Hence the need
for an algorithm like APS or MCMC to accelerate the search.
We run APS with the following input parameters:
\begin{eqnarray}
\Delta\chi^2&=&12.59 \nonumber\\
&&\text{(according to the likelihood ratio test,}\nonumber\\
&&\text{the
95\% confidence limit for a 6-dimensional}\nonumber\\
&&\text{ parameter space
ought to be }\chi^2_\text{min}+12.59\text{)}\nonumber\\
R_g&=&0.1\nonumber\\
N_w&=&100\nonumber\\
N_c&=&250\nonumber\\
N_g&=&15\nonumber
\end{eqnarray}
We will consider both equations (\ref{eq:covraw}) and (\ref{eq:cov_nn})
as covariograms, though we find that the choice has little effect on our
results.

To gauge the appropriateness of applying APS to this problem, we consider
the two following tests.  In Figure \ref{fig:chimin} we plot the minimum discovered value
of $\chi^2$ yielded by APS and MCMC as a function of the number of samples drawn
by each.  If APS failed to find a suitable value of $\chi^2_\text{min}$, the use of
$\Delta\chi^2$  to find $\chi^2_\text{lim}$ would be questionable.  However, as we can see,
the $\chi^2_\text{min}$ yielded by APS ($1273.3$) 
is within a few of the $\chi^2_\text{min}$ yielded
by MCMC ($1270.7$).  Our use of $\chi^2_\text{lim}=\chi^2_\text{min}+\Delta\chi^2$ is therefore
justified.

The reader will note that
MCMC discovers its $\chi^2_\text{min}$ much more rapidly than APS, however, 
we will see
in Figures \ref{fig:contours_gauss} through \ref{fig:contours_br} that this rapid
descent to $\chi^2_\text{min}$ does not translate into rapid convergence to
the final credible limit.  Because MCMC infers its limits based on the number
of samples drawn from the Bayesian posterior probability, MCMC still requires
many tens (or hundreds) of thousands of points to be sampled after $\chi^2_\text{min}$
has been discovered.  Because APS is designed to locate points at the precise value
$\chi^2=\chi^2_\text{lim}$, it requires much fewer samples after discovering 
$\chi^2_\text{min}$ to infer the correct confidence limit.

The second question we consider is whether or not our assumption that
$\vec{\theta}\rightarrow\chi^2$ can be modeled by a Gaussian Process is 
valid.  In Figure
\ref{fig:goodness} we plot the average fractional error in $\mu$ as a proxy for $\chi^2$ as a
function of the number of steps sampled (note: the vertical axis actually shows the average
fractional error over the preceding 2,000 steps so that the early, imprecise models do not
contaminate the later, precise models).  As APS samples more points, its Gaussian Process
model becomes more accurate and the fractional error converges to a few percent.  Thus our
use of Gaussian Processes to model $\vec{\theta}\rightarrow\chi^2$ is also justified.

\begin{figure}
\includegraphics[scale=0.3]{chimin_plot.eps}
\caption{
The minimum value of $\chi^2$ found by APS and MCMC on the WMAP 7 problem
as a function of 
steps sampled by each.
}
\label{fig:chimin}
\end{figure}

\begin{figure}
\includegraphics[scale=0.3]{goodness_of_fit.eps}
\caption{
The average value of $|\mu-\chi^2|/\chi^2$ as a function of the number of points
sampled by APS in the WMAP 7 problem.  
This is shown as a check on our assumption that
$\vec{\theta}\rightarrow\chi^2$ can be approximated as a Gaussian Process.
As the number of sample points increases, the average error yielded by either
covariogram quickly converges to a few percent in $\chi^2$.
}
\label{fig:goodness}
\end{figure}

We now consider the ultimate results, namely the 95\% confidence limits on
our 6-dimensional cosmological parameter space yielded by APS.
To provide a baseline against which to test APS, we run 4 independent MCMC
chains using CosmoMC.  Like the MCMC code used in Section \ref{sec:toy},
CosmoMC periodically adjusts its proposal density by learning the covariance
matrix of the points it has already sampled.  At each step, it proposes
a new point in parameter space by randomly selecting an eigen vector
of that learned covariance matrix and stepping along it.
We integrate the posterior by taking the output chains, discarding the first
50\% of steps as a burn in period, and thinning the remaining steps so that
we have a set of effectively independent samples drawn from the posterior.
Quantitatively, this is determined by keeping only every $L$th sample after
burn-in, where $L$ is set as the length such that the normalized correlation
\begin{equation}
\frac{\text{Cov}[\theta_n^{(i)},\theta_n^{(i+L)}]}
{\text{Var}[\theta_n]}\le 0.01
\end{equation}
where $n$ is the index over the $N_p$ parameters and $(i)$ is the index
over the samples drawn.

We find the 95\% credible
limits in two-dimensional slices of our parameter space using
a total of 460,000 MCMC steps.  These are the solid
black contours in Figures \ref{fig:contours_gauss} through 
\ref{fig:contours_br} 
below.  They are referred to below as the ``final credible limits''
returned by MCMC.  To gauge
convergence as a function of time, we also plot the 95\% credible limit contours
taken from our set of four MCMC chains after they have sampled
a total of 100,000 points.  
These are the dotted green
contours in Figures \ref{fig:contours_gauss} through \ref{fig:contours_br}.
They are referred to as the ``pre-convergence'' credible limits.
Note that, even though MCMC has discovered its final value of $\chi^2_\text{min}$
by this point (see Figure \ref{fig:chimin}), this rapid convergence in
$\chi^2_\text{min}$ has not translated into convergence for the credible limits
on parameter space.

In the case of APS, the confidence limit is interpreted 
as that set of points found
which satisfy $\chi^2\le\chi^2_\text{lim}$.  Figures \ref{fig:contours_gauss}
and \ref{fig:contours_nn} plot these points after APS has sampled 100,000 
points in the case where $\chi^2_\text{lim}\equiv\chi^2_\text{min}+12.59$
using equations (\ref{eq:covraw}) and (\ref{eq:cov_nn}) as covariograms,
respectively.
We find that the confidence limits returned by APS overlap with
the final credible limits of MCMC, but are generally too permissive,
they include regions of parameter space which MCMC ultimately excludes.
It is tempting to say that
this is due to the fact that the $\chi^2_\text{min}$ discovered by
APS is $\sim3$ greater than the $\chi^2_\text{min}$ discovered by
MCMC (therefore, $\chi^2_\text{lim}$ is $\sim 3$ too large).
However, Figure \ref{fig:contours_br} uses the same data as
Figure \ref{fig:contours_gauss}, but plots only those APS points
such that $\chi^2\le 1283.3$, i.e. $\chi^2\le\chi^2_\text{min, MCMC}+12.59$.
Still we find that the APS confidence limit includes regions of parameter
space excluded by the final MCMC contour.  Thus, the difference in
parameter space constraints between APS and MCMC amounts to difference
between the frequentist confidence limit (see Appendix \ref{sec:frequentism})
and the Bayesian credible limit.  The fact that the two contraints overlap
significantly but not entirely indicates the value in performing multiple
statistical tests before drawing conclusions from one's data set.
It is also worth noting that in all cases, APS has thoroughly explored
the correct confidence limit in fewer samples than has MCMC.  Compare the
coverage of the red scatter plot to the green pre-convergence MCMC contours in
Figures \ref{fig:contours_gauss} through \ref{fig:contours_br}.

As a final note, we do not plot any confidence or 
credible limits for $\tau$.
This is because $\tau$ needs CMB polarization data for meaningful constraints.  For the sake
of simplicity, we did not consider CMB polarization data here.  Thus, $\tau$ has become a
``nuisance parameter'': it effects the fit of our theoretical models, but cannot itself be
meaningfully constrained.  MCMC handles such parameters by effectively integrating over them
when delivering the two-dimensional contours of Figures
\ref{fig:contours_gauss} through \ref{fig:contours_br}.  
APS handles such nuisance parameters by
simply allowing them to vary and returning any combinations of useful-plus-nuisance parameters
that satisfy the $\chi^2\le\chi^2_\text{lim}$ criterion.  The fact that this nuisance
paramter did not significantly impede the convergence of APS relative to MCMC speaks 
well of
APS' ability to handle actual physical problems which 
often include parameters that have
little physical interest but are required to, e.g., 
calibrate noise distributions or
systematics.

\begin{figure*}
\subfigure[]{
\includegraphics[scale=0.15]{scatter_100k_0_1.eps}
}
\subfigure[]{
\includegraphics[scale=0.15]{scatter_100k_0_2.eps}
}
\subfigure[]{
\includegraphics[scale=0.15]{scatter_100k_0_3.eps}
}
\subfigure[]{
\includegraphics[scale=0.15]{scatter_100k_0_4.eps}
}
\subfigure[]{
\includegraphics[scale=0.15]{scatter_100k_1_2.eps}
}
\subfigure[]{
\includegraphics[scale=0.15]{scatter_100k_1_3.eps}
}
\subfigure[]{
\includegraphics[scale=0.15]{scatter_100k_1_4.eps}
}
\subfigure[]{
\includegraphics[scale=0.15]{scatter_100k_2_3.eps}
}
\subfigure[]{
\includegraphics[scale=0.15]{scatter_100k_2_4.eps}
}
\subfigure[]{
\includegraphics[scale=0.15]{scatter_100k_3_4.eps}
}
\caption{
95\% confidence limits in two-dimensional slices of parameter space as
determined by MCMC and APS after 100,000 points have been sampled.  Solid (black)
contours are the baseline result found by 4 MCMC chains
after sampling a total of 460,000 points.  
Dotted (green) contours are the result found by a 4 MCMC
chains after each has sampled 25,000 points (for a total of 100,000
sampled points).  Red points show the $\chi^2\le\chi^2_\text{min}+12.59$
points found by APS after sampling a total of 100,000 points.
These plots use equation (\ref{eq:covraw}) for the covariogram.
}
\label{fig:contours_gauss}
\end{figure*}

\begin{figure*}
\subfigure[]{
\includegraphics[scale=0.15]{scatter_nn_100k_0_1.eps}
}
\subfigure[]{
\includegraphics[scale=0.15]{scatter_nn_100k_0_2.eps}
}
\subfigure[]{
\includegraphics[scale=0.15]{scatter_nn_100k_0_3.eps}
}
\subfigure[]{
\includegraphics[scale=0.15]{scatter_nn_100k_0_4.eps}
}
\subfigure[]{
\includegraphics[scale=0.15]{scatter_nn_100k_1_2.eps}
}
\subfigure[]{
\includegraphics[scale=0.15]{scatter_nn_100k_1_3.eps}
}
\subfigure[]{
\includegraphics[scale=0.15]{scatter_nn_100k_1_4.eps}
}
\subfigure[]{
\includegraphics[scale=0.15]{scatter_nn_100k_2_3.eps}
}
\subfigure[]{
\includegraphics[scale=0.15]{scatter_nn_100k_2_4.eps}
}
\subfigure[]{
\includegraphics[scale=0.15]{scatter_nn_100k_3_4.eps}
}
\caption{
95\% confidence limits in two-dimensional slices of parameter space as
determined by MCMC and APS after 100,000 points have been sampled.  Solid (black)
contours are the baseline result found by 4 MCMC chains
after sampling a total of 460,000 points.  
Dotted (green) contours are the result found by a 4 MCMC
chains after each has sampled 25,000 points (for a total of 100,000
sampled points).  Red points show the $\chi^2\le\chi^2_\text{min}+12.59$
points found by APS after sampling a total of 100,000 points.
These plots use equation (\ref{eq:cov_nn}) for the covariogram.
}
\label{fig:contours_nn}
\end{figure*}
\begin{figure*}
\subfigure[]{
\includegraphics[scale=0.15]{scatter_br_100k_0_1.eps}
}
\subfigure[]{
\includegraphics[scale=0.15]{scatter_br_100k_0_2.eps}
}
\subfigure[]{
\includegraphics[scale=0.15]{scatter_br_100k_0_3.eps}
}
\subfigure[]{
\includegraphics[scale=0.15]{scatter_br_100k_0_4.eps}
}
\subfigure[]{
\includegraphics[scale=0.15]{scatter_br_100k_1_2.eps}
}
\subfigure[]{
\includegraphics[scale=0.15]{scatter_br_100k_1_3.eps}
}
\subfigure[]{
\includegraphics[scale=0.15]{scatter_br_100k_1_4.eps}
}
\subfigure[]{
\includegraphics[scale=0.15]{scatter_br_100k_2_3.eps}
}
\subfigure[]{
\includegraphics[scale=0.15]{scatter_br_100k_2_4.eps}
}
\subfigure[]{
\includegraphics[scale=0.15]{scatter_br_100k_3_4.eps}
}
\caption{
The black and green contours are as in Figure \ref{fig:contours_gauss}.
The red points represent the $\chi^2\le 1283.3$ points sampled by the
APS run used to generate Figure \ref{fig:contours_gauss}.
}
\label{fig:contours_br}
\end{figure*}

\section{Conclusions}
\label{sec:conclusions}

In Section \ref{sec:toy} we showed that APS is more robust
than MCMC against multi-modal likelihood functions.
In Section \ref{sec:wmap} we showed that, on more well-behaved
likelihood functions, APS discovers the same parameter constraints
as MCMC and that APS does not converge any more slowly than does MCMC.
We also showed a great deal of qualitative similarity but quantitative
difference between the Bayesian credible limits drawn by MCMC
and the frequentist confidence limits drawn by APS.  The different
convergence properties and different outputs of APS and MCMC lend themselves
to different use cases.

As was demonstrated in Figure \ref{fig:chimin}, MCMC is significantly
more efficient than APS at finding the true value of $\chi^2_\text{min}$
in a given parameter space.  This makes it ideally suited to problems where
the uncertainty of the data set is poorly understood and must be fit
for.  However, as we saw in Figure \ref{fig:toy}, this rapid convergence to
$\chi^2_\text{min}$ can also result in a myopic disregard for other regions of
interest in parameter space.  Once MCMC has been used to learn the uncertainty
model of the data, it may be advisable to run APS as a guarantor that no
alternative best-fit models have been ignored by MCMC.

In cases where the uncertainty model is well-understood and the contraints in
parameter space can safely be described by either the likelihood ratio's
$\chi^2\le\chi^2_\text{min}+\Delta\chi^2$ or an imposed
$\chi^2\le\chi^2_\text{lim}$, APS' robustness against multi-modality and
comparatively rapid convergence in a confidence limit sense (recall Figures
\ref{fig:contours_gauss} through \ref{fig:contours_br}) make it ideal for rapidly
drawing constraints on parameter space.

One advantage of MCMC that APS cannot overcome is its comparative ease to
implement.  To this end,
we make our code available at \verb|https://github.com/uwssg/APS/|.  
The code is presented as a series of C++ modules with directions
indicating where the user can interface with 
the desired $\vec{\theta}\rightarrow\chi^2$ function.  
Those with questions about the code or the algorithm should not
hesitate to contact the authors.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section*{Acknowledgments}
The authors would like to thank the referee for useful comments and suggestions
which we hope have made this paper more useful to astrophysical audiences.
They would also like to thank Christopher Genovese for useful discussions
regarding statistial theory and interpretation.
SFD would like to thank Eric Linder, Arman Shafieloo, and Jacob Vanderplas 
for useful conversations
about Gaussian processes.  SFD would also like to acknowledge the hospitality of
the Institute for the Early Universe at Ewha Womans University in 
Seoul, Korea, who
hosted him while some of this work was done.
We acknowledge support from DOE award grant number DESC0002607 and NSF grant
IIS-0844580.

\begin{appendix}
\section{Frequentist Confidence Intervals}
\label{sec:frequentism}

Frequentist confidence limits are not nearly as common in the astrophysical
literature as their Bayesian counterparts.  For this reason, we define them below.
Note, however, that, though we have chosen to use a Frequentist
interpretation for the purposes of this current work, \APS is ultimately agnostic
about the user's chosen statistical interpretation.

Suppose we are going to conduct some experiment resulting in a random data set
$\vec{d}$.  Suppose also that we have some theory by which the distribution of
$\vec{d}$ is controlled by a set of parameters $\vec{\theta}$.  We would like to
use $\vec{d}$ to make some statement about the true values of $\vec{\theta}^{(0)}$
i.e., the values of $\vec{\theta}$ actually realized by this Universe.
Rather than adopt the Bayesian perspective 
and talk about the peak and width of the
the posterior
``probability of $\vec{\theta}$ given $\vec{d}$,'' Frequentists attempt to
construct sets of estimators
$\vec{\theta}^{(\text{CL})}(\vec{d})$ such that the
probability that 
$\vec{\theta}^{(0)}\in\vec{\theta}^{(\text{CL})}(\vec{d})$
is $(1-\alpha)\%$, i.e. if the experiment were repeated many times, yielding
many different sets $\vec{d}$, and the confidence limit set
$\vec{\theta}^{(\text{CL})}$
was calculated for each of those data sets, the true
value of $\vec{\theta}^{(0)}$ would fall within $\vec{\theta}^{(\text{CL})}$ 
$(1-\alpha)\%$ of the time.
This is the original definition of confidence intervals posited by Neyman (1937;
see his equation 20 and the attendant discussion).  It also underlies the
confidence belt construction discussed 
in Chapter 9 of Eadie {\it et al}. (1971), Chapter 20 of Stuart and Ord (1991)
and section II B of Feldman and Cousins (1998).

The specific example we consider in Section \ref{sec:wmap}
of this paper is the WMAP 7 year data
release measuring the anisotropy in the Cosmic Microwave Background.  This 
data set takes the form of a
power spectrum composed of 1199 $C_\ell$s ($2\le\ell\le1200$) 
characterizing the
anisotropy power on different angular scales.  
These 1199 measurements represent
independent Gaussian random variables. 
The probability density on $C_\ell$ space associated with
measuring a particular set of $C_\ell$s given a particular candidate set of
$\vec{\theta}^{(c)}$ is
$$P\propto\exp
\big[-\frac{1}{2}
\sum_{\ell,\ell^\prime}(C_\ell(\vec{d})-C_\ell(\vec{\theta}^{(c)})) 
\text{Cov}^{-1}_{\ell\ell^\prime}
(C_{\ell^\prime}(\vec{d})-C_{\ell^\prime}(\vec{\theta}^{(c)}))\big]$$
where 
$C_\ell(\vec{d})$ are the 1199 measured values of $C_\ell$,
$C_\ell(\vec{\theta}^{(c)})$ are the 1199 values of $C_\ell$ predicted
by the candidate theory,
and $\text{Cov}^{-1}_{\ell\ell^\prime}$ is the inverse of the covariance matrix
relating the measured $C_\ell$s.
A change of variables to
$$\chi^2\equiv \sum_{\ell\ell^\prime}
(C_\ell(\vec{d})-\vec{C}_\ell(\vec{\theta}^{(c)}))
\text{Cov}^{-1}_{\ell\ell^\prime}
(C_{\ell^\prime}(\vec{d})-C_{\ell^\prime}(\vec{\theta}^{(c)}))$$
gives the well-known result that the 1199 independent, Gaussian-distributed
$C_\ell$s give a $\chi^2$ statistic distributed according to the eponymous
$\chi^2$ distribution with 1199 degrees of freedom.  The $\chi^2$ distribution
is well-understood.  In the case of 1199 degrees of freedom, 95\% of the
probability is enclosed by $\chi^2\le1281$.  
We may use this fact to construct a
95\% confidence limit.

Recall the definition of Frequentist confidence limits.  The true
value of the parameters
$\vec{\theta}^{(0)}$ is fixed, but $\vec{d}$ (in the specific example above, the
set of 1199 $C_\ell$s) is random and will change each time we conduct the
experiment.  In this case, ``conducting the experiment'' means creating a new Universe
with a new Cosmic Microwave Background, randomly generated from the same
$\vec{\theta}^{(0)}$, and measuring it with WMAP;
while this is impossible, we ask the readers to
suspend their disbelief for the sake of this illustration.  
Because of what we have just noted about $\chi^2$, in
95\% of our repeated WMAP experiments $\chi^2$ calculated relative to the true
parameter set $\vec{\theta}^{(0)}$ will be less than or equal to 1281.  If, for
each set of $C_\ell$s we find all of the combinations of $\vec{\theta}$ that
yield $\chi^2\le1281$, that set of $\vec{\theta}$ will contain $\vec{\theta}^{(0)}$
95\% of the time.
Taking our one realization of $C_\ell$ and finding all of the values of
$\vec{\theta}^{(\text{CL})}$ which give $\chi^2\le1281$,
we can be confident that we have contained $\vec{\theta}^{(0)}$ with 95\%
probability.  This is what is meant by a 95\% Frequentist confidence limit.

In the large data limit, Frequentist confidence limits are expected to
give comparable results to Bayesian inference (as we see in Figure
\ref{fig:contours_gauss}).  In the limit of small data, Frequentist confidence limits
may differ from Bayesian limits.  If statistical fluctuations dominate the
signal, it is possible for Frequentist methods to return an empty confidence
limit (no points in parameter space fit the data).  While this may seem an
unpalatable outcome, it is useful to know that one's data set is noisy so that
one can interpret the explored likelihood surface with all due caution.
Lyons (2002 and 2008) discusses this distinction, as well as the
comparative strengths and weaknesses of other statistical perspectives, at
greater length than this work.

The Frequentist approach to confidence limits described herein
is actually a conservative version of
the non-parametric confidence ball approach popular among academic statisticians
(see the introduction to Baraud 2004).  Statisticians typically are concerned
with using data to constrain the form of a function in arbitrary function space.
They use the data to derive an estimator of the function 
from which it was drawn (in our example, the smooth theoretical $C_\ell$ function)
and then use
theoretical considerations to draw a hypersphere in function space about that
fit constituting the $(1-\alpha)\%$ confidence limit
\cite{bd,li,baraud,cai,davies}.
Note that these hyperspheres (referred to in the literature as ``confidence
balls'') are calculated from the data alone without reference to any
underlying physical model or set of (in Wilks' 1938 words) ``admissible
hypotheses.''  In this way, they function identically to the $\chi^2$ test
adopted in the present work.

Astrophysicists can utilize the
``confidence ball'' formalism by demanding that their theoretical
models give predicted functions that fall within 
the hypersphere drawn in function space.  This is how
Bryan {\it et al}. (2007) originally drew their constraints on the CMB.  It is
also how Genovese {\it et al}. (2004) confirmed the statistical significance of
the harmonic peaks in the WMAP 1-year CMB data.  Because the confidence
hypersphere is drawn a-priori, without any consideration for the admissible
hypotheses within the space of physical parameters, this method would be
straightforward to interface with \APS (indeed, it was included in the Bryan
{\it et al}. draft of the code).  
We leave such an implementation to future work.

The more familiar likelihood ratio test detailed
by Wilks (1938)  and Neyman and Pearson (1933) is an asymptotic simplification
of Frequentist confidence limits
for cases in which we know that the true model must exist
within some limited set of hypotheses.  
In this case, one assumes that the adopted $\vec{\theta}$ parametrization
is the only possible way of explaining that data.  In that case, the
$\chi^2_\text{min}$ on parameter space must, by definition, be the smallest
achievable $\chi^2$.  $\Delta\chi^2\equiv\chi^2-\chi^2_\text{min}$
is then distributed according to the $\chi^2$ distribution with $N_p$ 
degrees of freedom.  This is the assumption underlying our use of
$\Delta\chi^2$ in the test of Section \ref{sec:toy} and \ref{sec:wmap} above.

\end{appendix}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{thebibliography}{99}

\bibitem[Atchad\'e and Rosenthal 2005]{mcmcefficiency}
Atchad\'e, Y.~F. and Rosenthal, J.~S., 2005, Bernoulli {\bf 11}, 815

\bibitem[Baraud 2004]{baraud}
Baraud,~Y. 2004, Annals of Statistics {\bf 32} 528

\bibitem[Bentley 1975]{kdtree}
Bentley,~J.~L. 1974, Communications of the Associaton for Computing Machinery
{\bf 18}, 509

\bibitem[Beran and D\"umbgen 1998]{bd}
Beran,~R. and D\"umben,~L. 1998, Annals of Statistics {\bf 26} 1826

\bibitem[Berg\'e {\it et al}. 2012]{psf}
Berg\'e,~J., Price,~S., Amara,~A., and Rhodes,~J. 2012,
Monthly Notices of the Royal Astronomical Societ {\bf 419}, 2356

\bibitem[Brooks and Gelman 1998]{mcmcconvergence}
Brooks, S. and Gelman, A., 1998, 
Journal of Computational and Graphical Statistics
{\bf 7} 434

\bibitem[Bryan 2007]{brentsthesis}
Bryan, B., 2007, Ph.D. thesis
\verb|http://reports-archive.adm.cs.cmu.edu/anon/|
\verb|ml2007/abstracts/07-122.html|

\bibitem[Bryan {\it et al}. 2007]{bryan}
Bryan, B., Schneider, J., Miller, C.~J., Nichol, R.~C., Genovese, C., and
Wasserman, L., 2007,
Astrophys. \ J. {\bf 665}, 25

\bibitem[Cai and Low 2006]{cai}
Cai,~T.~T. and Low,~M.~G. 2006, Annals of Statistics {\bf 34} 202

\bibitem[Davies {\it et al}. 2009]{davies}
Davies,~P.~L., Kovac,~A., and Meise,~M. 2009, Annals of Statistics,
{\bf 37} 2597

\bibitem[Dunkley {\it et al}. 2005]{mcmcdunkley}
Dunkley, J. Bucher, M., Ferreira, P.~G., Moodley, K., and Skordis, C.,
2005,  
Mon. \ Not. \ R. \ Astron. \ Soc. {\bf 356},
925

\bibitem[Eadie {\it et al}. 1971]{eadie}
Eadie,~W.~T., Drijard,~D., James,~F.~E., Roos,~M., and Sadoulet,~B.,
1971, ``Statistical Methods in Experimental Physics''
(North-Holland, Amsterdam).

\bibitem[Feldman and Cousins 1998]{feldman}
Feldman,~G.~J. and Cousins,~R.~D. 1998, Phys. Rev. D. {\bf 57} 3873

\bibitem[Gelman {\it et al}. 1996]{gelman}
Gelman, A., Roberts, G. O., and Gilks, W. R. 1996,
Bayesian Statistics {\bf 5}, 599

\bibitem[Genovese {\it et al}. 2004]{genovese}
Genovese,~C.~R., Miller,~C.~J., Nichol,~R.~C., Arjunwadkar,~M., and
Wasserman,~L. 2004, Statistical Science {\bf 19} 308

\bibitem[Gilks {\it et al}. 1996]{mcmc}
Gilks, W.~R., Richardson, S., and Spiegelhalter, D.~J. ed.,
``Markov Chain Monte Carlo in Practice,'' (Chapman \& Hall, 1996),
Boca Raton, F.L.

\bibitem[Habib {\it et al}. 2007]{habib}
Habib,~S., Heitmann,~K., Higdon,~D., Nakhleh,~C., and
Williams,~B. 2007, Physical Review D {\bf 76} 083503

\bibitem[James 1972]{minuit}
James,~F. 1972, Proceedings of the CERN Computing and Data Processing School,
\verb|http://seal.web.cern.ch/seal/documents/minuit.mntutorial.pdf|

\bibitem[Jarosik {\it et al.} 2011]{wmap7}
N.~Jarosik {\it et al}., 2011, Astrophys. \ J. \ Suppl. \ Ser. {\bf 192} 14

\bibitem[Li 1989]{li}
Li,~K.-C. 1989, Annals of Statistics {\bf 17} 1001

\bibitem[Lewis and Bridle 2002]{cosmomc}
Lewis, A. and Bridle, S., 2002, Phys.\ Rev. \ D {\bf 66}, 103511;
\verb|http://cosmologist.info/cosmomc/readme.html|

\bibitem[Lewis {\it et al}. 2000]{camb}
Lewis, A., Challinor, A., and Lasenby, A., 2000, 
Astrophys. \ J. {\bf 538}, 473

\bibitem[Lyons 2002]{lyonsbf}
Lyons,~L. 2002, ``Bayes or Frequentism?''
\verb|http://www-cdf.fnal.gov/physics/statistics/|
\verb|statistics_recommendations.html|

\bibitem[Lyons 2008]{lyonsall}
Lyons,~L. 2008, Annals of Applied Statistics, {\bf 2} 887
[arXiv:0811.1663]

\bibitem[Nelder and Mead 1965]{simplex}
Nelder,~J.~A. and Mead,~R. 1965, The Computer Journal, {\bf 7} 308

\bibitem[Neyman and Pearson 1933]{np}
Neyman,~J. and Pearson,~E.~S. 1933, Philosophical Transactions
of the Royal Society of London, Series A, {\bf 231} 289

\bibitem[Neyman 1937]{neyman}
Neyman,~J. 1937, Philos. Trans. R. Soc. London {\bf A236} 333.
Reprinted in ``A Selection of Early Statistical Papers of J. Neyman''
(University of California Press, Berkeley, 1967)

\bibitem[Rasmussen and Williams 2006]{gp}
Rasmussen, C.~E. and Williams, C.~K.~I., 2006, ``Processes for Machine Learning''
\verb|http://www.GaussianProcess.org/gpml|

\bibitem[Shafieloo {\it et al}. 2012]{ericgp}
Shafieloo,~A., Kim,~A.~G., and Linger,~E.~V. 2012,
Physical Review D {\bf 85}, 123530 [arXiv:1204.2272]

\bibitem[Stuart and Ord 1991]{stuart}
Stuart,~A. and Ord,~J.~K., 1991, ``Classical Inference and Relationship,''
5th ed., Kendall's Advanced Theory of Statistics, Vol. 2
(Oxford University Press, New York).

\bibitem[Vanderplas and Connolly 2012]{gradientdescent}
Vanderplas, J. and Connolly, A., 2012, [in preparation]

\bibitem[Walsh 2004]{walsh}
Walsh,~B. 2004, Lecture Notes for EEB 581,
\verb|http://www.web.mit.edu/~wingated/www/|
\verb|introductions/mcmc-gibbs-intro.pdf|

\bibitem[WMAP 2010]{wmap7likelihood}
WMAP, 2010
\verb|http://lambda.gsfc.nasa.gov/product/map/|
\verb|current/likelihood_info.cfm|

\bibitem[Wilks 1938]{wilks}
Wilks,~S.~S. 1938, Annals of Mathematical Statistics, {\bf 9} 60

\end{thebibliography}

\label{lastpage}

\end{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2014-07-20 aps4p5/aps.cpp

@@ -1322,6 +1322,17 @@ int aps::get_nn(array_1d<double> &pt){
     return neigh.get_data(0);
 }
 
+double aps::get_chival(int dex){
+    if(dex>=gg.get_pts() || dex<0){
+        printf("WARNING asked aps for chi at %d but pts %d\n",
+        dex,gg.get_pts());
+        
+        exit(1);
+    }
+    
+    return gg.get_fn(dex);
+}
+
 int aps::get_n_centers(){
     return center_dexes.get_dim();
 }

### 2014-07-20 aps4p5/aps.h

@@ -92,6 +92,7 @@ public:
     
     array_1d<double>* get_pt(int);
     int get_nn(array_1d<double>&);
+    double get_chival(int);
     
 private:
     Ran *dice;

### 2014-07-20 aps4p5/aps_ellipse_brute.cpp

@@ -88,11 +88,15 @@ for(i=0;i<ncenters;i++){
 FILE *output;
 output=fopen(brute_name,"a");
 
-int last_assessed=0;
-double dd,ddmax;
+int last_assessed=0,ic;
+double dd,ddmax,ddc;
+array_1d<int> found_it;
 
 found_all=0;
 printf("time to start searching\n");
+for(i=0;i<ncenters;i++)found_it.set(i,0);
+
+
 while(aps_test.get_called()<100000 && found_all==0){
     aps_test.search();    
     
@@ -107,10 +111,13 @@ while(aps_test.get_called()<100000 && found_all==0){
             dd=euclideanDistance(*true_centers(i),*aps_test.get_pt(j));
           
             if(dd>ddmax)ddmax=dd;
+            
+            if(aps_test.get_chival(j)<11.0)found_it.set(i,1);
         }
         
-        if(ddmax<1.0){
-            found_all=1;
+        found_all=1;
+        for(i=0;i<ncenters;i++){
+            if(found_it.get_data(i)==0)found_all=0;
         }
     
     }

### 2014-07-20 aps4p5/aps_ellipse_brute.cpp
added the brute test scripts to the repo

@@ -124,8 +124,11 @@ while(aps_test.get_called()<100000 && found_all==0){
     
 }
 
-fprintf(output,"seed %d ddmax %e called %d\n",
+fprintf(output,"seed %d ddmax %e called %d ",
 seed,ddmax,aps_test.get_called());
+fprintf(output,"found ");
+for(i=0;i<ncenters;i++)fprintf(output,"%d ",found_it.get_data(i));
+fprintf(output,"\n");
 fclose(output);
 
 printf("that took %e\n",double(time(NULL))-before);

### 2014-07-20 aps4p5/test_mcmc.sh

@@ -1,0 +1,11 @@
+make mcmcBrute
+
+for ((i=0 ; i<100 ; i++)) do
+./mcmcBrute -1 5 2
+sleep 5
+done
+
+for ((i=0 ; i<100 ; i++)) do
+./mcmcBrute -1 5 4
+sleep 5
+done

### 2014-07-20 aps4p5/drafts/draft_140719.tex
edited the draft

@@
 
-Bryan {\it et al.} (2007) show that this algorithm is robust against multi-modal
+\subsection{Modifications to the APS algorithm in the present work}
+
+Bryan {\it et al.} (2007) show that the algorithm above 
+is robust against multi-modal
 likelihood functions, using it to identify a disjoint region of high likelihood in
@@
 
-\subsubsection{Characterization of known high likelihood regions}
+We are greatly indebted to Eric Linder of the Lawrence Berkeley National
+Laboratory who first suggested to us the modifications described below
+in sections \ref{sec:focus} and \ref{sec:simplex}.
+
+\subsubsection{Characterization of known low-$\chi^2$ regions}
 \label{sec:focus}
 
-As the algorithm searches, it will keep track of a list of all of the discovered points
+As it searches, APS will keep track of a list of all of the discovered points
 which are believed to be local minima of $\chi^2$ (how it finds these will be described below
-in Section \ref{sec:simplex}).  After each cycle of the main algorithm (steps 1A-3A above),
+in Section \ref{sec:simplex}).  These points will be stored in the set
+$\{\Gamma\}$.  After each cycle of the main algorithm (steps 1A-3A above),
 our version of APS attempts ``focused searches'' about these local minima.
 
-For each local minimum $\vec{M}$, APS performs the following:
+For each local minimum $\vec{M}$ in $\{\Gamma\}$, APS performs the following:
 \begin{itemize}
@@
 
-\subsubsection{Extending the boundary}
+\subsubsection{Extension of the $\chi^2=\chi^2_\text{lim}$ contour}
 \label{sec:unitSphere}
@@
 perpendicularly to its ``radii'' (the vector $\vec{B}^{(j)}-\vec{M}$ would be a radius of
-$\chi^2=\chi^2_\text{lim}$ described a sphere).
-
-something about APS wide helping to discover new corners of the surface as well
+$\chi^2=\chi^2_\text{lim}$ described a sphere).  It is, however, also possible
+that the $S$-maximization in steps (1A-3A) could also learn something about the
+shape of the $\chi^2=\chi^2_\text{lim}$ contour, even if it does not place
+points directly on that contour.  To explore that possibility, we add the
+following steps to the end of the sequence (1A-3A).
+
+\begin{itemize}
+\item(1C) For each point $\vec{M}$ in $\{\Gamma\}$, store, 
+in addition to the boundary
+points associated with $\vec{M}$, the projection of those boundary points
+onto a unit sphere in parameter space, i.e. for each boundary point, draw a line
+connecting that boundary point with $\vec{M}$ and keep track of
+where a unit sphere centered on $\vec{M}$ intersects that line.  Call this set
+of projectsion $\vec{s}^{(i)}$.
+\\
+\item(2C) Call the point for which $\chi^2$ has been evaluated in step (3A)
+$\vec{P}$., find the
+point $\vec{M}$ in $\{\Gamma\}$ that is nearest to $\vec{P}$ in parameter space.
+\\
+\item(3C) Find the
+projection of $\vec{P}$ onto the unit sphere surrounding $\vec{M}$.  
+Call this projected point $\vec{u}$.  Find the point among
+$\vec{s}^{(i)}$ that is nearest to $\vec{u}$.  Record the distance between them.
+Keep track of this distance in the set $\{D\}$ ($\{D\}$ will be a running list
+of the $\vec{u}-\vec{s}^{(i)}$ distance for all points $\vec{P}$ found by steps
+1A-3A).
+\\
+\item(4C) If the distance between $\vec{u}$ and its nearest neighbor in
+$\vec{s}^{(i)}$ is greater than the 2/3 quantile of $\{D\}$,
+then perform bisection between $\vec{M}$ and the point evaluated in step (3A).
+\end{itemize}
+
+This way, points on the $\chi^2=\chi^2_\text{lim}$ contour can be found whose
+directions away from $\vec{M}$ in parameter space have not yet been explored.
+Because we expect the unit sphere surrounding $\vec{M}$ to become more densely
+populated as the algorithm runs, we periodically delete the list $\{D\}$ so that
+new directions away from $\vec{M}$ can still be explored.
 
@@
 \label{sec:simplex}
+
+The final modification to the original APS algorithm which we propose in this
+work is to utilize function minimization to identify local minima in $\chi^2$.
+One of the principal drawbacks of the algorithm presented in steps (1A-3A) is
+that it makes no account of ``near misses.''  If step (3A) finds that $\chi^2$
+is lower than average but still greater than $\chi^2_\text{lim}$, the original
+APS algorithm does nothing with that information.  We have tried to improve upon
+that with our incorporation of bisection in sections \ref{sec:focus} and
+\ref{sec:unitSphere} above.  Our final innovation is to allow APS to use points
+with 
+\begin{equation}
+\label{eqn:simplexCriterion}
+\chi^2_\text{lim}\le\chi^2\le\chi^2_\text{sig}
+\end{equation}
+to seed a simplex function minimizer as developed by Nelder and Mead (1965).
+
+We set
+$\chi_\text{sig}=\chi^2_\text{min}+G\times(\chi^2_{1/10}-\chi^2_\text{min})$
+where $G$ is a user-set parameter, $\chi^2_\text{min}$ is the minimum discovered
+$\chi^2$ value and $\chi^2_{1/10}$ is the 1/10 quantile of the $\chi^2$ values
+discovered by steps (1A-3A) (not those discovered by any of the modifications
+discussed in sections \ref{sec:focus}, \ref{sec:unitSphere} or
+\ref{sec:simplex}).
+
+Because the Nelder-Mead simplex minimizer can be very time intensive, we balance our
+algorithm by demanding that the number of $\chi^2$ evaluations devoted to all
+other variants of search (i.e. steps (1A-3A) and (1B-5B)) equal the number of
+$\chi^2$ evaluations devoted to the simplex minimizer before a new simplex
+minimization can occur, and if no new points are known that satisfy the
+criterion in equation \ref{eqn:simplexCriterion} above, no simplex minimization
+will be undertaken, regardless.
+
+If the simplex minimizer converges to a point such that
+$\chi^2<\chi^2_\text{lim}$, our APS determines whether or not it
+has found a new local minimum to be added to $\{\Gamma\}$ by comparing the end
+point $\vec{\gamma}$ of the minimization to each of the points already in
+$\{\Gamma\}$.  For each $\vec{M}$ in $\{\Gamma\}$, APS calculates the mid-point
+between $\vec{\gamma}$ and $\vec{M}$.  If, for all $\vec{M}$, 
+$\chi^2$ at that mid-point is greater
+than $\chi^2_\text{lim}$, then it is deemed that $\vec{\gamma}$ is a new local
+minimum and is added to $\{\Gamma\}$.  If not, $\vec{\gamma}$ is assumed to be a
+part of the one of the previously identified loci of low $\chi^2$ and it is not
+added to $\{\Gamma\}$.
 

### 2014-07-20 aps4p5/drafts/draft_140719.tex
more writing draft

--- 2014-07-20 aps4p5/drafts/draft_140719.tex
@@
 parameter space are actually interesting enough to warrant
- $\vec{\theta}\rightarrow\chi^2$ evalutation.
+ $\vec{\theta}\rightarrow\chi^2$ evaluation.
 
@@
 \\
-\item(2A) Generate some number $N_c$ of candidate points.  Use the points
-already sampled by APS to predict the value of $\chi^2$ at each of these points.
+\item(2A) Generate some number $N_c$ of candidate points.  
+For each of these points, find the $N_g$ nearest points already sampled
+by APS. Use these nearest neighbor points and their corresponding $\chi^2$ values
+to predict the value of $\chi^2$ at each of these points.
  We will signify this prediction by $\mu$.  Calculate some value $\sigma$
@@
 \begin{equation}
-\label{eq:sstat}
+\label{eqn:sstat}
 S=\sigma-|\mu-\chi^2_\text{lim}|
@@
 
-The $|\mu-\chi^2_\text{lim}|$ term in equation (\ref{eq:sstat}) provides an
+The $|\mu-\chi^2_\text{lim}|$ term in equation (\ref{eqn:sstat}) provides an
 incentive for APS to choose points that it believes will lie on the boundary of
@@
 
-Bryan {\it et al.} (2007) show that this algorithm is robust against multi-modal
-likelihood functions, using it to identify a disjoint region of high likelihood in
-cosmological parameter space as constrained by the 1-year WMAP CMB anisotropy power
-spectrum.  This is the principal advantage to selecting points for evaluation according
-to the statistic (\ref{eq:sstat}).  Unfortunately, it is possible that the focus on exploring
-unknown regions of parameter space might slow down convergence in the case of uni-modal
-likelihood functions.  Therefore, we propose the following modifications to the algorithm
-to ensure that it adequately characterizes known regions of high likelihood while
-simultaneously searching for new regions to explore.
-
-\subsubsection{Characterization of known high likelihood regions}
-\label{sec:focus}
-
-As the algorithm searches, it will keep track of a list of all of the discovered points
-which are believed to be local minima of $\chi^2$ (how it finds these will be described below
-in Section \ref{sec:simplex}).  After each cycle of the main algorithm (steps 1A-3A above),
-our version of APS attempts ``focused searches'' about these local minima.
-
-For each local minimum $\vec{M}$, APS performs the following:
-\begin{itemize}
-\item (1B) If fewer than $N_p$ focused searches have been proposed around this minimum, randomly
-choose a point from a small sphere surrounding $\vec{M}$ in parameter space and 
-evaluate $\chi^2$ at that point.  
-\\
-\item(2B) If the point does not satisfy $\chi^2=\chi^2_\text{lim}$ (which is highly probable), 
-use bisection
-along the line connecting the proposed point to $\vec{M}$ to find a point which does
-satisfy $\chi^2=\chi^2_\text{lim}$, i.e. find a point on that line with
-$\chi^2<\chi^2_\text{lim}$ -- often $\vec{M}$ itself -- and a point along that line
-with $\chi^2>\chi^2_\text{lim}$ and iteratively step half the distance between them until
-arriving at the desired $\chi^2_\text{lim}$ \cite{minuit}.  The points discovered this way with
-$\chi^2=\chi^2_\text{lim}$ are stored as ``boundary points'' about $\vec{M}$.  Each local
-minimum will have its own list of boundary points.
-\\
-\item(3B) If $N_p$ boundary points have been found for $\vec{M}$, skip steps (1B) and (2B).
-Instead, examine the boundary points associated with $\vec{M}$ and identify the points
-corresponding to both the maximum and minimum values of each parameter $\theta_i$.  These
-points will be referred to as $\vec{B}^{(j)}$.  There will be $2\times N_p$ of them, one for
-the maximum and one for the minimum in each of our $N_p$ parameters.  
-\\
-\item(4B) For each point
-$\vec{B}^{(j)}$, propose some number $N_v$ of new points (we have chosen
-$N_v=20$ for the sake of speed) $\vec{d}^{(k)} = \vec{B}^{(j)}+\vec{v}$ where $\vec{v}$ is a
-random, small vector in parameter space constrained to be perpendicular to
-$\vec{B}^{(j)}-\vec{M}$.  There will now be $2\times N_p \times N_v$ points $\vec{d}^{(k)}$
-($N_v$ such points for each of the $2\times N_p$ extremal boundary points $\vec{B}^{(j)}$).
-\\
-\item(5B) At each point $\vec{d}^{(k)}$, evaluate equation (\ref{eq:sstat}).  Evaluate $\chi^2$
-at the point $\vec{d}^{(k)}$ which maximizes $S$. If
-$\chi^2 \neq \chi^2_\text{lim}$, bisection is again used and a new boundary point is found.
-\end{itemize}
-
-We are indebted to Eric Linder of Lawrence Berkeley National Lab who first suggested to us
-both the use of bisection to isolate $\chi^2_\text{lim}$ and the idea of stepping
-intentionally along the $\chi^2=\chi^2_\text{lim}$ surface by stepping perpendicularly to the
-direction $\vec{B}^{(j)}-\vec{M}$.
-
-To reiterate, our version of APS alternates between performing steps (1A-3A) and steps
-(1B-5B).  If there are $N_m$ distinct known local minima in $\chi^2$, or algoritm will perform
-$N_m$ iterations of steps (1A-3A), and then perform steps (1B-5B) once on each local minimum.
-
-\subsubsection{Extending the boundary}
-\label{sec:unitSphere}
-
-The modification suggested in section \ref{sec:focus} is designed to explore 
-$\chi^2=\chi^2_\text{lim}$ surface by starting from its extreme corners and stepping
-perpendicularly to its ``radii'' (the vector $\vec{B}^{(j)}-\vec{M}$ would be a radius of
-$\chi^2=\chi^2_\text{lim}$ described a sphere).
-
-something about APS wide helping to discover new corners of the surface as well
-
-\subsubsection{Function minimization}
-\label{sec:simplex}
+In the present work, we introduce modifications and expansions to this basic algorithm as
+described below.
 
@@
 defining it as $\chi^2_\text{min}+\Delta\chi^2$.  Here, $\chi^2_\text{min}$
-is the minimum value of $\chi^2$ discovered by APS.  $\Delta \chi^2$ is set
+is the minimum value of $\chi^2$ discovered by APS.  
+$\Delta \chi^2$ is set
 based on the properties of the $N_p$-dimensional parameter space being explored.
@@
 In this case, the user is relying on APS to eventually find the true
-$\chi^2_\text{min}$ of the likelihood surface being explored.  We show in
-Section \ref{sec:wmap} that, with the inclusion of the gradient descent search
-discussed above, this is a safe assumption to make.
-
-\subsection{User-specified Parameters of APS}
+$\chi^2_\text{min}$ of the likelihood surface being explored.  
+Section \ref{sec:simplex} below outlines an additional means of search
+by which our APS is capable of efficiently locating $\chi^2_\text{lim}$.
+
+\subsection{Modified search algorithms}
+
+Bryan {\it et al.} (2007) show that the APS search algorithm as presented in steps (1A-3A)
+is robust against multi-modal
+likelihood functions, using it to identify a disjoint region of high likelihood in
+cosmological parameter space as constrained by the 1-year WMAP CMB anisotropy power
+spectrum.  This is the principal advantage to selecting points for evaluation according
+to the statistic (\ref{eqn:sstat}).  Unfortunately, it is possible that the focus on 
+exploring
+unknown regions of parameter space might slow down convergence in the case of uni-modal
+likelihood functions.  Therefore, we propose the following modified search
+to ensure that it adequately characterizes known regions of high likelihood while
+simultaneously searching for new regions to explore.
+
+These modified searches will require APS to keep track of information it learns about the
+$\vec{\theta}\rightarrow\chi^2$ fuction as it searches.  
+To facilitate this discussion, we introduce the folling notations.
+Symbols denoted with $\vec{x}$ represent points in parameter space.  Variables denoted with
+$\{\hat{X}\}$ denote sets of points in parameter space.  Variables denoted $\{X\}$ are sets
+of numbers stored by APS.
+
+We are greatly indebted to Eric Linder of the Lawrence Berkeley National
+Laboratory who first suggested to us the modifications described below
+in sections \ref{sec:focus} and \ref{sec:simplex}.
+
+\subsubsection{Characterization of known low-$\chi^2$ regions}
+\label{sec:focus}
+
+As it searches, APS will keep track of a list of all of the discovered points
+which are believed to be local minima of $\chi^2$ 
+(how it finds these will be described below
+in Section \ref{sec:simplex}).  These points will be stored in the set
+$\{\hat{\Gamma}\}$.  After each cycle of the main algorithm (steps 1A-3A above),
+our version of APS attempts ``focused searches'' about the points in $\{\hat{\Gamma}\}$.
+
+For each point $\vec{M}$ in $\{\hat{\Gamma}\}$, APS performs the following:
+\begin{itemize}
+\item (1B) If fewer than $N_p$ focused searches have been proposed around this minimum, 
+randomly choose a point from a small sphere surrounding $\vec{M}$ in parameter space and 
+evaluate $\chi^2$ at that point.  
+\\
+\item(2B) If the point does not satisfy $\chi^2=\chi^2_\text{lim}$ (which is highly probable), 
+use bisection
+along the line connecting the proposed point to $\vec{M}$ to find a point which does
+satisfy $\chi^2=\chi^2_\text{lim}$, i.e. find a point on that line with
+$\chi^2<\chi^2_\text{lim}$ -- often $\vec{M}$ itself -- and a point along that line
+with $\chi^2>\chi^2_\text{lim}$ and iteratively step half the distance between them until
+arriving at the desired $\chi^2_\text{lim}$ \cite{minuit}.  The points discovered this way \
+with $\chi^2=\chi^2_\text{lim}$ are stored in the set $\{\hat{B}(\vec{M})\}$ of 
+``boundary points'' about $\vec{M}$ (each $\vec{M}$ in $\{\hat{\Gamma}\}$ will have its own
+set $\{\hat{B}(\vec{M})\}$).  
+\\
+\item(3B) If $N_p$ boundary points have been found for $\vec{M}$, skip steps (1B) and (2B).
+Instead, examine the points in $\{\hat{B}(\vec{M})\}$ and identify the points
+corresponding to both the maximum and minimum values of each parameter $\theta_i$.  These
+points will be referred to as $\{\hat{B}_\text{MaxMin}(\vec{M})\}$.  
+There will be $2\times N_p$ points in $\{\hat{B}_\text{MaxMin}(\vec{M})\}$, one for
+the maximum and one for the minimum in each of our $N_p$ parameters.  
+\\
+\item(4B) For each point $\vec{b}$ in $\{\hat{B}_\text{MaxMin}(\vec{M})\}$, 
+propose some number $N_v$ of new points (we have chosen
+$N_v=20$ for the sake of speed) $\vec{v} = \vec{b}+\vec{\epsilon}$ where $\vec{\epsilon}$ is a
+random, small vector in parameter space constrained to be perpendicular to
+$\vec{b}-\vec{M}$.  There will now be $2\times N_p \times N_v$ points $\vec{v}$
+($N_v$ such points for each of the $2\times N_p$ extremal boundary points in
+$\{\hat{B}_\text{MaxMin}(\vec{M})\}$).
+\\
+\item(5B) At each point $\vec{v}$ above, evaluate equation (\ref{eqn:sstat}).  
+Evaluate $\chi^2$ at the point $\vec{v}$ which maximizes $S$. If
+$\chi^2 \neq \chi^2_\text{lim}$, bisection is again used and a new boundary point is found.
+\end{itemize}
+
+To reiterate, our version of APS alternates between performing steps (1A-3A) and steps
+(1B-5B).  If there are $N_m$ distinct known local minima in $\{\hat{\Gamma}\}$, 
+our algoritm will perform
+$N_m$ iterations of steps (1A-3A), and then perform steps (1B-5B) once for each
+$\vec{M}$ in $\{\hat{\Gamma}\}$.
+
+\subsubsection{Extension of the $\chi^2=\chi^2_\text{lim}$ contour}
+\label{sec:unitSphere}
+
+The modification suggested in section \ref{sec:focus} is designed to explore 
+$\chi^2=\chi^2_\text{lim}$ surface by starting from its extreme corners and stepping
+perpendicularly to its ``radii'' (the vector $\vec{b}-\vec{M}$ would be a radius of
+$\chi^2=\chi^2_\text{lim}$ described a sphere).  It is, however, also possible
+that the $S$-maximization in steps (1A-3A) could also learn something about the
+shape of the $\chi^2=\chi^2_\text{lim}$ contour, even if it does not place
+points directly on that contour.  To explore that possibility, we add the
+following steps to the end of the sequence (1A-3A).
+
+\begin{itemize}
+\item(1C) For each point $\vec{M}$ in $\{\hat{\Gamma}\}$, store, 
+in addition to the boundary
+points $\{\hat{B}(\vec{M})\}$, the projection of those boundary points
+onto a unit sphere in parameter space, i.e. for each point $\vec{b}$ in $\{\hat{B}(\vec{M})\}$, draw a line
+connecting $\vec{b}$ with $\vec{M}$ and keep track of
+where a unit sphere centered on $\vec{M}$ intersects that line.  Call this set
+of projectsion $\{\hat{U}\}$.
+\\
+\item(2C) Call the point for which $\chi^2$ has been evaluated in step (3A)
+$\vec{P}$., find the
+point $\vec{M}$ in $\{\hat{\Gamma}\}$ that is nearest to $\vec{P}$ in parameter space.
+\\
+\item(3C) Find the
+projection of $\vec{P}$ onto the unit sphere surrounding $\vec{M}$.  
+Call this projected point $\vec{p}$.  Find the point $\vec{u}$ in
+$\{\hat{U}\}$ that is nearest to $\vec{p}$.  Record the parameter space 
+distance between $\vec{u}$ and $\vec{p}$.
+Add this distance to the set $\{D\}$ ($\{D\}$ will be a running list
+of the $\vec{u}-\vec{p}$ distance for all points $\vec{P}$ found by steps
+1A-3A, regardless of which point in $\vec{M}$ they are nearest).
+\\
+\item(4C) If the $\vec{p}-\vec{u}$ distance is greater than the 2/3 quantile of $\{D\}$,
+then perform bisection between $\vec{M}$ and the point evaluated in step (3A).
+\end{itemize}
+
+This way, points on the $\chi^2=\chi^2_\text{lim}$ contour can be found whose
+directions away from $\vec{M}$ in parameter space have not yet been explored.
+Because we expect the unit spheres surrounding the points $\vec{M}$ to become more densely
+populated as the algorithm runs, we periodically delete the list $\{D\}$ so that
+new directions away from $\vec{M}$ can still be explored.
+
+\subsubsection{Function minimization}
+\label{sec:simplex}
+
+The final modification to the original APS algorithm which we propose in this
+work is to utilize function minimization to identify local minima in $\chi^2$.
+One of the principal drawbacks of the algorithm presented in steps (1A-3A) is
+that it makes no account of ``near misses.''  If step (3A) finds that $\chi^2$
+is lower than average but still greater than $\chi^2_\text{lim}$, the original
+APS algorithm does nothing with that information.  We have tried to improve upon
+that with our incorporation of bisection in sections \ref{sec:focus} and
+\ref{sec:unitSphere} above.  Our final innovation is to allow APS to use points
+with 
+\begin{eqnarray}
+\chi^2_\text{lim}&\le&\chi^2\le\chi^2_\text{threshold}\label{eqn:simplexCriterion}\\
+\chi_\text{threshold}&=&\chi^2_\text{min}+G\times(\chi^2_{1/10}-\chi^2_\text{min})
+\label{eqn:threshold}
+\end{eqnarray}
+to seed a simplex function minimizer as developed by Nelder and Mead (1965).
+Here, $G$ is a user-set parameter, $\chi^2_\text{min}$ is the minimum discovered
+$\chi^2$ value and $\chi^2_{1/10}$ is the 1/10 quantile of the $\chi^2$ values
+discovered by steps (1A-3A) (not those discovered by any of the modifications
+discussed in sections \ref{sec:focus}, \ref{sec:unitSphere} or
+\ref{sec:simplex}).
+
+Because the Nelder-Mead simplex minimizer can be very time intensive, we balance our
+algorithm by demanding that the number of $\chi^2$ evaluations devoted to all
+other variants of search (i.e. steps (1A-3A) and (1B-5B)) equal the number of
+$\chi^2$ evaluations devoted to the simplex minimizer before a new simplex
+minimization can occur.
+
+If the simplex minimizer converges to a point such that
+$\chi^2<\chi^2_\text{lim}$, our APS determines whether or not it
+has found a new local minimum to be added to $\{\hat{\Gamma}\}$ by comparing the end
+point $\vec{y}$ of the minimization to each of the points already in
+$\{\hat{\Gamma}\}$.  For each $\vec{M}$ in $\{\hat{\Gamma}\}$, APS calculates the mid-point
+\begin{equation}
+\label{eqn:midpt}
+\vec{m}=0.5\times\left(\vec{y}+\vec{M}\right)
+\end{equation}
+and $\vec{M}$.  If, for all $\vec{M}$, 
+$\chi^2$ at $\vec{m}$ is greater
+than $\chi^2_\text{lim}$, then it is deemed that $\vec{\gamma}$ is a new local
+minimum and is added to $\{\hat{\Gamma}\}$.  If not, $\vec{\gamma}$ is assumed to be a
+part of the one of the previously identified loci of low $\chi^2$ and it is not
+added to $\{\hat{\Gamma}\}$.  Regardless of whether or not $\vec{y}$ passes the test in
+equation (\ref{eqn:midpt}), $\vec{y}$ is added to the set $\{\hat{Y}\}$ of all
+simplex end points discovered by APS.
+
+As a further guard against excessive simplex searching, all points $\vec{w}$ 
+that pass the test in equation(\ref{eqn:simplexCriterion}) are further required
+to satisfy the condition
+\begin{eqnarray}
+\chi^2(\vec{z})&\le&\chi^2(\vec{w})-0.25\times\left(\chi^2(\vec{w})-\chi^2(\vec{y})\right)\\
+\vec{z}&\equiv&0.5\times\left(\vec{w}+\vec{y}\right)\\
+\vec{y}&\in&\{\hat{Y}\}
+\end{eqnarray}
+before it is allowed to seed a new simplex search.  In this way, our APS avoids seeding simplex
+searches with points that are inside of already known regions of low $\chi^2$.
+
+\subsection{Code implementing APS}
 \label{sec:user}
 
-We make code implementing APS for an arbitrary $\vec{\theta}\rightarrow\chi^2$
+We make code implementing our extended APS for 
+an arbitrary $\vec{\theta}\rightarrow\chi^2$
 function available at \verb|https://github.com/uwssg/APS/|.
-Below, we present a list of user-specified parameters 
-required to run the algorithm.  Documentation is also provided with
-the code.
+The code is written in C++.  It provides the option for several
+user-specified parameters.  We list them below along with the C++ 
+methods (belonging to the class \verb|aps|) used to set them.  
+This code includes several user-specified parameters which we
+outline below.
 
 \begin{itemize}
-\item $N_s$ -- the number of randomly selected points selected in step (1A).
-\\
-
-\item $\chi^2_\text{lim}$ -- the target threshold value of $\chi^2$.  Alternatively,
-the user can specify
-\\
-\item $\Delta \chi^2$ -- a threshold such that APS sets
-$\chi^2_\text{lim}=\chi^2_\text{min}+\Delta\chi^2$ where $\chi^2_\text{min}$ is the
-minimum value of $\chi^2$ discovered by APS at any given time.  This is the framework
-adopted in this paper.
-\\
-
-\item $N_c$ -- the number of candidate points considered in setp (2A).  Recall
-that the code will only evaluate $\vec{\theta}\rightarrow\chi^2$ on the
-candidate point which maximizes equation (\ref{eq:sstat}).
-\\
-
-\item $N_g$ -- in order to speed up step (2A), the code
-as presented only uses the $N_g$ nearest neighbors (in parameter space) of
-each candidate point as inputs to prediction $\mu(\vec{\theta})$.
-\\
-
-\item $R_g$ -- (\verb|grat| in the code) the ratio determining whether a point
-will be considered a candidate for initiating a gradient descent search. 
-Gradient descent will only be considered for points for which
-$$\chi^2-\chi^2_\text{lim}\le R_g\times \chi^2_\text{lim}$$
-\\
-
-\item $N_w$ -- the maximum number of gradient descent searches that the code
-will allow to go on in parallel with the ``usual'' APS search at one time.  A
-gradient descent search is terminated once it has either
-found a point which satisfies $\chi^2\le\chi^2_\text{lim}$ or 
-has reduced its step
-size such that it is effectively no longer moving through parameter space.  
-If all $N_w$ gradient descent slots are already full when another gradient descent
-candidate is found, the code will add the new
-gradient descent only if its present value of $\chi^2$ is closer to
-$\chi^2_\text{lim}$ than the farthest active gradient descent.  In that case,
-the farthest active gradient descent will be terminated.
-\\
-
-\item $\{\text{min}_i,\text{max}_i\}$ -- 
-the minimum and maximum values of the $N_p$ parameters over which you want APS
-to search.
+\item The parameter $G$ from equation (\ref{eqn:threshold}) above can be set using the method
+\verb|aps.set_grat|
+\\
+\item $\chi^2_\text{lim}$ can be manually set using the method \verb|aps.set_target|
+\\
+\item $\Delta \chi^2$ can be set in the constructor for the class \verb|aps|.
+\\
+\item The number $N_g$ of nearest neighbors used in step 2A can be set in the constructor
+for the class \verb|aps|
+\\
+\item $N_c$ from step 2A can be set using the method \verb|aps.set_samples|
+\\
+\item The minimum and maximum bounds in parameter space can be set using the 
+method \verb|aps.set_max| and \verb|aps.set_min|.  These bounds will constrain the range of
+points sampled in step 2A.  It will not contraint searches performed with bisection or the 
+simplex minimizer.
+\\
+\item If there is a characteristic length scale associated with a parameter, this can be set
+using the method \verb|aps.set_characteristic_length|.  This length scale will be used to
+normalize all distances in parameter space.  If it is not set, it will default to the
+difference between the maximum and minimum values set by \verb|aps.set_max| and
+\verb|aps.set_min|.
 \end{itemize}
+
+More complete documentation is made available with the source code.
+
+\subsection{Interpreting APS outputs}
+\label{sec:bayes}
+
+Originally, APS was conceived as a way to draw Frequentist confidence limits. 
+$\chi^2_\text{lim}$ is determined, either in a vacuum or as some $\Delta\chi^2$ away from
+$\chi^2_\text{min}$ (see Appendix \ref{sec:frequentism} for a more thorough discussion).
+This is by far the most direct way to interpret APS outputs: a list of points is given with
+corresponding values of $\chi^2$.  All of the points with $\chi^2\le\chi^2_\text{lim}$ are
+taken to circumscribe the desired confidence limit.  There is, however, also a way to use the
+outputs from APS to draw a Bayesian credible limit.
+
+As discussed in the introduction, Bayesian parameter constraints assume that there is some
+probability distribution (``the posterior'') over $\vec{\theta}$.  
+$(1-\alpha)\%$ credible limits are drawn by
+integrating over this probability distribution and selecting that region which contains
+$(1-\alpha)\%$ of the total probability.  This integral can be very time-consuming if the
+dimensionality of $\vec{\theta}$ is large.  Markov Chain Monte Carlo methods overcome this
+difficulty by randomly drawing samples from the posterior and assuming that, if enough samples
+are drawn, the distribution of those samples will be sufficient to reconstruct the attributes
+of the full posterior probability density.
+
+APS outputs cannot be used this way, since APS does not select points to evaluate
+probabilisticaly, but in an attempt to recreate the behavior of the $\chi^2$ function in
+interesting regions of parameter space.  We can still, however, use these points to approximate
+the integral over the posterior probability density.  The algorithm we propose is as follows.
+The total set of points sampled by APS will be referred to below as $\{\hat{A}\}$.
+
+\begin{itemize}
+\item(1D) For each point $\vec{a}$ in $\{\hat{A}\}$ initialize two vectors $\vec{a}_\text{max}$ and
+$\vec{a}_\text{min}$.
+
+\item(2D) Select the $3N_p+1$ nearest neighbor points to $\vec{a}$ from
+$\{\hat{A}\}$ (this will obviously include $\vec{a}$ itself, which can be discarded).  Call
+this set of nearest neighbors $\{\hat{N}\}$.
+\\
+\item(3D) For each point $\vec{n}$ in $\{\hat{N}\}$, calculate the vector
+$\vec{\delta}=\|\vec{n}-\vec{a}|$ (where the vertical bars denote the absolute value of the
+components of $\vec{\delta}$, not any kind of vector normalization).  For each component
+$\delta_i$ of $\vec{\delta}$, starting with the largest and working towards the smallest:
+\\
+\begin{itemize}
+\item(3Da) If $n_i>a_i$ and $a_{\text{max},i}$ has not been set, set $a_{\text{max},i}=n_i$.
+\\
+\item(3Db) If $n_i<a_i$ and $a_{\text{min},i}$ has not been set, set $a_{\text{min},i}=n_i$.
+\\
+\item(3Dc) If both $a_\text{max,i}$ and $a_\text{min,i}$ have already been set, move on to the
+next largest component of $\vec{\delta}$ and repeat at step (3Da).
+\end{itemize}
+\item(4D) Choose the next neighbor $\vec{n}$ and repeat from step (3D). Continue until all
+$a_{\text{max},i}$ and $a_{\text{min},i}$ have been set.  Once all $a_{\text{max},i}$ and
+$a_{\text{min},i}$ have been set, move on to step (5D).
+\\
+\item(5D) $\vec{a}_\text{max}$ an $\vec{a}_\text{min}$ should now describe an asymmetric,
+rectangular hyperbox about $\vec{a}$.  To symmetrize it, step through all $N_p$ dimensions
+and reset either $a_{\text{max},i}$ or $a_{\text{min},i}$ so that 
+$d_\text{min}=a_i-a_{\text{min},i}$ and
+$d_\text{max}=a_{\text{max},i}-a_i$ are equal to the minimum of $\{d_\text{max},d_\text{min}\}$
+from the asymmetric hyperbox.
+\end{itemize}
+
+At this point, each $\vec{a}$ in $\{\hat{A}\}$ should have a corresponding symmetric hyperbox
+surrounding it.  To integrate the posterior probability density, we will assume that all of the
+parameter space points in each hyperbox correspond to a $\chi^2$ equal to $\chi^2(\vec{a})$.
+The posterior probability associated with each box is then
+\begin{eqnarray}
+P(\vec{a})&=&\frac{V(\vec{a})}{P_\text{tot}}\times\exp\left[-\chi^2(\vec{a})/2\right]
+\label{eqn:posterior}\\
+P_\text{tot}&\equiv&\sum_{\vec{a}} V(\vec{a})\times\exp\left[-\chi^2(\vec{a})/2\right]
+\end{eqnarray}
+where $V(\vec{a})$ is the hypervolume associated with the hyperbox surrouning $\vec{a}$.
+To plot the $(1-\alpha)\%$ credible limit, one need only plot the points in hyperspace that
+enclose the lowest-$\chi^2$ hyperboxes containing $(1-\alpha)\%$ of the total probability.
+We test this algorithm on a function with an exactly known credible limit
+in Section \ref{sec:toy} and on a real, physical dataset in Section \ref{sec:wmap}.  In 
+Section \ref{sec:toy}, we find that our results agree with the known credible limit.
+In Section \ref{sec:wmap}, we find that our results agree with results derived from traditional
+Markov Chain Monte Carlo methods.
+
+The functionality described by this section is provided in by the classes defined in the source
+code file \verb|aps_extractor.cpp| in our software package.
 
@@
 provide well-motivated prediction uncertainties $\sigma$, a requirement of
-equation (\ref{eq:sstat}).  We present below an introduction to the
+equation (\ref{eqn:sstat}).  We present below an introduction to the
 formalism of Gaussian processes drawn heavily
@@
 $C_{ij}(\vec{\theta}^{(i)},\vec{\theta}^{(j)})$.
-If we have $N_d$ measurements of $\chi^2$, then we assume
+If we have $N_g$ measurements of $\chi^2$, then we assume
 they are distributed according to
 \begin{equation}
-\label{eq:likelihood}
+\label{eqn:likelihood}
 P(\vec{\chi}^2)=
-\frac{\exp\bigg[-\frac{1}{2}K^{-1}\sum_{i,j}^{N_d}
+\frac{\exp\bigg[-\frac{1}{2}K^{-1}\sum_{i,j}^{N_g}
 (\chi^{2,(i)}-\bar{\chi}^2)C^{-1}_{ij}
-(\chi^{2,(j)}-\bar{\chi}^2)\bigg]}{\sqrt{2\pi}^{N_d}\text{det}|KC|^{N_d/2}}
+(\chi^{2,(j)}-\bar{\chi}^2)\bigg]}{\sqrt{2\pi}^{N_g}\text{det}|KC|^{N_g/2}}
 \end{equation}
@@
 \begin{equation}
-\label{eq:covraw}
+\label{eqn:GaussianCovar}
 C_{ij}=\exp\left[-\frac{1}{2}D^2_{ij}\right]
@@
 \begin{equation}
-D^2_{ij}\equiv\sum_n^{N_p}\left(\frac{\theta^{(i)}_n-\theta^{(j)}_n}{\text{max}_n-\text{min}_n}\right)^2
+D^2_{ij}\equiv\sum_n^{N_p}
+\left(\frac{\theta^{(i)}_n-\theta^{(j)}_n}{\ell_n\times(\text{max}_n-\text{min}_n)}\right)^2
 \end{equation}
-We discuss the normalization $K$ below.
+where $\text{max}_n-\text{min}_n$ is the difference between the maximum and minimum values of
+the $n$th parameter taken from the $N_g$ points used to seed the Gaussian process and $\ell_n$
+is a ``hyper-parameter'' associated with the $n$th parameter.  We both the normalization
+paramter $K$ below in section \ref{sec:kriging}.  We discuss the setting of
+hyper-parameters in section \ref{sec:hyperparams}.
 The exponential form of $C_{ij}$ quantifies the assumption 
@@
 C_{ij}=\phantom{\sin
-\bigg[\frac{2(1+\tilde{\vec{\theta}}^{(i)}\cdot\tilde{\vec{\theta}}^{(j)})}
-{\sqrt{(1+2(1+\tilde{\vec{\theta}}^{(i)}\cdot\tilde{\vec{\theta}}^{(i)}))
-(1+2(1+\tilde{\vec{\theta}}^{(j)}\cdot\tilde{\vec{\theta}}^{(j)}))}}\bigg]}
-\label{eq:cov_nn}\\
+\bigg[\frac{2(\sigma_0+\tilde{\vec{\theta}}^{(i)}\cdot\tilde{\vec{\theta}}^{(j)})}
+{\sqrt{(1+2(\sigma_1+\tilde{\vec{\theta}}^{(i)}\cdot\tilde{\vec{\theta}}^{(i)}))
+(1+2(\sigma_1+\tilde{\vec{\theta}}^{(j)}\cdot\tilde{\vec{\theta}}^{(j)}))}}\bigg]}
+\label{eqn:NNCovar}\\
 \frac{2}{\pi}\sin^{-1}
-\bigg[\frac{2(1+\tilde{\vec{\theta}}^{(i)}\cdot\tilde{\vec{\theta}}^{(j)})}
-{\sqrt{(1+2(1+\tilde{\vec{\theta}}^{(i)}\cdot\tilde{\vec{\theta}}^{(i)}))
-(1+2(1+\tilde{\vec{\theta}}^{(j)}\cdot\tilde{\vec{\theta}}^{(j)}))}}\bigg]
+\bigg[\frac{2(\sigma_0+\tilde{\vec{\theta}}^{(i)}\cdot\tilde{\vec{\theta}}^{(j)})}
+{\sqrt{(1+2(\sigma_1+\tilde{\vec{\theta}}^{(i)}\cdot\tilde{\vec{\theta}}^{(i)}))
+(1+2(\sigma_1+\tilde{\vec{\theta}}^{(j)}\cdot\tilde{\vec{\theta}}^{(j)}))}}\bigg]
 \nonumber
@@
 parameters $\vec{\theta}$ recentered and renormalized relative to the span
-$(\text{max}_n-\text{min}_n)$ in each dimension.  We also consider this
+$(\text{max}_n-\text{min}_n)$ in each dimension.  $\sigma_0$ and $\sigma_1$
+are hyperparameters.  We also consider this
 covariogram in Section \ref{sec:wmap} and find no appreciable difference
-in performance when compared to the use of covariogram (\ref{eq:covraw}).
-
-The normalization constant $K$ in equation (\ref{eq:likelihood}) --known as the
+in performance when compared to the use of covariogram (\ref{eqn:GaussianCovar}).
+
+\subsection{Normalizing the covariogram}
+\label{sec:kriging}
+
+The normalization constant $K$ in equation (\ref{eqn:likelihood}) --known as the
 ``Kriging parameter'' for the geophysicist who pioneered the overall method -- 
@@
 
-Recall from equation (\ref{eq:likelihood}) that we are treating our
+Recall from equation (\ref{eqn:likelihood}) that we are treating our
 input data as though they were samples drawn from a probability distribution.
@@
 \begin{equation}
-\label{eq:kriging}
+\label{eqn:kriging}
 K=\frac{\sum_{i,j}^{N_g}(\chi^{2,(i)}-\bar{\chi}^2)
@@
 Figure \ref{fig:gp} applies the Gaussian process of equations (\ref{sig}), 
-(\ref{mu}), and (\ref{eq:kriging}) with covariogram (\ref{eq:covraw})
+(\ref{mu}), and (\ref{eqn:kriging}) with covariogram (\ref{eqn:GaussianCovar})
 to a toy one-dimensional function.  Inspection shows many desirable
@@
 Far from the $\theta^{(i)}$, $\sigma$ is larger, and the $S$ statistic in
-equation (\ref{eq:sstat}) will induce the \APS
+equation (\ref{eqn:sstat}) will induce the \APS
 algorithm to examine the true value of $\chi^2$.
@@
 2-$\sigma$ uncertainty bounds.  We set $K$ according to 
-equation (\ref{eq:kriging}) wth $N_g=5$ for the 5 sampled points.
+equation (\ref{eqn:kriging}) wth $N_g=5$ for the 5 sampled points.
 }
@@
 \end{figure}
+
+\subsection{Setting hyper-parameters}
+\label{sec:hyperparams}
+
+The covariance functions $C_{ij}$ presented above (indeed, any covariance function)
+each contain arbitrary hyper-parameters whose values can be set to fine-tune
+the Gaussian Process model ($\ell_n$ in equation \ref{eqn:GaussianCovar}
+and $\{\sigma_0,\sigma_1\}$ in equation \ref{eqn:NNCovar}).  
+In order to set the hyper-parameters used in the APS Gaussian Process, we randomly
+select a set $\{\hat{V}\}$ of up to 3000 points from those already sampled by APS 
+and use these to construct
+the error function $E(\vec{h})$ where $\vec{h}$ is a vector specifying the hyper-parameters
+required by the chosen $C_{ij}$.  $E$ is defined as
+\begin{equation}
+\label{eqn:hyperparamError}
+E(\vec{h})=\sum_{\vec{v}\in\{\hat{V}\}}\left(\mu(\vec{v},\vec{h})-\chi^2(\vec{v})\right)^2
+\end{equation}
+where $\mu(\vec{v},\vec{h})$ is calculated according to equation \ref{mu} using
+hyper-parameters $\vec{h}$ (and ignoring $\vec{v}$ as its own nearest neighbor) and
+$\chi^2(\vec{v})$ is already known, since $\vec{v}$ has been chosen from the points previously
+sampled by APS.  If $\vec{h}$ has more than two dimensions, we minimize $E$ on $\vec{h}$-space
+using the simplex
+minimizer of Nelder and Mead (1965).  Otherwise, we do a simple grid search for the value of
+$\vec{h}$ which minimizes $E$.  This optimal value of $\vec{h}$ is used to set
+the hyper-parameters for our Gaussian Process model.
+
+As APS samples more points in parameter space, we expect it to learn more about the
+$\vec{\theta}\rightarrow\chi^2$ function and thus be capable of a more accurate Gaussian
+Process, so we allow it to periodically reset the hyper-parameters $\vec{h}$.  However,
+minimizing $E$ can be a very time-consuming process, so we have APS keep track of the amount of
+time elapsed since the algorithm began.  We divide this value by the number of calls made to
+the $\vec{\theta}\rightarrow\chi^2$ function and compare this amortized time to the average
+amount of time spent on one call to $\vec{theta}\rightarrow\chi^2$.  APS us only allowed to
+reset its hyper-parameters if the difference between these two times is less than 0.1 second.
 
@@
 therefore, we will use for our test a toy $\chi^2$ function
-with multiple locii of low $\chi^2$.  We assume an $N_p=6$ dimensional
-$\vec{\theta}$ and a $\chi^2$ that goes as
+with multiple locii of low $\chi^2$.
+
+We construct our toy $\chi^2$ as a series of low-$\chi^2$ ellipses in a 5-dimensional
+parameter space (we choose a 5-dimensional parameter space because it allows us to
+reasonably show all of the 2-dimensional sub-spaces in the plots which follow).
+We consider functions with 2, 3, and 4 discrete minima in $\chi^2$.
+The location of the low-$\chi^2$ regions are generated using a pseudo random number
+generator taken from Marsaglia (2003) with parameters taken from Press {\it et al.} (2007).
+$\chi^2(\vec{\theta})$ for this toy function is calculated as 
 \begin{equation}
-\chi^2=1300+\frac{d_1^2}{2}+\frac{d_2^2}{2}-153\exp[-2d_1^2]-100\exp[-d_2^2]
-\label{eq:toychisq}
+\label{eqn:toychi}
+\chi^2(\vec{\theta})=
+\sum_i\left(\frac{\theta_i-c_{\text{nearest},i}}{w_{\text{nearest},i}}\right)^2
 \end{equation}
-where $d_1$ is the Euclidean parameter-space distance from 
-$\vec{\theta}=\{3,0,0,0,0,0\}$ and
-$d_2$ is the Euclidean parameter-space distance from 
-$\vec{\theta}=\{-3,0,0,0,0,0\}$.  
-Figure \ref{fig:toychisq} plots this $\chi^2$ as a function of $\theta_0$ with
-all other parameters $\theta_i=0$.
-This $\chi^2$ function has three features that
-make it tricky for MCMC algorithms to handle.  It has two locii of low $\chi^2$
-which are disjoint.  Those locii are narrow compared to the full
-$N_p$-dimensional extent of the function (we choose to explore
-$-10\le\theta_i\le10$).  There is a false minimum in $\chi^2$ near
-$\vec{\theta}=\{0,0,0,0,0,0\}$.
-
-\begin{figure}
-\includegraphics[scale=0.3]{toychi_plot.eps}
+where $c_{\text{nearest},i}$ is the $i$th coordinate of the center of the nearest low-$\chi^2$
+region and $w_{\text{nearest},i}$ is the (pseudo-randomly generated) width of that ellipse in
+the $i$th coordinate.  We refrain from introducing any interesting parameter degeneracies so
+that the $\chi^2(\vec{\theta})$ function remains integrable in 2-dimensional sub-spaces.
+This gives us a hard-truth ``control'' against which to evaluate our test APS runs below.
+Readers wishing to re-create this $\chi^2$ function can find it in our software
+package as the class \verb|ellipses_integrable| in the source code files \verb|chisq.h| and
+\verb|chisq.cpp|.
+
+To test the performance of our APS algorithm, we ran APS on iterations of this
+$\vec{\theta}\rightarrow\chi^2$ function with 2, 3, and 4 discrete minima in $\chi^2$.  We
+allowed APS to sample a total of 10,000 points in the 5-dimensional parameter space.  
+We had APS adaptively set $\chi^2_\text{lim}=\chi^2_\text{min}+11.0$ -- $11.0$ being the $95\%$
+confidence limit bound for $\chi^2$ probability distribution with 5 degrees of freedom.
+Figures
+\ref{fig:toyFreq2}, \ref{fig:toyFreq3}, and \ref{fig:toyFreq4} plot the
+$\chi^2\le\chi^2_\text{lim}$ points found by APS (black points) against the
+known $\Delta\chi^2=11.0$ contours of the function (red contours).  Obviously, APS did a very
+good job of exploring all of the available low-$\chi^2$ regions.
+
+\begin{figure*}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c2_0_1_frequentist.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c2_0_2_frequentist.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c2_0_3_frequentist.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c2_0_4_frequentist.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c2_1_2_frequentist.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c2_1_3_frequentist.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c2_1_4_frequentist.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c2_2_3_frequentist.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c2_2_4_frequentist.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c2_3_4_frequentist.eps}
+}
 \caption{
-A one-dimensional slice of the toy $\chi^2$ function equation
-(\ref{eq:toychisq}).  The other five parameters are set to zero.
-The red line indicates the value $\chi^2=1281$.
-}
-\label{fig:toychisq}
-\end{figure}
-
-To test the performance of APS versus MCMC exploring equation
-(\ref{eq:toychisq}), we run 200 instances of each algorithm on the function.
-We run APS including the gradient descent modification introduced at the end of
-section \ref{sec:algorithm}.  
-Each instantiation of APS is begun with parameters (see Section \ref{sec:user})
-\begin{eqnarray}
-\Delta\chi^2&=&12.59\nonumber\\
-R_g&=&0.1\nonumber\\
-N_w&=&50\nonumber\\
-N_c&=&250\nonumber\\
-N_g&=&15\nonumber
-\end{eqnarray}
-
-Each instantiation of MCMC is comprised of four
-independent chains, each of which is allowed to sample up to 250,000 points
-in the parameter space.  Initially, the chains propose steps randomly
-in the cardinal directions of the parameter space.  The steps in each direction
-are drawn from a Gaussian 
-distribution with standard deviation $2.0$.
-After 4,000 steps have been taken, the MCMC approximates the covariance matrix of
-the steps taken so far and adjusts its proposal algorithm so that it proposes steps
-along the eigen vectors of this matrix, with characteristic step size equal to
-the $(2.38)/\sqrt{N_p}$ times the
-square root of the corresponding eigen values ($N_p=6$ is the
-dimensionality of our parameter space) \cite{gelman}.  
-The code reassesses this proposal distribution after 
-every subsequent 2,000 steps.
-
-Figure \ref{fig:toyneg} shows how many instantiations
-of each algorithm successfully found 
-the low $\chi^2$ region centered on $\theta_0=-3$
-as a function of the number of
-sampled points it took them to do so.  
-Figure \ref{fig:toypos} shows the same
-for the $\theta_0=3$  minimum.  The solid black curve represents
-APS.  The dashed red curve
-represents MCMC.  
-For the purposes of these Figures, ``finding'' a minimimum means evaluating
-$\vec{\theta}\rightarrow\chi^2$ at a point where $\chi^2\le1281$ near either
-$\vec{\theta}=\{-3,0,\dots\}$ or $\vec{\theta}=\{3,0,\dots\}$.
-Though APS and MCMC find the minimum at $\{-3,0,\dots\}$
-(which is wider in parameter space)
-with comparable rapidity, MCMC only finds the $\{3,0,\dots\}$
-minimum  50\% of the time.  The other 50\% of the time, 
-MCMC is complacent having
-found the other minimum and calibrates itself to thoroughly exploring
-that region of low $\chi^2$.
-Because APS selects points based on how interesting they are
-likely to be -- i.e. equation (\ref{eq:sstat}) -- APS finds both regions of low
-$\chi^2$ 100\% of the time.
-We also considered the performance of a uniform random sampling of our parameter
-space.  In 1 million samplings, only 9 of the
-200 instantiations found the $\{-3,0,\dots\}$ minimum.  
-Only 5 instantiations found the $\{3,0,\dots\}$ minimum.
-
-\begin{figure}
-\subfigure[]{
-\includegraphics[scale=0.3]{udder_found_n.eps}
-\label{fig:toyneg}
-}
-\subfigure[]{
-\includegraphics[scale=0.3]{udder_found_p.eps}
-\label{fig:toypos}
+This plot shows all 10 2-dimensional sub-spaces of our 5-dimensional toy likelihood
+function with 2 minima in $\chi^2$.  The black points are $\chi^2\le\chi^2_\text{lim}$
+found by APS after 10,000 samplings.  The red contours are the known $\chi^2=\chi^2_\text{lim}$
+contours of the function.
+}
+\label{fig:toyFreq2}
+\end{figure*}
+
+\begin{figure*}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c3_0_1_frequentist.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c3_0_2_frequentist.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c3_0_3_frequentist.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c3_0_4_frequentist.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c3_1_2_frequentist.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c3_1_3_frequentist.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c3_1_4_frequentist.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c3_2_3_frequentist.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c3_2_4_frequentist.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c3_3_4_frequentist.eps}
 }
 \caption{
-We run MCMC and APS on a 6-dimensional parmeter space with a $\chi^2$ function
-given by equation (\ref{eq:toychisq}).
-The horizontal axis represents the number of points sampled by each algorithm.
-The vertical axis represents the number of instantiations of each algorithm that
-had found the high likelihood region at $\{-3,0,0,0,0,0\}$ (Figure
-\ref{fig:toyneg}) and $\{3,0,0,0,0,0\}$ (Figure \ref{fig:toypos}) in that
-many iterations.
-}
-\label{fig:toy}
-\end{figure}
+This plot shows all 10 2-dimensional sub-spaces of our 5-dimensional toy likelihood
+function with 3 minima in $\chi^2$.  The black points are $\chi^2\le\chi^2_\text{lim}$
+found by APS after 10,000 samplings.  The red contours are the known $\chi^2=\chi^2_\text{lim}$
+contours of the function.
+}
+\label{fig:toyFreq3}
+\end{figure*}
+
+\begin{figure*}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c4_0_1_frequentist.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c4_0_2_frequentist.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c4_0_3_frequentist.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c4_0_4_frequentist.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c4_1_2_frequentist.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c4_1_3_frequentist.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c4_1_4_frequentist.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c4_2_3_frequentist.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c4_2_4_frequentist.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c4_3_4_frequentist.eps}
+}
+\caption{
+This plot shows all 10 2-dimensional sub-spaces of our 5-dimensional toy likelihood
+function with 4 minima in $\chi^2$.  The black points are $\chi^2\le\chi^2_\text{lim}$
+found by APS after 10,000 samplings.  The red contours are the known $\chi^2=\chi^2_\text{lim}$
+contours of the function.
+}
+\label{fig:toyFreq4}
+\end{figure*}
+
+\begin{figure*}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c2_0_1_bayes.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c2_0_2_bayes.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c2_0_3_bayes.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c2_0_4_bayes.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c2_1_2_bayes.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c2_1_3_bayes.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c2_1_4_bayes.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c2_2_3_bayes.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c2_2_4_bayes.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c2_3_4_bayes.eps}
+}
+\caption{
+This plot shows all 10 2-dimensional sub-spaces of our 5-dimensional toy likelihood
+function with 2 minima in $\chi^2$.  The black points represent the $95\%$ Bayesian credible
+limit found by APS (and determined as described in Section \ref{sec:bayes}) after 10,000
+samplings.  The red contours represent the known $95\%$ Bayesian credible limit of the toy
+function.
+}
+\label{fig:toyBayes2}
+\end{figure*}
+
+\begin{figure*}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c3_0_1_bayes.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c3_0_2_bayes.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c3_0_3_bayes.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c3_0_4_bayes.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c3_1_2_bayes.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c3_1_3_bayes.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c3_1_4_bayes.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c3_2_3_bayes.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c3_2_4_bayes.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c3_3_4_bayes.eps}
+}
+\caption{
+This plot shows all 10 2-dimensional sub-spaces of our 5-dimensional toy likelihood
+function with 3 minima in $\chi^2$.  The black points represent the $95\%$ Bayesian credible
+limit found by APS (and determined as described in Section \ref{sec:bayes}) after 10,000
+samplings.  The red contours represent the known $95\%$ Bayesian credible limit of the toy
+function.
+}
+\label{fig:toyBayes3}
+\end{figure*}
+
+\begin{figure*}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c4_0_1_bayes.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c4_0_2_bayes.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c4_0_3_bayes.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c4_0_4_bayes.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c4_1_2_bayes.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c4_1_3_bayes.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c4_1_4_bayes.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c4_2_3_bayes.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c4_2_4_bayes.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{ellipse_d5_c4_3_4_bayes.eps}
+}
+\caption{
+This plot shows all 10 2-dimensional sub-spaces of our 5-dimensional toy likelihood
+function with 4 minima in $\chi^2$.  The black points represent the $95\%$ Bayesian credible
+limit found by APS (and determined as described in Section \ref{sec:bayes}) after 10,000
+samplings.  The red contours represent the known $95\%$ Bayesian credible limit of the toy
+function.
+}
+\label{fig:toyBayes4}
+\end{figure*}
 
@@
 \end{eqnarray}
-We will consider both equations (\ref{eq:covraw}) and (\ref{eq:cov_nn})
+We will consider both equations (\ref{eqn:covraw}) and (\ref{eqn:cov_nn})
 as covariograms, though we find that the choice has little effect on our
@@
 points in the case where $\chi^2_\text{lim}\equiv\chi^2_\text{min}+12.59$
-using equations (\ref{eq:covraw}) and (\ref{eq:cov_nn}) as covariograms,
+using equations (\ref{eqn:covraw}) and (\ref{eqn:cov_nn}) as covariograms,
 respectively.
@@
 points found by APS after sampling a total of 100,000 points.
-These plots use equation (\ref{eq:covraw}) for the covariogram.
+These plots use equation (\ref{eqn:covraw}) for the covariogram.
 }
@@
 points found by APS after sampling a total of 100,000 points.
-These plots use equation (\ref{eq:cov_nn}) for the covariogram.
+These plots use equation (\ref{eqn:cov_nn}) for the covariogram.
 }
@@
 
+\bibitem[Marsaglia 2003]{random}
+Marsaglia,~G. 2003, Journal of Statistical Software, {\bf 8} 1
+
 \bibitem[Nelder and Mead 1965]{simplex}
@@
 
+\bibitem[Press {\it et al}. 2007]{nr}
+Press,~W.~H., Teukolsky,~S.~A., Vetterling,~W.~T. and Flannery,~B.~P. 2007,
+``Numerical Recipes'' (3d edition), Cambridge University Press, 2007
+
 \bibitem[Rasmussen and Williams 2006]{gp}

### 2014-07-21 aps4p5/kde.cpp
futzing around with mcmc_extraction_runner

@@ -54,7 +54,7 @@ int kde::get_dex(array_1d<double> &pt, int ixx, double dxx){
 void kde::initialize_density(int ix1_in, double dx1_in, 
            int ix2_in, double dx2_in, int smoothby_in){
     
-    printf("\n\nINITIALIZING DENSITY\n\n");
+    //printf("\n\nINITIALIZING DENSITY\n\n");
     
     grid.reset();
     grid_wgt.reset();

### 2014-07-21 aps4p5/mcmc_extraction_runner.cpp

@@ -6,21 +6,14 @@ main(){
 
 mcmc_extractor test;
 
-int nparams=22;
-
-//nparams=46;
-
+int nparams;
 nparams = 81;
 
 test.set_nchains(4);
 test.set_nparams(nparams);
 
-test.set_cutoff(2500);
-
-
-//test.set_chainname("/Users/noldor/physics/recreate_getdist/planck_chains/planckTESTgibbs_chain");
-//test.set_chainname("chains/test_nogibbs_chains");
-//test.set_chainname("/Users/noldor/physics/planckLikelihood/base_planck_lowl/base/planck_lowl/base_planck_lowl");
+//test.set_cutoff(7500);
+//control has a total 453,051 samples
 
 test.set_chainname("/Users/noldor/physics/recreate_getdist/ieuchains_1304/wmap7_learn");
 
@@ -31,8 +24,9 @@ test.learn_thinby();
 printf("independent samples %d\n",test.get_nsamples());
 printf("thinby %d\nused %d\nkept %d\n",test.get_thinby(),test.get_total_used(),test.get_total_kept());
 printf("rows %d\n",test.get_total_rows());
+printf("best_covar %e\n",test.get_best_covar());
 
-test.print_samples("test_wmap7_samples10k.sav");
+test.print_samples("test_wmap7_samplescontrol.sav");
 
 kde kde_test;
 kde_test.set_data(test.get_samples());
@@ -63,14 +57,14 @@ for(i=0;i<6;i++){
     for(j=i+1;j<6;j++){
         if(i!=3 && j!=3){
  
-           sprintf(outname,"mcmc_processed/wmap_%d_%d_10k.sav",i,j);
+           sprintf(outname,"mcmc_processed/wmap_%d_%d_control.sav",i,j);
 
             kde_test.plot_boundary(ix.get_data(i),dx.get_data(i),ix.get_data(j),dx.get_data(j),0.95,outname,3);
         }
     }
 }
 
-test.plot_delta("mcmc_good_pts_test10k.sav",0.5*12.61);
+test.plot_delta("mcmc_good_pts_testcontrol.sav",0.5*12.61);
 
 
 array_1d<double> RR,VV,WW,mean,var;

### 2014-07-21 aps4p5/mcmc_extraction_runner.cpp
just fiddling around

@@ -12,12 +12,12 @@ nparams = 81;
 test.set_nchains(4);
 test.set_nparams(nparams);
 
-//test.set_cutoff(7500);
+test.set_cutoff(7500);
 //control has a total 453,051 samples
 
 test.set_chainname("/Users/noldor/physics/recreate_getdist/ieuchains_1304/wmap7_learn");
 
-test.set_keep_frac(0.5);
+test.set_keep_frac(0.9);
 
 test.learn_thinby();
 
@@ -26,7 +26,7 @@ printf("thinby %d\nused %d\nkept %d\n",test.get_thinby(),test.get_total_used(),t
 printf("rows %d\n",test.get_total_rows());
 printf("best_covar %e\n",test.get_best_covar());
 
-test.print_samples("test_wmap7_samplescontrol.sav");
+test.print_samples("test_wmap7_samples30ktest.sav");
 
 kde kde_test;
 kde_test.set_data(test.get_samples());
@@ -57,14 +57,14 @@ for(i=0;i<6;i++){
     for(j=i+1;j<6;j++){
         if(i!=3 && j!=3){
  
-           sprintf(outname,"mcmc_processed/wmap_%d_%d_control.sav",i,j);
+           sprintf(outname,"mcmc_processed/wmap_%d_%d_30ktest.sav",i,j);
 
             kde_test.plot_boundary(ix.get_data(i),dx.get_data(i),ix.get_data(j),dx.get_data(j),0.95,outname,3);
         }
     }
 }
 
-test.plot_delta("mcmc_good_pts_testcontrol.sav",0.5*12.61);
+test.plot_delta("mcmc_good_pts_test30ktest.sav",0.5*12.61);
 
 
 array_1d<double> RR,VV,WW,mean,var;

### 2014-07-21 aps4p5/aps_extraction_runner.cpp
fine-tuned aps_extraction_runner.cpp

@@ -7,14 +7,23 @@ test.set_filename("aps_output/WMAP/apsWMAP_unitSphere_NNLearn_output.sav");
 test.set_delta_chi(12.6);
 //test.set_delta_chi(100.0);
 
-test.set_cutoff(30000);
+test.set_cutoff(10000);
 
-test.write_good_points("aps_processed/apsSphere_good_points.sav");
+test.write_good_points("aps_processed/WMAP/frequentist/nnLearn_good_points_10k.sav");
 //test.plot_chimin("aps_chi_min_test.sav");
 //test.sample_posterior("aps_processed/apsSphere_samples.sav",10000);
 
-test.draw_bayesian_bounds("test_dir/apsSphere_lp_bayes_0_2.sav",0,2,0.95);
-test.draw_bayesian_bounds("test_dir/apsSphere_lp_bayes_4_5.sav",4,5,0.95);
+int i,j;
+char outname[letters];
+
+for(i=0;i<6;i++){
+    for(j=i+1;j<6;j++){
+        if(i!=3 && j!=3){
+            sprintf(outname,"aps_processed/WMAP/bayesian/nnLearn_%d_%d_10k_bayes.sav",i,j);
+            test.draw_bayesian_bounds(outname,i,j,0.95);
+        }
+    }
+}
 
 test.show_minpt();
 

### 2014-07-21 aps4p5/aps_extraction_runner.cpp
aps_extractor.cpp now thins the bayesian plots

@@ -19,7 +19,7 @@ char outname[letters];
 for(i=0;i<6;i++){
     for(j=i+1;j<6;j++){
         if(i!=3 && j!=3){
-            sprintf(outname,"aps_processed/WMAP/bayesian/nnLearn_%d_%d_10k_bayes.sav",i,j);
+            sprintf(outname,"aps_processed/WMAP/bayesianTest/nnLearn_%d_%d_10k_bayes.sav",i,j);
             test.draw_bayesian_bounds(outname,i,j,0.95);
         }
     }

### 2014-07-21 aps4p5/aps_extractor.cpp

@@ -474,6 +474,117 @@ void aps_extractor::draw_bayesian_bounds(char *filename, int ix, int iy, double 
     
     double sum=0.0;
     
+    array_2d<double> to_plot;
+    array_1d<double> min,max,center;
+    
+    to_plot.set_cols(2);
+    min.set(0,chisq_exception);
+    min.set(1,chisq_exception);
+    max.set(0,-1.0*chisq_exception);
+    max.set(1,-1.0*chisq_exception);
+    for(i=0;i<l_probability.get_dim() && sum<limit;i++){
+        ibox=l_prob_dexes.get_data(i);
+        sum+=exp(l_probability.get_data(ibox));
+        
+        j=to_plot.get_rows();
+        to_plot.set(j,0,box_max.get_data(ibox,ix));
+        to_plot.set(j,1,box_max.get_data(ibox,iy));
+        
+        j=to_plot.get_rows();
+        to_plot.set(j,0,box_max.get_data(ibox,ix));
+        to_plot.set(j,1,box_min.get_data(ibox,iy));
+        
+        j=to_plot.get_rows();
+        to_plot.set(j,0,box_min.get_data(ibox,ix));
+        to_plot.set(j,1,box_min.get_data(ibox,iy));
+        
+        j=to_plot.get_rows();
+        to_plot.set(j,0,box_min.get_data(ibox,ix));
+        to_plot.set(j,1,box_max.get_data(ibox,iy));
+        
+        if(box_max.get_data(ibox,ix)>max.get_data(0))max.set(0,box_max.get_data(ibox,ix));
+        if(box_max.get_data(ibox,iy)>max.get_data(1))max.set(1,box_max.get_data(ibox,iy));
+        
+        if(box_min.get_data(ibox,ix)<min.get_data(0))min.set(0,box_min.get_data(ibox,ix));
+        if(box_min.get_data(ibox,iy)<min.get_data(1))min.set(1,box_min.get_data(ibox,iy));
+        
+    }
+    
+   
+    center.set(0,0.5*(max.get_data(0)+min.get_data(0)));
+    center.set(1,0.5*(max.get_data(1)+min.get_data(1)));
+    
+    array_1d<double> dd;
+    array_1d<int> dex;
+    double nn;
+    for(i=0;i<to_plot.get_rows();i++){
+        
+        nn=0.0;
+        for(j=0;j<2;j++){
[…]

### 2014-07-21 aps4p5/aps_extractor.cpp
factored out plot_thinned_data into its own subroutine

@@ -468,20 +468,14 @@ void aps_extractor::draw_bayesian_bounds(char *filename, int ix, int iy, double 
         make_boxes();
     }
     
-    FILE *output;
+
     int i,j,ibox;
-    
-    
     double sum=0.0;
     
     array_2d<double> to_plot;
-    array_1d<double> min,max,center;
     
     to_plot.set_cols(2);
-    min.set(0,chisq_exception);
-    min.set(1,chisq_exception);
-    max.set(0,-1.0*chisq_exception);
-    max.set(1,-1.0*chisq_exception);
+
     for(i=0;i<l_probability.get_dim() && sum<limit;i++){
         ibox=l_prob_dexes.get_data(i);
         sum+=exp(l_probability.get_data(ibox));
@@ -502,15 +496,24 @@ void aps_extractor::draw_bayesian_bounds(char *filename, int ix, int iy, double 
         to_plot.set(j,0,box_min.get_data(ibox,ix));
         to_plot.set(j,1,box_max.get_data(ibox,iy));
         
-        if(box_max.get_data(ibox,ix)>max.get_data(0))max.set(0,box_max.get_data(ibox,ix));
-        if(box_max.get_data(ibox,iy)>max.get_data(1))max.set(1,box_max.get_data(ibox,iy));
-        
-        if(box_min.get_data(ibox,ix)<min.get_data(0))min.set(0,box_min.get_data(ibox,ix));
-        if(box_min.get_data(ibox,iy)<min.get_data(1))min.set(1,box_min.get_data(ibox,iy));
         
     }
     
-   
+    plot_thinned_data(to_plot,0.01,filename);    
+}
+
+void aps_extractor::plot_thinned_data(array_2d<double> &to_plot, double tol, char *filename){    
+    
+    array_1d<double> max,min,center;
+    int i,j;
+    for(i=0;i<to_plot.get_rows();i++){
+        if(i==0 || to_plot.get_data(i,0)>max.get_data(0))max.set(0,to_plot.get_data(i,0));
+        if(i==0 || to_plot.get_data(i,1)>max.get_data(1))max.set(1,to_plot.get_data(i,1));
+        
+        if(i==0 || to_plot.get_data(i,0)<min.get_data(0))min.set(0,to_plot.get_data(i,0));
+        if(i==0 || to_plot.get_data(i,1)<min.get_data(1))min.set(1,to_plot.get_data(i,1));
+    }
+    
[…]

### 2014-07-21 aps4p5/aps_extractor.h

@@ -41,6 +41,8 @@ private:
     void learn_chimin();
     void learn_nparams();
     void validate();
+    
+    void plot_thinned_data(array_2d<double>&,double,char*);
 };
 
 #endif

### 2014-07-21 aps4p5/aps_extraction_runner.cpp
aps_extract write_good_points can now also thin

@@ -19,12 +19,14 @@ char outname[letters];
 for(i=0;i<6;i++){
     for(j=i+1;j<6;j++){
         if(i!=3 && j!=3){
-            sprintf(outname,"aps_processed/WMAP/bayesianTest/nnLearn_%d_%d_10k_bayes.sav",i,j);
-            test.draw_bayesian_bounds(outname,i,j,0.95);
+            sprintf(outname,"aps_processed/WMAP/bayesian/nnLearn_%d_%d_10k_bayes.sav",i,j);
+            test.draw_bayesian_bounds(outname,i,j,0.95,0.01);
+            
+            sprintf(outname,"aps_processed/WMAP/frequentist/nnLearn_%d_%d_10k_frequentist.sav",i,j);
+            test.write_good_points(outname,i,j,0.01);
         }
     }
 }
 
-test.show_minpt();
 
 }

### 2014-07-21 aps4p5/aps_extractor.cpp

@@ -147,6 +147,46 @@ void aps_extractor::write_good_points(char *outname){
     printf("wrote good pts with chi_min %e\n",chi_min);
 }
 
+void aps_extractor::write_good_points(char *outname, int ix, int iy, double tol){
+    learn_chimin();
+    
+    int i,ct=0,j;
+    FILE *input=fopen(filename,"r");
+    
+    char word[letters];
+    for(i=0;i<nparams+extra_words;i++)fscanf(input,"%s",word);
+    
+    double nn;
+    array_1d<double> vv;
+    array_2d<double> to_plot;
+    
+    to_plot.set_cols(2);
+    while(fscanf(input,"%le",&nn)>0 && (ct<cutoff || cutoff<0)){
+        ct++;
+        vv.set(0,nn);
+        for(i=1;i<nparams;i++){
+            fscanf(input,"%le",&nn);
+            vv.set(i,nn);
+        }
+        fscanf(input,"%le",&nn);
+        
+        if(nn<=chi_min+delta_chi+tol){
+            j=to_plot.get_rows();
+            to_plot.set(j,0,vv.get_data(ix));
+            to_plot.set(j,1,vv.get_data(iy));
+        }
+        
+        for(i=0;i<extra_words-2;i++)fscanf(input,"%le",&nn);
+    }
+    
+    
+    fclose(input);
+    
+    plot_thinned_data(to_plot,tol,outname);
+    
+    printf("wrote good pts with chi_min %e\n",chi_min);
+}
+
 void aps_extractor::plot_chimin(char *outname){
     learn_nparams();
     
@@ -463,7 +503,7 @@ void aps_extractor::sample_posterior(char *outname,array_2d<double> &samples, in
     }
 }
 
-void aps_extractor::draw_bayesian_bounds(char *filename, int ix, int iy, double limit){
+void aps_extractor::draw_bayesian_bounds(char *filename, int ix, int iy, double limit, double tol){
     if(l_probability.get_dim()==0){
         make_boxes();
     }
@@ -499,7 +539,7 @@ void aps_extractor::draw_bayesian_bounds(char *filename, int ix, int iy, double 
         
     }
     
-    plot_thinned_data(to_plot,0.01,filename);    
+    plot_thinned_data(to_plot,tol,filename);    
 }
 
 void aps_extractor::plot_thinned_data(array_2d<double> &to_plot, double tol, char *filename){    

### 2014-07-21 aps4p5/aps_extractor.h

@@ -13,6 +13,7 @@ public:
     void set_filename(char*);
     void set_delta_chi(double);
     void write_good_points(char*);
+    void write_good_points(char*,int,int,double);
     void plot_chimin(char*);
     
     void show_minpt();
@@ -21,7 +22,7 @@ public:
     void sample_posterior(array_2d<double>&,int);
     void sample_posterior(char*,array_2d<double>&,int,int);
     
-    void draw_bayesian_bounds(char*,int,int,double);
+    void draw_bayesian_bounds(char*,int,int,double,double);
     
     void make_boxes();
     

### 2014-07-21 aps4p5/aps_extraction_runner.cpp
only learn chimin if you have to

@@ -3,13 +3,13 @@
 main(){
 
 aps_extractor test;
-test.set_filename("aps_output/WMAP/apsWMAP_unitSphere_NNLearn_output.sav");
+test.set_filename("aps_output/WMAP/apsWMAP_unitSphere_GaussLearn_output.sav");
 test.set_delta_chi(12.6);
 //test.set_delta_chi(100.0);
 
 test.set_cutoff(10000);
 
-test.write_good_points("aps_processed/WMAP/frequentist/nnLearn_good_points_10k.sav");
+test.write_good_points("aps_processed/WMAP/frequentist/gaussLearn_good_points_10k.sav");
 //test.plot_chimin("aps_chi_min_test.sav");
 //test.sample_posterior("aps_processed/apsSphere_samples.sav",10000);
 
@@ -19,10 +19,10 @@ char outname[letters];
 for(i=0;i<6;i++){
     for(j=i+1;j<6;j++){
         if(i!=3 && j!=3){
-            sprintf(outname,"aps_processed/WMAP/bayesian/nnLearn_%d_%d_10k_bayes.sav",i,j);
+            sprintf(outname,"aps_processed/WMAP/bayesian/gaussLearn_%d_%d_10k_bayes.sav",i,j);
             test.draw_bayesian_bounds(outname,i,j,0.95,0.01);
             
-            sprintf(outname,"aps_processed/WMAP/frequentist/nnLearn_%d_%d_10k_frequentist.sav",i,j);
+            sprintf(outname,"aps_processed/WMAP/frequentist/gaussLearn_%d_%d_10k_frequentist.sav",i,j);
             test.write_good_points(outname,i,j,0.01);
         }
     }

### 2014-07-21 aps4p5/aps_extractor.cpp

@@ -108,10 +108,13 @@ void aps_extractor::learn_chimin(){
 
 void aps_extractor::set_cutoff(int ii){
     cutoff=ii;
+    chi_min=-1.0;
 }
 
 void aps_extractor::write_good_points(char *outname){
-    learn_chimin();
+    if(chi_min<0.0){
+        learn_chimin();
+    }
     
     int i,ct=0;
     FILE *output=fopen(outname,"w");
@@ -148,7 +151,9 @@ void aps_extractor::write_good_points(char *outname){
 }
 
 void aps_extractor::write_good_points(char *outname, int ix, int iy, double tol){
-    learn_chimin();
+    if(chi_min<0.0){
+        learn_chimin();
+    }
     
     int i,ct=0,j;
     FILE *input=fopen(filename,"r");

### 2014-07-21 aps4p5/aps_extraction_runner.cpp
aps_extractor now lets you asset chi_lim

@@ -3,13 +3,14 @@
 main(){
 
 aps_extractor test;
-test.set_filename("aps_output/WMAP/apsWMAP_unitSphere_GaussLearn_output.sav");
+test.set_target(1283.3);
+test.set_filename("aps_output/WMAP/apsWMAP_unitSphere_GaussAssert_output.sav");
 test.set_delta_chi(12.6);
 //test.set_delta_chi(100.0);
 
-test.set_cutoff(10000);
+test.set_cutoff(27000);
 
-test.write_good_points("aps_processed/WMAP/frequentist/gaussLearn_good_points_10k.sav");
+test.write_good_points("aps_processed/WMAP/frequentist/gaussAssert_good_points_30k.sav");
 //test.plot_chimin("aps_chi_min_test.sav");
 //test.sample_posterior("aps_processed/apsSphere_samples.sav",10000);
 
@@ -19,10 +20,10 @@ char outname[letters];
 for(i=0;i<6;i++){
     for(j=i+1;j<6;j++){
         if(i!=3 && j!=3){
-            sprintf(outname,"aps_processed/WMAP/bayesian/gaussLearn_%d_%d_10k_bayes.sav",i,j);
+            sprintf(outname,"aps_processed/WMAP/bayesian/gaussAssert_%d_%d_30k_bayes.sav",i,j);
             test.draw_bayesian_bounds(outname,i,j,0.95,0.01);
             
-            sprintf(outname,"aps_processed/WMAP/frequentist/gaussLearn_%d_%d_10k_frequentist.sav",i,j);
+            sprintf(outname,"aps_processed/WMAP/frequentist/gaussLearn_%d_%d_30k_frequentist.sav",i,j);
             test.write_good_points(outname,i,j,0.01);
         }
     }

### 2014-07-21 aps4p5/aps_extractor.cpp

@@ -2,15 +2,17 @@
 
 aps_extractor::aps_extractor(){
     chi_min=-1.0;
+    chi_target=-1.0;
     delta_chi = chisq_exception;
     filename[0]=0;
     nparams=-1;
+    asserted=0;
     
     cutoff=-1;
     
     extra_words=5;
     
-    tol=1.0e-6;
+    global_tol=1.0e-6;
     
     box_max.set_name("box_max");
     box_min.set_name("box_min");
@@ -35,6 +37,11 @@ void aps_extractor::validate(){
         printf("WARNING, no filename\n");
         throw -1;
     }
+}
+
+void aps_extractor::set_target(double tt){
+    chi_target=tt;
+    asserted=1;
 }
 
 void aps_extractor::set_delta_chi(double nn){
@@ -104,6 +111,11 @@ void aps_extractor::learn_chimin(){
     }
     
     fclose(input);
+    
+    if(asserted==0){
+        chi_target=chi_min+delta_chi;
+    }
+    
 }
 
 void aps_extractor::set_cutoff(int ii){
@@ -134,7 +146,7 @@ void aps_extractor::write_good_points(char *outname){
         }
         fscanf(input,"%le",&nn);
         
-        if(nn<=chi_min+delta_chi+tol){
+        if(nn<=chi_target+global_tol){
             for(i=0;i<nparams;i++){
                 fprintf(output,"%le ",vv.get_data(i));
             }
@@ -147,7 +159,7 @@ void aps_extractor::write_good_points(char *outname){
     fclose(output);
     fclose(input);
     
-    printf("wrote good pts with chi_min %e\n",chi_min);
+    printf("wrote good pts with chi_min %e target %e\n",chi_min,chi_target);
 }
 
 void aps_extractor::write_good_points(char *outname, int ix, int iy, double tol){
@@ -175,7 +187,7 @@ void aps_extractor::write_good_points(char *outname, int ix, int iy, double tol)
         }
         fscanf(input,"%le",&nn);
         
-        if(nn<=chi_min+delta_chi+tol){
+        if(nn<=chi_target+global_tol){
             j=to_plot.get_rows();
             to_plot.set(j,0,vv.get_data(ix));
             to_plot.set(j,1,vv.get_data(iy));
@@ -189,7 +201,7 @@ void aps_extractor::write_good_points(char *outname, int ix, int iy, double tol)
     
     plot_thinned_data(to_plot,tol,outname);
     
[…]

### 2014-07-21 aps4p5/aps_extractor.h

@@ -27,13 +27,15 @@ public:
     void make_boxes();
     
     void set_cutoff(int);
+    void set_target(double);
 
 private:
 
     char filename[letters];
-    double chi_min,delta_chi,tol;
+    double chi_target;
+    double chi_min,delta_chi,global_tol;
     int nparams,extra_words;
-    int cutoff;
+    int cutoff,asserted;
     
     array_1d<int> l_prob_dexes;
     array_1d<double> min_pt,l_probability,chisq;

### 2014-07-21 aps4p5/aps_extraction_ellipse.cpp
added aps_extraction_ellipse.cpp to repo

@@ -1,0 +1,30 @@
+#include "aps_extractor.h"
+
+main(){
+
+aps_extractor test;
+test.set_filename("aps_output/ellipse/fullAnalysis/ellipse_d5_c2_output.sav");
+test.set_delta_chi(11.0);
+//test.set_delta_chi(100.0);
+
+//test.set_cutoff(15000);
+
+//test.write_good_points("aps_processed/ellipse/frequentist/ellipse_d5_c2_good.sav");
+//test.plot_chimin("aps_chi_min_test.sav");
+//test.sample_posterior("apsEllipse_samples.sav",10000);
+
+char bayesname[letters];
+int i,j;
+for(i=0;i<5;i++){
+    for(j=i+1;j<5;j++){
+        sprintf(bayesname,"aps_processed/ellipse/bayesianThinned/ellipse_d5_c2_%d_%d_bayes.sav",i,j);
+        test.draw_bayesian_bounds(bayesname,i,j,0.95,1.0e-3);
+        
+        sprintf(bayesname,"aps_processed/ellipse/frequentistThinned/ellipse_d5_c2_%d_%d_frequentist.sav",i,j);
+        test.write_good_points(bayesname,i,j,1.0e-3);
+    }
+}
+
+test.show_minpt();
+
+}

### 2014-07-21 aps4p5/aps_extractor.cpp
fixed bug in aps_extractor.cpp

@@ -251,9 +251,8 @@ void aps_extractor::make_boxes(){
         exit(1);
     }
     
-    if(!(delta_chi<chisq_exception)){
-        printf("WARNING delta_chi %e in sample_posterior\n");
-        exit(1);
+    if(chi_min<0.0){
+        learn_chimin();
     }
     
     l_prob_dexes.reset();

### 2014-07-21 aps4p5/drafts/draft_140719.tex
more writing

@@
 $\bar{\chi}^2$
-and variance dictated by the covariance function 
+and variance dictated by the ``covariogram''
 $C_{ij}(\vec{\theta}^{(i)},\vec{\theta}^{(j)})$.
@@
 
+As mentioned above, because our toy $\chi^2$ function is constructed without any
+parameter degeneracies, we are able to directly integrate the posterior and
+determine the true Bayesian credible limits in our parameter space.  Figures
+\ref{fig:toyBayes2}, \ref{fig:toyBayes3}, and \ref{fig:toyBayes4} plot the true
+95\% credible limits in all 10 2-dimensional sub-spaces of our 5-dimensional
+parameter space (red contours).  They also show the APS-determined credible
+limits as described in Section \ref{sec:bayes} (black points).  The
+APS-determined credible limits correspond well with the true credible limits,
+giving us confidence in APS's ability to map out Bayesian credible limits as well
+as Frequentist confidence limits.
+
+The tests above demonstrate APS's ability to correctly determin credible and
+confidence limits.  They do not, however, address the question of APS's chief
+advantage over Markov Chain Monte Carlo methods: its ability to successfully
+locate all of the low-$\chi^2$ regions in a given parameter space.  To
+demonstrate this, we ran APS 100 times on each of the likelihood functions
+plotted in Figures \ref{fig:toyFreq2}-\ref{fig:toyFreq4}.  Each time, the
+pseudo-random number generator driving APS was given a different seed.  In all
+cases, APS successfully located all of the local minima in $\chi^2$, usually in
+fewer than 10,000 evaluations of $\chi^2$ (one case of the 2-mode $\chi^2$
+function required 10,352 evaluations to find both low-$\chi^2$ regions; one
+required 12,549 evaluations).  For comparison, we also ran 100 Markov Chain
+Monte Carlo searches on these $\chi^2$ function, each run consisting of five
+independent chains.  These searches were not nearly as successful as APS at
+locating all of the low-$\chi^2$ regions.  In the case of the 2-mode function,
+50 of the 100 MCMC searches failed to identify both modes.  In the case of the
+3-mode function, 56 of the 100 MCMC searches failed to identify all of the
+modes.  In the case of the 4-mode function, 87 of the 100 MCM searches failed to
+identify all of the modes.  This is not to say that APS is perfect at
+identifying all of the modes of a likelihood function.  If we reduce the
+characteristic widths $w_{j,i}$ in equation (\ref{eqn:toychi}), we enter a regime
+in which neither APS nor MCMC is capable of reliably,
+identifying all of the modes.  However, this test does indicate that APS is more
+robust than MCMC against multi-modal likelihood functions, as was first
+demonstrated by Bryan {\it et al}. (2007).
+
 \begin{figure*}
@@
 for an algorithm like APS or MCMC to accelerate the search.
-We run APS with the following input parameters:
-\begin{eqnarray}
-\Delta\chi^2&=&12.59 \nonumber\\
-&&\text{(according to the likelihood ratio test,}\nonumber\\
-&&\text{the
-95\% confidence limit for a 6-dimensional}\nonumber\\
-&&\text{ parameter space
-ought to be }\chi^2_\text{min}+12.59\text{)}\nonumber\\
-R_g&=&0.1\nonumber\\
-N_w&=&100\nonumber\\
-N_c&=&250\nonumber\\
-N_g&=&15\nonumber
-\end{eqnarray}
-We will consider both equations (\ref{eqn:covraw}) and (\ref{eqn:cov_nn})
-as covariograms, though we find that the choice has little effect on our
+We will consider both equations (\ref{eqn:GaussianCovar}) and (\ref{eqn:NNCovar})
+as covariogram for our Gaussian Process, 
+though we find that the choice has little effect on our
 results.
@@
 function of the number of steps sampled (note: the vertical axis actually shows the average
-fractional error over the preceding 2,000 steps so that the early, imprecise models do not
+fractional error over the preceding 500 steps so that the early, imprecise models do not
 contaminate the later, precise models).  As APS samples more points, its Gaussian Process
-model becomes more accurate and the fractional error converges to a few percent.  Thus our
-use of Gaussian Processes to model $\vec{\theta}\rightarrow\chi^2$ is also justified.
+model becomes more accurate.  This effect is more pronounced for covariogram (\ref{eqn:GaussianCovar})
+than for covariogram (\ref{eqn:NNCovar}).  However, we will see below that this difference
+does not effect the parameter constraints determined by APS.
 
@@
 $\vec{\theta}\rightarrow\chi^2$ can be approximated as a Gaussian Process.
-As the number of sample points increases, the average error yielded by either
-covariogram quickly converges to a few percent in $\chi^2$.
+Though the squared exponential covariogram gives a better fit to the $\chi^2$ function,
+we see in Figures \ref{fig:contours_gauss} and \ref{fig:contours_nn}
+that this has little effect on the parameter constraints derived by APS.
 }
@@
 \frac{\text{Cov}[\theta_n^{(i)},\theta_n^{(i+L)}]}
-{\text{Var}[\theta_n]}\le 0.01
+{\text{Var}[\theta_n]}\le 0.1
 \end{equation}
@@
 limits in two-dimensional slices of our parameter space using
-a total of 460,000 MCMC steps.  These are the solid
-black contours in Figures \ref{fig:contours_gauss} through 
+a total of 460,000 MCMC steps.  These are assumed to be the
+true credible limits on the parameter space and are plotted
+as the thick red
+contours in Figures \ref{fig:contours_gauss} through 
 \ref{fig:contours_br} 
-below.  They are referred to below as the ``final credible limits''
-returned by MCMC.  To gauge
-convergence as a function of time, we also plot the 95\% credible limit contours
-taken from our set of four MCMC chains after they have sampled
-a total of 100,000 points.  
-These are the dotted green
-contours in Figures \ref{fig:contours_gauss} through \ref{fig:contours_br}.
-They are referred to as the ``pre-convergence'' credible limits.
-Note that, even though MCMC has discovered its final value of $\chi^2_\text{min}$
-by this point (see Figure \ref{fig:chimin}), this rapid convergence in
-$\chi^2_\text{min}$ has not translated into convergence for the credible limits
-on parameter space.
-
-In the case of APS, the confidence limit is interpreted 
-as that set of points found
-which satisfy $\chi^2\le\chi^2_\text{lim}$.  Figures \ref{fig:contours_gauss}
-and \ref{fig:contours_nn} plot these points after APS has sampled 100,000 
-points in the case where $\chi^2_\text{lim}\equiv\chi^2_\text{min}+12.59$
-using equations (\ref{eqn:covraw}) and (\ref{eqn:cov_nn}) as covariograms,
-respectively.
-We find that the confidence limits returned by APS overlap with
-the final credible limits of MCMC, but are generally too permissive,
-they include regions of parameter space which MCMC ultimately excludes.
-It is tempting to say that
-this is due to the fact that the $\chi^2_\text{min}$ discovered by
-APS is $\sim3$ greater than the $\chi^2_\text{min}$ discovered by
-MCMC (therefore, $\chi^2_\text{lim}$ is $\sim 3$ too large).
-However, Figure \ref{fig:contours_br} uses the same data as
-Figure \ref{fig:contours_gauss}, but plots only those APS points
-such that $\chi^2\le 1283.3$, i.e. $\chi^2\le\chi^2_\text{min, MCMC}+12.59$.
-Still we find that the APS confidence limit includes regions of parameter
-space excluded by the final MCMC contour.  Thus, the difference in
-parameter space constraints between APS and MCMC amounts to difference
-between the frequentist confidence limit (see Appendix \ref{sec:frequentism})
-and the Bayesian credible limit.  The fact that the two contraints overlap
-significantly but not entirely indicates the value in performing multiple
-statistical tests before drawing conclusions from one's data set.
-It is also worth noting that in all cases, APS has thoroughly explored
-the correct confidence limit in fewer samples than has MCMC.  Compare the
-coverage of the red scatter plot to the green pre-convergence MCMC contours in
-Figures \ref{fig:contours_gauss} through \ref{fig:contours_br}.
+below.  Against these control contours, we plot the Frequentist
+confidence limits (all points with $\chi^2\le\chi^2_\text{lim}$)
+discovered by APS after only 50,000 evaluations of $\chi^2$.
+We consider both an adaptive $\chi^2_\text{lim}=\chi^2_\text{min}+12.6$
+and an absoulte $\chi^2_\text{lim}=1283.3$ (this is the 95\% confidence limit
+for a $\chi^2$ probability distribution with 1199 degrees of freedom).
+These are the black points in Figures \ref{fig:contours_gauss}, \ref{fig:contours_nn},
+and \ref{fig:contours_br} respectively.  These confidence limits overlap very nicely
+with the credible limits discovered by MCMC after many more $\chi^2$ evaluations,
+comporting the the idea that, in the large data limit, Bayesiand credible limits
+and Frequentist confidence limits ought to align.
+
+For comparison of convergence properties, we plot the Bayesian credible limits discovered
+by MCMC after only 50,000 $\chi^2$ evaluations.  These are the black regions in
+Figure \ref{fig:mcmc_comparison}.  As can be seen, MCMC has not yet totally converged
+in all of the sub-spaces of cosmological parameter space.  This should be compared
+with the reasonably complete coverage of the control credible limits seen in
+Figures \ref{fig:contours_gauss}-\ref{fig:contours_br}.
+
+In Figure \ref{fig:contours_bayes}, we plot the 95\% Bayesian credible limit discovered
+by APS after 50,000 evalutations of $\chi^2$ and determined by the method outlined in
+Section \ref{sec:bayes}.  While we do not see the clean equivalence between the APS-derived credible limits and
+the true credible limits that we saw in Figures \ref{fig:toyBayes2}-\ref{fig:toyBayes4},
+we do find that the APS-derived credible limits roughly sketch the bounds of the 
+true credible limit.  This, combined with its robustness against multi-modal likelihood
+functions, suggests that APS may, at the very least be useful in cases where very little
+prior information is available and exploring the paramter space is at least as important as
+integrating it to find a rigorous bound.
 
@@
 \subfigure[]{
-\includegraphics[scale=0.15]{scatter_100k_0_1.eps}
-}
-\subfigure[]{
-\includegraphics[scale=0.15]{scatter_100k_0_2.eps}
-}
-\subfigure[]{
-\includegraphics[scale=0.15]{scatter_100k_0_3.eps}
-}
-\subfigure[]{
-\includegraphics[scale=0.15]{scatter_100k_0_4.eps}
-}
-\subfigure[]{
-\includegraphics[scale=0.15]{scatter_100k_1_2.eps}
-}
-\subfigure[]{
-\includegraphics[scale=0.15]{scatter_100k_1_3.eps}
-}
-\subfigure[]{
-\includegraphics[scale=0.15]{scatter_100k_1_4.eps}
-}
-\subfigure[]{
-\includegraphics[scale=0.15]{scatter_100k_2_3.eps}
-}
-\subfigure[]{
-\includegraphics[scale=0.15]{scatter_100k_2_4.eps}
-}
-\subfigure[]{
-\includegraphics[scale=0.15]{scatter_100k_3_4.eps}
+\includegraphics[scale=0.15]{gaussLearn_0_1_50k_frequentist.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{gaussLearn_0_2_50k_frequentist.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{gaussLearn_0_4_50k_frequentist.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{gaussLearn_0_5_50k_frequentist.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{gaussLearn_1_2_50k_frequentist.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{gaussLearn_1_4_50k_frequentist.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{gaussLearn_1_5_50k_frequentist.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{gaussLearn_2_4_50k_frequentist.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{gaussLearn_2_5_50k_frequentist.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{gaussLearn_4_5_50k_frequentist.eps}
 }
 \caption{
-95\% confidence limits in two-dimensional slices of parameter space as
-determined by MCMC and APS after 100,000 points have been sampled.  Solid (black)
-contours are the baseline result found by 4 MCMC chains
-after sampling a total of 460,000 points.  
-Dotted (green) contours are the result found by a 4 MCMC
-chains after each has sampled 25,000 points (for a total of 100,000
-sampled points).  Red points show the $\chi^2\le\chi^2_\text{min}+12.59$
-points found by APS after sampling a total of 100,000 points.
-These plots use equation (\ref{eqn:covraw}) for the covariogram.
+The 10 2-dimensional sub-spaces of our 6-dimensional parameter space
+($\tau$ is ignored because not real constraint can be gleaned from this data).
+The thick red contours are the 95\% Bayesian credible limits determined by MCMC after
+460,000 calls to $\chi^2$.  The black points are the 95\% Frequentist confidence limits
+determined by APS after 50,000 calls to $\chi^2$.  $\chi^2_\text{lim}=\chi^2_\text{min}+12.6$.
+Equation (\ref{eqn:GaussianCovar}) is used for the Gaussian Process covariogram.
 }
@@
 \subfigure[]{
-\includegraphics[scale=0.15]{scatter_nn_100k_0_1.eps}
-}
-\subfigure[]{
-\includegraphics[scale=0.15]{scatter_nn_100k_0_2.eps}
-}
-\subfigure[]{
-\includegraphics[scale=0.15]{scatter_nn_100k_0_3.eps}
-}
-\subfigure[]{
-\includegraphics[scale=0.15]{scatter_nn_100k_0_4.eps}
-}
-\subfigure[]{
-\includegraphics[scale=0.15]{scatter_nn_100k_1_2.eps}
-}
-\subfigure[]{
-\includegraphics[scale=0.15]{scatter_nn_100k_1_3.eps}
-}
-\subfigure[]{
-\includegraphics[scale=0.15]{scatter_nn_100k_1_4.eps}
-}
-\subfigure[]{
-\includegraphics[scale=0.15]{scatter_nn_100k_2_3.eps}
-}
-\subfigure[]{
-\includegraphics[scale=0.15]{scatter_nn_100k_2_4.eps}
-}
-\subfigure[]{
-\includegraphics[scale=0.15]{scatter_nn_100k_3_4.eps}
+\includegraphics[scale=0.15]{nnLearn_0_1_50k_frequentist.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{nnLearn_0_2_50k_frequentist.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{nnLearn_0_4_50k_frequentist.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{nnLearn_0_5_50k_frequentist.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{nnLearn_1_2_50k_frequentist.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{nnLearn_1_4_50k_frequentist.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{nnLearn_1_5_50k_frequentist.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{nnLearn_2_4_50k_frequentist.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{nnLearn_2_5_50k_frequentist.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{nnLearn_4_5_50k_frequentist.eps}
 }
 \caption{
-95\% confidence limits in two-dimensional slices of parameter space as
-determined by MCMC and APS after 100,000 points have been sampled.  Solid (black)
-contours are the baseline result found by 4 MCMC chains
-after sampling a total of 460,000 points.  
-Dotted (green) contours are the result found by a 4 MCMC
-chains after each has sampled 25,000 points (for a total of 100,000
-sampled points).  Red points show the $\chi^2\le\chi^2_\text{min}+12.59$
-points found by APS after sampling a total of 100,000 points.
-These plots use equation (\ref{eqn:cov_nn}) for the covariogram.
+The 10 2-dimensional sub-spaces of our 6-dimensional parameter space
+($\tau$ is ignored because not real constraint can be gleaned from this data).
+The thick red contours are the 95\% Bayesian credible limits determined by MCMC after
+460,000 calls to $\chi^2$.  The black points are the 95\% Frequentist confidence limits
+determined by APS after 50,000 calls to $\chi^2$.  $\chi^2_\text{lim}=\chi^2_\text{min}+12.6$.
+Equation (\ref{eqn:NNCovar}) is used for the Gaussian Process covariogram.
 }
@@
 \end{figure*}
+
 \begin{figure*}
 \subfigure[]{
-\includegraphics[scale=0.15]{scatter_br_100k_0_1.eps}
-}
-\subfigure[]{
-\includegraphics[scale=0.15]{scatter_br_100k_0_2.eps}
-}
-\subfigure[]{
-\includegraphics[scale=0.15]{scatter_br_100k_0_3.eps}
-}
-\subfigure[]{
-\includegraphics[scale=0.15]{scatter_br_100k_0_4.eps}
-}
-\subfigure[]{
-\includegraphics[scale=0.15]{scatter_br_100k_1_2.eps}
-}
-\subfigure[]{
-\includegraphics[scale=0.15]{scatter_br_100k_1_3.eps}
-}
-\subfigure[]{
-\includegraphics[scale=0.15]{scatter_br_100k_1_4.eps}
-}
-\subfigure[]{
-\includegraphics[scale=0.15]{scatter_br_100k_2_3.eps}
-}
-\subfigure[]{
-\includegraphics[scale=0.15]{scatter_br_100k_2_4.eps}
-}
-\subfigure[]{
-\includegraphics[scale=0.15]{scatter_br_100k_3_4.eps}
+\includegraphics[scale=0.15]{gaussAssert_0_1_50k_frequentist.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{gaussAssert_0_2_50k_frequentist.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{gaussAssert_0_4_50k_frequentist.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{gaussAssert_0_5_50k_frequentist.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{gaussAssert_1_2_50k_frequentist.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{gaussAssert_1_4_50k_frequentist.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{gaussAssert_1_5_50k_frequentist.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{gaussAssert_2_4_50k_frequentist.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{gaussAssert_2_5_50k_frequentist.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{gaussAssert_4_5_50k_frequentist.eps}
 }
 \caption{
-The black and green contours are as in Figure \ref{fig:contours_gauss}.
-The red points represent the $\chi^2\le 1283.3$ points sampled by the
-APS run used to generate Figure \ref{fig:contours_gauss}.
+The 10 2-dimensional sub-spaces of our 6-dimensional parameter space
+($\tau$ is ignored because not real constraint can be gleaned from this data).
+The thick red contours are the 95\% Bayesian credible limits determined by MCMC after
+460,000 calls to $\chi^2$.  The black points are the 95\% Frequentist confidence limits
+determined by APS after 50,000 calls to $\chi^2$.  $\chi^2_\text{lim}=1283.3$.
+Equation (\ref{eqn:GaussianCovar}) is used for the Gaussian Process covariogram.
 }
@@
 
+\begin{figure*}
+\subfigure[]{
+\includegraphics[scale=0.15]{mcmc_comparison_0_1_50k.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{mcmc_comparison_0_2_50k.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{mcmc_comparison_0_4_50k.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{mcmc_comparison_0_5_50k.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{mcmc_comparison_1_2_50k.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{mcmc_comparison_1_4_50k.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{mcmc_comparison_1_5_50k.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{mcmc_comparison_2_4_50k.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{mcmc_comparison_2_5_50k.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{mcmc_comparison_4_5_50k.eps}
+}
+\caption{
+The 10 2-dimensional sub-spaces of our 6-dimensional parameter space
+($\tau$ is ignored because not real constraint can be gleaned from this data).
+The thick red contours are the 95\% Bayesian credible limits determined by MCMC after
+460,000 calls to $\chi^2$.  The black regions are the 95\% Bayesian credible limits
+determined by MCMC after 50,000 calls to $\chi^2$.
+}
+\label{fig:mcmc_comparison}
+\end{figure*}
+
+\begin{figure*}
+\subfigure[]{
+\includegraphics[scale=0.15]{gaussLearn_0_1_50k_bayes.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{gaussLearn_0_2_50k_bayes.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{gaussLearn_0_4_50k_bayes.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{gaussLearn_0_5_50k_bayes.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{gaussLearn_1_2_50k_bayes.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{gaussLearn_1_4_50k_bayes.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{gaussLearn_1_5_50k_bayes.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{gaussLearn_2_4_50k_bayes.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{gaussLearn_2_5_50k_bayes.eps}
+}
+\subfigure[]{
+\includegraphics[scale=0.15]{gaussLearn_4_5_50k_bayes.eps}
+}
+\caption{
+The 10 2-dimensional sub-spaces of our 6-dimensional parameter space
+($\tau$ is ignored because not real constraint can be gleaned from this data).
+The thick red contours are the 95\% Bayesian credible limits determined by MCMC after
+460,000 calls to $\chi^2$.  The black points are the 95\% Bayesian credible limits determined
+by APS according to the method outlined in Section \ref{sec:bayes}.  While the APS credible limit
+covers the entire MCMC limit, it also covers regions of parameter space excluded by MCMC.
+}
+\label{fig:contours_bayes}
+\end{figure*}
+
 \section{Conclusions}
@@
 
-In Section \ref{sec:toy} we showed that APS is more robust
+In Section \ref{sec:toy} we indicated that APS is more robust
 than MCMC against multi-modal likelihood functions.
 In Section \ref{sec:wmap} we showed that, on more well-behaved
-likelihood functions, APS discovers the same parameter constraints
+likelihood functions, APS discovers comparable parameter constraints
 as MCMC and that APS does not converge any more slowly than does MCMC.
@@
 \section*{Acknowledgments}
-The authors would like to thank the referee for useful comments and suggestions
+The authors would like to thank the referee for comments and suggestions
 which we hope have made this paper more useful to astrophysical audiences.

### 2014-07-21 aps4p5/aps_extraction_ellipse.cpp

@@ -3,7 +3,7 @@
 main(){
 
 aps_extractor test;
-test.set_filename("aps_output/ellipse/fullAnalysis/ellipse_d5_c2_output.sav");
+test.set_filename("aps_output/ellipse/fullAnalysis/ellipse_d5_c4_output.sav");
 test.set_delta_chi(11.0);
 //test.set_delta_chi(100.0);
 
@@ -17,10 +17,10 @@ char bayesname[letters];
 int i,j;
 for(i=0;i<5;i++){
     for(j=i+1;j<5;j++){
-        sprintf(bayesname,"aps_processed/ellipse/bayesianThinned/ellipse_d5_c2_%d_%d_bayes.sav",i,j);
+        sprintf(bayesname,"aps_processed/ellipse/bayesianThinned/ellipse_d5_c4_%d_%d_bayes.sav",i,j);
         test.draw_bayesian_bounds(bayesname,i,j,0.95,1.0e-3);
         
-        sprintf(bayesname,"aps_processed/ellipse/frequentistThinned/ellipse_d5_c2_%d_%d_frequentist.sav",i,j);
+        sprintf(bayesname,"aps_processed/ellipse/frequentistThinned/ellipse_d5_c4_%d_%d_frequentist.sav",i,j);
         test.write_good_points(bayesname,i,j,1.0e-3);
     }
 }

### 2014-07-21 aps4p5/aps_extraction_runner.cpp

@@ -2,28 +2,32 @@
 
 main(){
 
+char sizeflag[100];
+char outname[letters];
+
 aps_extractor test;
-test.set_target(1283.3);
-test.set_filename("aps_output/WMAP/apsWMAP_unitSphere_GaussAssert_output.sav");
+//test.set_target(1283.3);
+test.set_filename("aps_output/WMAP/apsWMAP_unitSphere_GaussLearn_output.sav");
 test.set_delta_chi(12.6);
 //test.set_delta_chi(100.0);
 
-test.set_cutoff(27000);
+//test.set_cutoff(47000);
+sprintf(sizeflag,"100k");
 
-test.write_good_points("aps_processed/WMAP/frequentist/gaussAssert_good_points_30k.sav");
+sprintf(outname,"aps_processed/WMAP/frequentist/gaussLearn_good_points_%s.sav",sizeflag);
+test.write_good_points(outname);
 //test.plot_chimin("aps_chi_min_test.sav");
 //test.sample_posterior("aps_processed/apsSphere_samples.sav",10000);
 
 int i,j;
-char outname[letters];
 
 for(i=0;i<6;i++){
     for(j=i+1;j<6;j++){
         if(i!=3 && j!=3){
-            sprintf(outname,"aps_processed/WMAP/bayesian/gaussAssert_%d_%d_30k_bayes.sav",i,j);
+            sprintf(outname,"aps_processed/WMAP/bayesian/gaussLearn_%d_%d_%s_bayes.sav",i,j,sizeflag);
             test.draw_bayesian_bounds(outname,i,j,0.95,0.01);
             
-            sprintf(outname,"aps_processed/WMAP/frequentist/gaussLearn_%d_%d_30k_frequentist.sav",i,j);
+            sprintf(outname,"aps_processed/WMAP/frequentist/gaussLearn_%d_%d_%s_frequentist.sav",i,j,sizeflag);
             test.write_good_points(outname,i,j,0.01);
         }
     }

### 2014-07-21 aps4p5/mcmc_brute.cpp

@@ -28,7 +28,7 @@ ncenters=atoi(argv[3]);
 
 char outname[letters];
 
-sprintf(outname,"aps_output/ellipse/scriptOutput/ellipse_mcmc_d%d_c%d.sav",dim,ncenters);
+sprintf(outname,"aps_output/ellipse/scriptOutput/ellipse_mcmc_d%d_c%d_100k.sav",dim,ncenters);
 
 Ran chaos(seed);
 int i,j;
@@ -90,7 +90,7 @@ while(iteration<20 && found_all==0){
     
     found_all=1;
     iteration++;
-    mcmc_test.sample(200);
+    mcmc_test.sample(1000);
     
     ddmax=-1.0;
     for(i=0;i<ncenters;i++){

### 2014-07-21 aps4p5/mcmc_extraction_runner.cpp

@@ -12,12 +12,12 @@ nparams = 81;
 test.set_nchains(4);
 test.set_nparams(nparams);
 
-test.set_cutoff(7500);
+test.set_cutoff(12500);
 //control has a total 453,051 samples
 
 test.set_chainname("/Users/noldor/physics/recreate_getdist/ieuchains_1304/wmap7_learn");
 
-test.set_keep_frac(0.9);
+test.set_keep_frac(0.5);
 
 test.learn_thinby();
 
@@ -26,7 +26,7 @@ printf("thinby %d\nused %d\nkept %d\n",test.get_thinby(),test.get_total_used(),t
 printf("rows %d\n",test.get_total_rows());
 printf("best_covar %e\n",test.get_best_covar());
 
-test.print_samples("test_wmap7_samples30ktest.sav");
+test.print_samples("test_wmap7_samples50k.sav");
 
 kde kde_test;
 kde_test.set_data(test.get_samples());
@@ -57,14 +57,17 @@ for(i=0;i<6;i++){
     for(j=i+1;j<6;j++){
         if(i!=3 && j!=3){
  
-           sprintf(outname,"mcmc_processed/wmap_%d_%d_30ktest.sav",i,j);
+             //sprintf(outname,"mcmc_processed/wmap_%d_%d_50k.sav",i,j);
 
-            kde_test.plot_boundary(ix.get_data(i),dx.get_data(i),ix.get_data(j),dx.get_data(j),0.95,outname,3);
+            //kde_test.plot_boundary(ix.get_data(i),dx.get_data(i),ix.get_data(j),dx.get_data(j),0.95,outname,3);
+            
+            sprintf(outname,"mcmc_processed/wmap_scatter_%d_%d_50k.sav",i,j);
+            kde_test.plot_density(ix.get_data(i),dx.get_data(i),ix.get_data(j),dx.get_data(j),0.95,outname,3);
         }
     }
 }
 
-test.plot_delta("mcmc_good_pts_test30ktest.sav",0.5*12.61);
+test.plot_delta("mcmc_good_pts_test50k.sav",0.5*12.61);
 
 
 array_1d<double> RR,VV,WW,mean,var;

### 2014-07-21 aps4p5/test_mcmc.sh

@@ -1,11 +1,16 @@
 make mcmcBrute
 
 for ((i=0 ; i<100 ; i++)) do
+./mcmcBrute -1 5 3
+sleep 2
+done
+
+for ((i=0 ; i<100 ; i++)) do
 ./mcmcBrute -1 5 2
-sleep 5
+sleep 2
 done
 
 for ((i=0 ; i<100 ; i++)) do
 ./mcmcBrute -1 5 4
-sleep 5
+sleep 2
 done

### 2014-07-21 aps4p5/aps_interface.cpp
removed old software files

@@ -1,313 +1,0 @@
-#include <stdio.h>
-#include <stdlib.h>
-#include <time.h>
-#include <math.h>
-#include "likelihoodinterface.h"
-
-main(int iargc, char *argv[]){
-   
-   int i,j,k,l,ii,dim,nstart=2000,nend=10000,nc=200,llswit=0;
-   int resumeswitch=0,setcovar=0,setlike=0;
-   char paramname[letters],keyword[letters],word[letters];
-   char resumename[letters];
-   double nn,mm,*min,*max,exc;
-   
-   double **guess,**buff;
-   int nguess;
-   
-   nguess=0;
-   
-   likelihood *likeness;
-   covariance_function *cv;
-   likelihood_function *lk;
-   
-   FILE *input,*output;
-   
-   
-   for(j=0;argv[1][j]!=0;j++){
-     paramname[j]=argv[1][j];
-   }
-   paramname[j]=0;
-   
-   input=fopen(paramname,"r");
-   while(fscanf(input,"%s",keyword)==1){
-     if(compare_char(keyword,"#dim")==1){
-       
-       fscanf(input,"%d",&dim);
-       
-       min=new double[dim];
-       max=new double[dim];
-       
-     }
-     else{
-       if(compare_char(keyword,"#param")==1){
-         fscanf(input,"%d %s %le %le",&ii,word,&nn,&mm);
-	 
-	 if(nn>mm){
-	   max[ii]=nn;
-	   min[ii]=mm;
-	 }
-	 else{
-	   max[ii]=mm;
-	   min[ii]=nn;
-	 }
-	 
-	 //printf("set lims %d %e %e\n",ii,min[ii],max[ii]);
-	 
-       }
-       else if(compare_char(keyword,"#covariance")==1){
-           setcovar=1;
-           fscanf(input,"%s",word);
-	   if(compare_char(word,"gauss")==1){
-	       cv=new gaussian_covariance();
-	   }
-	   else if(compare_char(word,"nn")==1){
-	       cv=new nn_covariance();
-	   }
-	   else{
-	       cv=new gaussian_covariance();
-	   }
-         
-       }
-       else if(compare_char(keyword,"#likelihood")==1){
-           setlike=1;
-           fscanf(input,"%s",word);
-	   if(compare_char(word,"wmap")==1){
-	       #ifdef _WMAP7_
-	       lk=new wmap_likelihood();
-	       #else
-	       lk=new udder_likelihood();
-	       #endif
-	   }
[…]

### 2014-07-21 aps4p5/likelihoodinterface.cpp

@@ -1,2207 +1,0 @@
-#include <stdio.h>
-#include <stdlib.h>
-#include <math.h>
-
-#ifdef OPENMP
-#include <omp.h>
-#else
-#define omp_get_num_threads() 1
-#define omp_get_thread_num() 0
-#endif
-
-#include <time.h>
-#include "likelihoodinterface.h"
-//#include "eigen_wrapper.h"
-
-int ipower(int arg, int ee){
-  int ans,i;
-  
-  ans=1;
-  for(i=0;i<ee;i++)ans=ans*arg;
-  
-  return ans;
-}
-
-likelihood::~likelihood(){
-  int i;
-  double **dd,*d;
-
-  //Call subroutines with delswit=-1 so that they delete
-  //pointers that were allocated on their initial call  
-  
-  sample_pts(-1);
-  
-  delete [] nodes;
-  node_sample(-2);
-  grad_sample(-2);
- 
-  
-  for(i=0;i<ngw;i++){
-    if(gw[i].chisq<chiexcept)delete [] gw[i].center;
-  }
-  
-  for(i=0;i<nparams;i++)delete [] pnames[i];
-  delete pnames;
-  
-  delete dice;
-  delete [] minpt;
-  if(initialized==1)delete [] lingerflag;
-  
-  if(calledmufit==1){
-      delete [] mumufit;
-      delete [] diffmufit;
-      delete [] chimufit;
-      delete [] ctmufit;
-  }
-  
-
-}
-
-likelihood::likelihood(){
- 
- printf("WARNING called the likelihood default constructor\n");
-  printf("you shouldn't do that...");
-  printf("I'm going to freeze the program now\n");
-  scanf("%lf",&junk);
-}
-
-void likelihood::set_timingname(char *word){
-
-
-    int i;
-    for(i=0;word[i]!=0;i++)timingname[i]=word[i];
-    timingname[i]=0;
-}
-
-void likelihood::set_deltachi(double xx){
-    deltachi=xx;
-}
-
-void likelihood::set_seed(int ii){
-    seed=ii;
-}
-
-likelihood::likelihood(int nn,double *mns,double *mxs,
-covariance_function *cv, likelihood_function *lk){
-
- //This routine doesn't actually do much.
- //It just builds some of the necessary matrices
- //and assigns the values you want to some of the
- //class member variables.  The real action of
- //beginning to explore parameter space is in
- //likelihoodinterface.initiailize()
-
[…]

### 2014-07-21 aps4p5/likelihoodinterface.h

@@ -1,206 +1,0 @@
-#include "goto_tools.h"
-#include "gaussian_process.h"
-#define likeletters 500
-#define chiexcept 1.0e10
-
-//these are variables which allow the different likelihood_function classes
-//to identify themselves by type
-enum{LK_TYPE_UDDER,LK_TYPE_WMAP,LK_TYPE_OTHER};
-
-#ifdef _WMAP7_
-
-//the two external function definitions below are only needed to
-//interface with CAMB and the WMAP 7 likelihood function
-//as provided in aps_cmb_module.cpp
-
-extern "C" void \
-camb_wrap_(double*,double*,double*,double*,double*,double*);
-
-extern "C" void wmaplikeness_(double*,double*,double*,double*,double*);
-#endif
-
-class likelihood_function{
-
-protected:
-    //maximum and minimum parameter values are not required for all likelihood
-    //functions
-    double *maxs,*mins;
-    
-    //setmaxmin keeps track of whether or not *maxs and *mins have been
-    //allocated; dim is the dimensionality of parameter space
-    int setmaxmin,dim;
-    
-public:
-    
-    likelihood_function();
-    ~likelihood_function();
-    void set_max_min(int,double*,double*);
-    virtual double operator()(double*){};
-
-    virtual int get_type(){
-        return LK_TYPE_OTHER;
-    };
-    
-    //these are used for the case of the udder likelihood function;
-    //they return the iterations on which either the {-3,0,0,0,0,0}
-    //region of highlikelihood or the {3,0,0,0,0,0} region of high
-    //likelihood were found
-    virtual int get_fn3(){};
-    virtual int get_fp3(){};
-
-};
-
-#ifdef _WMAP7_
-class wmap_likelihood : public likelihood_function{
-
-public:
-    wmap_likelihood(){};
-    ~wmap_likelihood(){};
-    virtual double operator()(double*);
-    virtual int get_type();
-
-};
-#endif
-
-class udder_likelihood : public likelihood_function{
-
-private:
-    int foundp3,foundn3,called;
-    
-    //foundp3 and foundn3 keep track of when APS found the two different
-    //high likelihood regions in the udder likelihood function
-    
-    //their values can be accessed by the outside using the public
-    //functions get_fp3 and get_fn3 below
-
-public:
-    udder_likelihood();
-    ~udder_likelihood(){};
-    virtual double operator()(double*);
[…]

### 2014-07-21 aps4p5/aps_runner_ellipses.cpp
starting to document code

@@ -11,13 +11,16 @@ main(int iargc, char *argv[]){
 int seed=99;
 int dim,ncenters;
 
-
 seed=atoi(argv[1]);
 dim=atoi(argv[2]);
 ncenters=atoi(argv[3]);
 
 char timingname[letters],outname[letters];
+
+//what is the name of the file where APS will store its timing information
 sprintf(timingname,"aps_output/ellipse/fullAnalysis/ellipse_d%d_c%d_timing.sav",dim,ncenters);
+
+//what is the name of the file where APS will output the points it sampled
 sprintf(outname,"aps_output/ellipse/fullAnalysis/ellipse_d%d_c%d_output.sav",dim,ncenters);
 
 
@@ -27,29 +30,38 @@ if(seed<0){
 
 printf("seed %d\n",seed);
 
-Ran chaos(seed);
-
+//declare the covariogram for APS's Gaussian process
 matern_covariance cv;
 
+//declare the chisquared function APS will be searching
 ellipses_integrable chisq(dim,ncenters);
 
-
+//declare APS
+//the '20' below is the number of nearest neighbors to use when seeding the
+//Gaussian process
+//
+//the '11.0' is the \Delta\chi^2 corresponding to a 95% confidence limit
+//on a 5-dimensional parameter space
 aps aps_test(dim,20,11.0,seed);
 
+//pass chisq to the aps object
 aps_test.assign_chisquared(&chisq);
+
+//pass the covariogram to the aps object
 aps_test.assign_covariogram(&cv);
 
+//how often will APS stop and write its output
 aps_test.set_write_every(1000);
+
+//set the G parameter from equation (4) in the paper
 aps_test.set_grat(1.0);
 
+//set the maximum and minimum values in parameter space
 array_1d<double> max,min;
 max.set_name("driver_max");
 min.set_name("driver_min");
 max.set_dim(dim);
 min.set_dim(dim);
-
-aps_test.set_timingname(timingname);
-aps_test.set_outname(outname);
 
 
 int i,j;
@@ -59,15 +71,23 @@ for(i=0;i<dim;i++){
     max.set(i,100.0);
 }
 
-printf("initializing\n");
-aps_test.initialize(1000,min,max);
-printf("done initializing\n");
 
+aps_test.set_timingname(timingname);
+aps_test.set_outname(outname);
+
+//initialize aps with 1000 random samples
+aps_test.initialize(1000,min,max);
+
+//set N_c from step 2A in the paper
[…]

### 2014-07-21 aps4p5/aps.cpp
comment out code writing minima.sav

@@ -2946,7 +2946,7 @@ void aps::write_pts(){
     double mu,sig;
     FILE *output;
     
-    
+    /*
     output=fopen(minimaname,"w");
     fprintf(output,"known_minima %d\n",known_minima.get_dim());
     for(i=0;i<known_minima.get_dim();i++){
@@ -2957,7 +2957,7 @@ void aps::write_pts(){
         fprintf(output,"%d\n",forbidden_candidates.get_data(i));
     }
     fclose(output);
-    
+    */
     
     /*
     array_1d<double> correct_ans;

### 2014-07-21 aps4p5/aps_runner_ellipses.cpp
added placeholders in directories for aps_ellipse outputs

@@ -18,10 +18,10 @@ ncenters=atoi(argv[3]);
 char timingname[letters],outname[letters];
 
 //what is the name of the file where APS will store its timing information
-sprintf(timingname,"aps_output/ellipse/fullAnalysis/ellipse_d%d_c%d_timing.sav",dim,ncenters);
+sprintf(timingname,"timingFiles/ellipse_d%d_c%d_timing.sav",dim,ncenters);
 
 //what is the name of the file where APS will output the points it sampled
-sprintf(outname,"aps_output/ellipse/fullAnalysis/ellipse_d%d_c%d_output.sav",dim,ncenters);
+sprintf(outname,"outputFiles/ellipse_d%d_c%d_output.sav",dim,ncenters);
 
 
 if(seed<0){

### 2014-07-21 aps4p5/goto_tools.h
listed all of the imported .cpp files in readme.txt

@@ -20,17 +20,11 @@ struct Ran{
 //this structure will be based on the Xorshift random number generator
 // discovered by George Marsaglia and published in
 //Journal of Statistical Software, volume 8, no. 14 pp 1-6
-//
-//those who have access would do well to replace it with the 
-//structure Ran from 
+
+//parameters are drawn from the table on page 347 of 
 //Numerical Recipes (3rd edition) 
 //William H. press, Saul A. Teukolsky, William T. Vetterling, Brian P. Flannery
 //Cambridge University Press, 2007
-//p 342
-
-
-//parameters are drawn from the table on page 347 of Numerical Recipes
-
 
 unsigned long long x;
 Ran(unsigned long long seed){

### 2014-07-21 aps4p5/drafts/draft_140719.tex
latest draft of paper

@@
 function $f$ which is an implicit function of $\vec{\theta}$.  The researcher
-is able to calculate some statistic (e.g. $\chi^2$) quantifying the fit of a 
-model $f(\vec{\theta})$ to the data and wishes to find a confidence limit
+is able to calculate a $\chi^2$ value quantifying the fit of a 
+model $f(\vec{\theta})$ to the data and wishes to find either a confidence limit
 defined by some rule, e.g $\chi^2\le\chi^2_\text{lim}$.  This is a general
@@
 \ref{sec:frequentism}.  We discuss alternatives for determining
-$\chi^2_\text{lim}$ in Section \ref{sec:chi} below.
+$\chi^2_\text{lim}$ in Section \ref{sec:chi} below.  We will also consider
+the case that the reseracher wishes to find Bayesian credible limt
+to some probability $(1-\alpha)\%$.
 
@@
 probability.  In this case, ``credible limits'' are phrased as target values
-for the probability integral, i.e. 
-$0.68$ for ``1-$\sigma$'' limits, $0.95$ for ``2-$\sigma$'' limits
+for the probability integral $(1-\alpha)\%$, i.e. 
+68\% for ``1-$\sigma$'' limits, 95\% for ``2-$\sigma$'' limits
 and so on \cite{mcmc}.
@@
 associated with traditional MCMC methods (see Section \ref{sec:wmap}).
-We show that APS is capable of yielding both Frequentist confidence
-limits (which we define in Appendix \ref{sec:frequentism}) as well as more traditional
-Bayesian credible limits on the space of $\vec{\theta}$.
+We show that, while APS is designed specifically to
+yield Frequentist confidence
+limits (which we define in Appendix \ref{sec:frequentism}),
+it is also capable of roughly approximating
+Bayesian credible limits on the space of
+$\vec{\theta}$ (see Sections \ref{sec:bayes},\ref{sec:toy} and \ref{sec:wmap}).
+This rough approximation may principally be of use in cases where little 
+prior information about $\vec{\theta}$ exists and the researcher
+requires APS's robustness against multi-modality to be certain that she
+has discovered all of the best fits to her data.
 
@@
 limits of MCMC, which represent an integral over the sampled points containing some
-fraction of the total posterior probability.
+fraction of the total posterior probability.  We present an alternative
+interpretation for yielding credible limits from APS in Section \ref{sec:bayes}.
 
@@
 
-The first possibility, a user-defined $\chi^2_\text{lim}$ is based
+The first possibility, a user-defined $\chi^2_\text{lim}$, is based
 on the theory of Frequentist confidence intervals, which we briefly discuss
@@
 Section \ref{sec:simplex} below outlines an additional means of search
-by which our APS is capable of efficiently locating $\chi^2_\text{lim}$.
+by which our expanded APS is capable of efficiently locating $\chi^2_\text{lim}$.
 
@@
 unknown regions of parameter space might slow down convergence in the case of uni-modal
-likelihood functions.  Therefore, we propose the following modified search
-to ensure that it adequately characterizes known regions of high likelihood while
+likelihood functions.  Therefore, we augment steps (1A-3A) with
+the following modified search algorithms
+to ensure that APS adequately characterizes known regions of high likelihood while
 simultaneously searching for new regions to explore.
@@
 $\vec{\theta}\rightarrow\chi^2$ fuction as it searches.  
-To facilitate this discussion, we introduce the folling notations.
+To facilitate this discussion, we introduce the following notation.
 Symbols denoted with $\vec{x}$ represent points in parameter space.  Variables denoted with
@@
 \begin{itemize}
-\item (1B) If fewer than $N_p$ focused searches have been proposed around this minimum, 
+\item (1B) If fewer than $N_p$ focused searches have been proposed about $\vec{M}$, 
 randomly choose a point from a small sphere surrounding $\vec{M}$ in parameter space and 
@@
 along the line connecting the proposed point to $\vec{M}$ to find a point which does
-satisfy $\chi^2=\chi^2_\text{lim}$, i.e. find a point on that line with
+satisfy $\chi^2=\chi^2_\text{lim}$, i.e. find a point on that line for which
 $\chi^2<\chi^2_\text{lim}$ -- often $\vec{M}$ itself -- and a point along that line
-with $\chi^2>\chi^2_\text{lim}$ and iteratively step half the distance between them until
-arriving at the desired $\chi^2_\text{lim}$ \cite{minuit}.  The points discovered this way \
+for which $\chi^2>\chi^2_\text{lim}$ and iteratively step half the distance between them until
+arriving at the desired $\chi^2_\text{lim}$ \cite{minuit}.  The points discovered in this way \
 with $\chi^2=\chi^2_\text{lim}$ are stored in the set $\{\hat{B}(\vec{M})\}$ of 
@@
 \\
-\item(5B) At each point $\vec{v}$ above, evaluate equation (\ref{eqn:sstat}).  
-Evaluate $\chi^2$ at the point $\vec{v}$ which maximizes $S$. If
+\item(5B) At each point $\vec{v}$ above, find the $N_g$ nearest neighbor points as in step (2A) above
+and evaluate equation (\ref{eqn:sstat}).  
+Find the point $\vec{v}$ which maximizes $S$ and evaluate $\vec{\theta}\rightarrow\chi^2$ at
+that point.  If
 $\chi^2 \neq \chi^2_\text{lim}$, bisection is again used and a new boundary point is found.
+Store the resulting $\chi^2=\chi^2_\text{lim}$ point in $\{\hat{B}(\vec{M})\}$
 \end{itemize}
 
-To reiterate, our version of APS alternates between performing steps (1A-3A) and steps
+The modification suggested in section \ref{sec:focus} is designed to explore 
+$\chi^2=\chi^2_\text{lim}$ surface by starting from its extreme corners and stepping
+perpendicularly to its ``radii'' (the vector $\vec{b}-\vec{M}$ would be a radius of
+$\chi^2=\chi^2_\text{lim}$ described a sphere), with the hopes of being able
+to extend the bounds of $\{\hat{B}_\text{MaxMin}(\vec{M})\}$.  In this way,
+our extended APS attempts to plum the full Frequentist confidence limit $\chi^2\le\chi^2_\text{lim}$
+as rapidly as possible.  However, we do not want this to come at the expense of the wide-ranging
+exploration of steps (1A-3A).  Our version of APS alternates between performing steps (1A-3A) and steps
 (1B-5B).  If there are $N_m$ distinct known local minima in $\{\hat{\Gamma}\}$, 
@@
 
-The modification suggested in section \ref{sec:focus} is designed to explore 
-$\chi^2=\chi^2_\text{lim}$ surface by starting from its extreme corners and stepping
-perpendicularly to its ``radii'' (the vector $\vec{b}-\vec{M}$ would be a radius of
-$\chi^2=\chi^2_\text{lim}$ described a sphere).  It is, however, also possible
+It is, however, also possible
 that the $S$-maximization in steps (1A-3A) could also learn something about the
 shape of the $\chi^2=\chi^2_\text{lim}$ contour, even if it does not place
-points directly on that contour.  To explore that possibility, we add the
-following steps to the end of the sequence (1A-3A).
+points directly on that contour.  To explore that possibility, our
+extended APS performs the
+following modified search after each iteration of steps (1A-3A).
 
@@
 where a unit sphere centered on $\vec{M}$ intersects that line.  Call this set
-of projectsion $\{\hat{U}\}$.
+of projections $\{\hat{U}\}$.
 \\
 \item(2C) Call the point for which $\chi^2$ has been evaluated in step (3A)
-$\vec{P}$., find the
+$\vec{P}$.  Find the
 point $\vec{M}$ in $\{\hat{\Gamma}\}$ that is nearest to $\vec{P}$ in parameter space.
@@
 then perform bisection between $\vec{M}$ and the point evaluated in step (3A).
+Add the resulting $\chi^2=\chi^2_\text{lim}$ point to $\{\hat{B}(\vec{M})\}$.
 \end{itemize}
 
-This way, points on the $\chi^2=\chi^2_\text{lim}$ contour can be found whose
-directions away from $\vec{M}$ in parameter space have not yet been explored.
+Step (4C) ensures that, if the point $\vec{P}$ is in a novel direction from
+$\vec{M}$, APS will find a point on the $\chi^2=\chi^2_\text{lim}$ contour
+in that direction.  Because $\vec{P}$ was
+chosen by maximizing equation (\ref{eqn:sstat}), we assume that there was something
+``interesting'' about that direction from $\vec{M}$ (perhaps a sharp corner or distended
+extension of the $\chi^2=\chi^2_\text{lim}$ contour).  In this way, our extended APS
+will attempt to cover the surface of $\chi^2=\chi^2_\text{lim}$ in the most informative
+way possible.
+
 Because we expect the unit spheres surrounding the points $\vec{M}$ to become more densely
@@
 \chi^2_\text{lim}&\le&\chi^2\le\chi^2_\text{threshold}\label{eqn:simplexCriterion}\\
-\chi_\text{threshold}&=&\chi^2_\text{min}+G\times(\chi^2_{1/10}-\chi^2_\text{min})
+\chi_\text{threshold}&\equiv&\chi^2_\text{min}+G\times(\chi^2_{1/10}-\chi^2_\text{min})
 \label{eqn:threshold}
@@
 Because the Nelder-Mead simplex minimizer can be very time intensive, we balance our
-algorithm by demanding that the number of $\chi^2$ evaluations devoted to all
-other variants of search (i.e. steps (1A-3A) and (1B-5B)) equal the number of
+algorithm by keeping track of how many times $\vec{\theta}\rightarrow\chi^2$
+is called by the different searches and
+demanding that the number of $\chi^2$ evaluations devoted to all
+other variants of search (i.e. steps (1A-3A), (1B-5B), and (1C-4C)) equal the number of
 $\chi^2$ evaluations devoted to the simplex minimizer before a new simplex
@@
 If the simplex minimizer converges to a point such that
-$\chi^2<\chi^2_\text{lim}$, our APS determines whether or not it
+$\chi^2<\chi^2_\text{lim}$, our extended APS determines whether or not it
 has found a new local minimum to be added to $\{\hat{\Gamma}\}$ by comparing the end
@@
 \label{eqn:midpt}
-\vec{m}=0.5\times\left(\vec{y}+\vec{M}\right)
+\vec{m}(\vec{M})=0.5\times\left(\vec{y}+\vec{M}\right)
 \end{equation}
-and $\vec{M}$.  If, for all $\vec{M}$, 
-$\chi^2$ at $\vec{m}$ is greater
-than $\chi^2_\text{lim}$, then it is deemed that $\vec{\gamma}$ is a new local
-minimum and is added to $\{\hat{\Gamma}\}$.  If not, $\vec{\gamma}$ is assumed to be a
+If, for all $\vec{M}$, 
+$\chi^2$ at $\vec{m}(\vec{M})$ is greater
+than $\chi^2_\text{lim}$, then it is deemed that $\vec{y}$ is a new local
+minimum and is added to $\{\hat{\Gamma}\}$.  If not, $\vec{y}$ is assumed to be a
 part of the one of the previously identified loci of low $\chi^2$ and it is not
@@
 As a further guard against excessive simplex searching, all points $\vec{w}$ 
-that pass the test in equation(\ref{eqn:simplexCriterion}) are further required
+that pass the test in equation (\ref{eqn:simplexCriterion}) are further required
 to satisfy the condition
 \begin{eqnarray}
-\chi^2(\vec{z})&\le&\chi^2(\vec{w})-0.25\times\left(\chi^2(\vec{w})-\chi^2(\vec{y})\right)\\
+\chi^2(\vec{z})&>&\chi^2(\vec{w})-0.25\times\left(\chi^2(\vec{w})-\chi^2(\vec{y})\right)\\
 \vec{z}&\equiv&0.5\times\left(\vec{w}+\vec{y}\right)\\
@@
 \end{eqnarray}
-before it is allowed to seed a new simplex search.  In this way, our APS avoids seeding simplex
+before they are allowed to seed a new simplex search.  In this way, our 
+extended APS avoids seeding simplex
 searches with points that are inside of already known regions of low $\chi^2$.
@@
 function available at \verb|https://github.com/uwssg/APS/|.
-The code is written in C++.  It provides the option for several
+The code is written in C++.  It is designed for users to use
+with any $\vec{theta}\rightarrow\chi^2$ function.  Users need only
+to write a sub-class of the \verb|chisquared| class defined in
+\verb|chisq.h| and \verb|chisq.cpp| which evaluates their desired
+$\vec{\theta}\rightarrow\chi^2$ and pass an instantiation of this
+sub-class to an instantiation of the \verb|aps| class defined in
+\verb|aps.h| and \verb|aps.cpp|.  It is this \verb|aps| class which
+does the work of performing the searches described above.
+The user should not need to modify this class (though she is 
+certainly welcome to).
+
+Our \verb|aps| class provides the option for several
 user-specified parameters.  We list them below along with the C++ 
@@
 method \verb|aps.set_max| and \verb|aps.set_min|.  These bounds will constrain the range of
-points sampled in step 2A.  It will not contraint searches performed with bisection or the 
+points sampled in step 2A.  They will not contraint searches performed with bisection or the 
 simplex minimizer.
@@
 difference between the maximum and minimum values set by \verb|aps.set_max| and
-\verb|aps.set_min|.
+\verb|aps.set_min|.  Setting a characteristic length scale will principally
+effect what points are considered nearest neighbors for the purposes of step 2A.
+If the characteristic length scale of a parameter is large, that parameter will
+be down-weighted when calculating the parameter-space distance between two points.
+If the characteristic length scale of a parameter is small, that parameter will
+be up-weighted in length calculations.
 \end{itemize}
@@
 
-Originally, APS was conceived as a way to draw Frequentist confidence limits. 
-$\chi^2_\text{lim}$ is determined, either in a vacuum or as some $\Delta\chi^2$ away from
-$\chi^2_\text{min}$ (see Appendix \ref{sec:frequentism} for a more thorough discussion).
+Originally, APS was conceived as a way to draw Frequentist confidence limits $\chi^2\le\chi^2_\text{lim}$. 
 This is by far the most direct way to interpret APS outputs: a list of points is given with
 corresponding values of $\chi^2$.  All of the points with $\chi^2\le\chi^2_\text{lim}$ are
-taken to circumscribe the desired confidence limit.  There is, however, also a way to use the
-outputs from APS to draw a Bayesian credible limit.
+taken to cover the desired confidence limit.  There is, however, also a way to use the
+outputs from APS to draw an approximate Bayesian credible limit.
 
@@
 
-APS outputs cannot be used this way, since APS does not select points to evaluate
-probabilisticaly, but in an attempt to recreate the behavior of the $\chi^2$ function in
+APS outputs cannot be used this way, since APS does not select its points
+probabilisticaly, but in an attempt to recreate the behavior of the 
+$\vec{\theta}\rightarrow\chi^2$ function in
 interesting regions of parameter space.  We can still, however, use these points to approximate
@@
 
+For each $\vec{a}$ in $\{\hat{A}\}$:
 \begin{itemize}
-\item(1D) For each point $\vec{a}$ in $\{\hat{A}\}$ initialize two vectors $\vec{a}_\text{max}$ and
-$\vec{a}_\text{min}$.
+\item(1D) Initialize two vectors $\vec{a}_\text{max}$ and
+$\vec{a}_\text{min}$.  Set all of the values in these vectors to some nonsense number.
 
@@
 \item(3D) For each point $\vec{n}$ in $\{\hat{N}\}$, calculate the vector
-$\vec{\delta}=\|\vec{n}-\vec{a}|$ (where the vertical bars denote the absolute value of the
-components of $\vec{\delta}$, not any kind of vector normalization).  For each component
+$\vec{\delta}$ such that $\delta_i=|n_i-a_i|$ (here the vertical bars denote the absolute value).  
+For each component
 $\delta_i$ of $\vec{\delta}$, starting with the largest and working towards the smallest:
@@
 \item(3Dc) If both $a_\text{max,i}$ and $a_\text{min,i}$ have already been set, move on to the
-next largest component of $\vec{\delta}$ and repeat at step (3Da).
+next largest $\delta_i$ and repeat at step (3Da).
 \end{itemize}
@@
 \\
-\item(5D) $\vec{a}_\text{max}$ an $\vec{a}_\text{min}$ should now describe an asymmetric,
+\item(5D) $\vec{a}_\text{max}$ adn $\vec{a}_\text{min}$ should now describe an asymmetric,
 rectangular hyperbox about $\vec{a}$.  To symmetrize it, step through all $N_p$ dimensions
@@
 $d_\text{max}=a_{\text{max},i}-a_i$ are equal to the minimum of $\{d_\text{max},d_\text{min}\}$
-from the asymmetric hyperbox.
+from the asymmetric hyperbox.  Return to step (1D) and repeat for the next $\vec{a}$.
 \end{itemize}
@@
 At this point, each $\vec{a}$ in $\{\hat{A}\}$ should have a corresponding symmetric hyperbox
-surrounding it.  To integrate the posterior probability density, we will assume that all of the
-parameter space points in each hyperbox correspond to a $\chi^2$ equal to $\chi^2(\vec{a})$.
+surrounding it described by the corresponding 
+$\vec{a}_\text{max}(\vec{a})$ and $\vec{a}_\text{min}(\vec{a})$.  
+To integrate the posterior probability density, we will assume that all of the
+parameter space points in each hyperbox correspond have $\chi^2=\chi^2(\vec{a})$
+($\chi^2)\vec{a})$ being already known because $\vec{a}$ is one of the points for which APS evaluated
+$\vec{theta}\rightarrow\chi^2$).
 The posterior probability associated with each box is then
@@
 \end{eqnarray}
-where $V(\vec{a})$ is the hypervolume associated with the hyperbox surrouning $\vec{a}$.
+where $V(\vec{a})$ is the hypervolume of with the hyperbox surrounding $\vec{a}$.
 To plot the $(1-\alpha)\%$ credible limit, one need only plot the points in hyperspace that
@@
 We test this algorithm on a function with an exactly known credible limit
-in Section \ref{sec:toy} and on a real, physical dataset in Section \ref{sec:wmap}.  In 
-Section \ref{sec:toy}, we find that our results agree with the known credible limit.
-In Section \ref{sec:wmap}, we find that our results agree with results derived from traditional
+in Section \ref{sec:toy} (see Figures \ref{fig:toyBayes2}-\ref{fig:toyBayes4}) 
+and on a real, physical dataset in Section \ref{sec:wmap} (see Figure \ref{fig:contours_bayes}).  
+In  Section \ref{sec:toy}, we find that our results agree with the known credible limit.
+In Section \ref{sec:wmap}, we find that our results roughly approximate those derived from traditional
 Markov Chain Monte Carlo methods.
@@
 \begin{eqnarray}
-\mu&=&\sum_{i,j} C_{qi}C^{-1}_{ij}\chi^{2,(j)}\nonumber\\
-\sigma^2&=&C_{qq}-\sum_{i,j}C_{qi}C^{-1}_{ij}C_{jq}\label{sig}
+\mu&=&\sum_{i,j}^{N_g} C_{qi}C^{-1}_{ij}\chi^{2,(j)}\nonumber\\
+\sigma^2&=&K\times\left(C_{qq}-\sum_{i,j}^{N_g}C_{qi}C^{-1}_{ij}C_{jq}\right)\label{sig}
 \end{eqnarray}
@@
 \label{mu}
-\mu=\bar{\chi}^2+\sum_{i,j} C_{qi}C^{-1}_{ij}
+\mu=\bar{\chi}^2+\sum_{i,j}^{N_g} C_{qi}C^{-1}_{ij}
 (\chi^{2,(j)}-\bar{\chi}^2)
@@
 
-The form of the covariance function 
+The form of the covariogram
 $C_{ij}(\vec{\theta}^{(i)},\vec{\theta}^{(j)})$ must be
@@
 the $n$th parameter taken from the $N_g$ points used to seed the Gaussian process and $\ell_n$
-is a ``hyper-parameter'' associated with the $n$th parameter.  We both the normalization
+is a ``hyper-parameter'' associated with the $n$th parameter.  We discuss the normalization
 paramter $K$ below in section \ref{sec:kriging}.  We discuss the setting of
@@
 that distant points should
-not be very correlated.  We will use this covariogram in
-our tests in Section \ref{sec:toy} and \ref{sec:wmap}.
+not be very correlated.
 
@@
 $(\text{max}_n-\text{min}_n)$ in each dimension.  $\sigma_0$ and $\sigma_1$
-are hyperparameters.  We also consider this
-covariogram in Section \ref{sec:wmap} and find no appreciable difference
-in performance when compared to the use of covariogram (\ref{eqn:GaussianCovar}).
+are hyperparameters.  
+In Section \ref{sec:wmap}, we test APS using both covarigorams (\ref{eqn:GaussianCovar})
+and (\ref{eqn:NNCovar}) and find no appreciable difference
+in the resulting parameter constraints.
 
@@
 To maximize the probability of the observed data (i.e. maximizing the value
-of $P(\vec{\chi}^2)$) we must set $K$ equal to
+of $P(\vec{\chi}^2)$ in equation \ref{eqn:likelihood}) we must set $K$ equal to
 \begin{equation}
@@
 
-The covariance functions $C_{ij}$ presented above (indeed, any covariance function)
+The covariograms $C_{ij}$ presented above in equations (\ref{eqn:GaussianCovar})
+and (\ref{eqn:NNCovar}) (indeed, any covariogram)
 each contain arbitrary hyper-parameters whose values can be set to fine-tune
@@
 the $\vec{\theta}\rightarrow\chi^2$ function and compare this amortized time to the average
-amount of time spent on one call to $\vec{theta}\rightarrow\chi^2$.  APS us only allowed to
-reset its hyper-parameters if the difference between these two times is less than 0.1 second.
+amount of time spent on one call to $\vec{\theta}\rightarrow\chi^2$.  APS us only allowed to
+reset its hyper-parameters if the difference between these two times is less than 0.1 second
+(of course, we do require that APS optimize its hyper parameters at least once, so that
+we can have confidence in our Gaussian Process model).
 
@@
 
-To test the performance of our APS algorithm, we ran APS on iterations of this
+To test the performance of APS, we ran APS on versions of this
 $\vec{\theta}\rightarrow\chi^2$ function with 2, 3, and 4 discrete minima in $\chi^2$.  We
 allowed APS to sample a total of 10,000 points in the 5-dimensional parameter space.  
-We had APS adaptively set $\chi^2_\text{lim}=\chi^2_\text{min}+11.0$ -- $11.0$ being the $95\%$
+We had APS adaptively set $\chi^2_\text{lim}=\chi^2_\text{min}+\Delta\chi^2$, 
+$\Delta\chi^2=11.0$ being the 95\%
 confidence limit bound for $\chi^2$ probability distribution with 5 degrees of freedom.
-Figures
-\ref{fig:toyFreq2}, \ref{fig:toyFreq3}, and \ref{fig:toyFreq4} plot the
+We used a Mat\`ern covagiogram with $\nu=3/2$ for our Gaussian Process (Rasmussen and Williams equation
+4.17). Figures \ref{fig:toyFreq2}, \ref{fig:toyFreq3}, and \ref{fig:toyFreq4} plot the
 $\chi^2\le\chi^2_\text{lim}$ points found by APS (black points) against the
@@
 
-The tests above demonstrate APS's ability to correctly determin credible and
+The tests above demonstrate APS's ability to correctly determine credible and
 confidence limits.  They do not, however, address the question of APS's chief
@@
 Monte Carlo searches on these $\chi^2$ function, each run consisting of five
-independent chains.  These searches were not nearly as successful as APS at
+independent chains.  We allowed the MCMCs to run
+until they had sampled 100,000 points in parameter space.  
+These searches were not nearly as successful as APS at
 locating all of the low-$\chi^2$ regions.  In the case of the 2-mode function,
@@
 Section \ref{sec:toy} demonstrates the robustness of APS against multi-modal
-$\chi^2$ functions.  In this section, we demonstrate 
-that, on well-behaved likelihood functions, APS' rate of convergence is
-comparable to that of MCMC.  
-We also show that APS can be used on actual astrophysical
-problems by using for our $\chi^2$ function the likelihood function on the
+$\chi^2$ functions.  However, it does have some short-comings as a test of
+APS's abilities.  As was already noted, there are no interesting parameter
+degeneracies in equation (\ref{eqn:toychi}).  Furthermore,
+because this function was constructed artificially, there is no noise.  At
+any point in parameter space, $\chi^2$ falls smoothly and monotonically towards
+the associated local minimum of $\chi^2$.  To really have confidence in APS
+abilities, we must test it on actual physical data drawn from the universe,
+with all of the noisy and ill-behaved properties that entails.
+To that end, we ill no consider for our $\vec{\theta}\rightarrow\chi^2$ 
+function the likelihood function on the
 7-year data release of the WMAP CMB satellite \cite{wmap7,wmap7likelihood}.
@@
 \item$\tau$ -- the optical depth to the surface of last scattering.  $\tau$ cannot be
-constrained without polarizationd data, which we do not consider here.  However, it functions
+constrained without polarization data, which we do not consider here.  However, it functions
 as a nuisance parameter to be marginalized over.  The fact that APS can handle it as well as
-MCMC is an important test of APS' feasibility.
+MCMC is an important test of APS's feasibility.
 \\
@@
 Evaluating the $\chi^2$ using the likelihood code provided by the WMAP team
-\cite{wmap7likelihood} takes an additional 2 seconds.  Hence the need
+\cite{wmap7likelihood} takes an additional 2 seconds (this can, of course, be sped up by using a faster
+machine and multiple cores).  Hence the need
 for an algorithm like APS or MCMC to accelerate the search.
 We will consider both equations (\ref{eqn:GaussianCovar}) and (\ref{eqn:NNCovar})
-as covariogram for our Gaussian Process, 
+as the covariogram for our Gaussian Process, 
 though we find that the choice has little effect on our
-results.
+results.  We will also consider setting $\chi^2_\text{lim}=\chi^2+\Delta\chi^2$ 
+as wel as setting $\chi^2_\text{lim}$ by hand.
 
@@
 many tens (or hundreds) of thousands of points to be sampled after $\chi^2_\text{min}$
-has been discovered.  Because APS is designed to locate points at the precise value
-$\chi^2=\chi^2_\text{lim}$, it requires much fewer samples after discovering 
-$\chi^2_\text{min}$ to infer the correct confidence limit.
+has been discovered.  Because it is learning the shape of the function
+directly, APS does not impose this requirement.
 
@@
 $\vec{\theta}\rightarrow\chi^2$ can be approximated as a Gaussian Process.
-Though the squared exponential covariogram gives a better fit to the $\chi^2$ function,
+Though the squared exponential covariogram (\ref{eqn:GaussianCovar})
+gives a better fit to the $\chi^2$ function than the arcsine covariogram
+(\ref{eqn:NNCovar}),
 we see in Figures \ref{fig:contours_gauss} and \ref{fig:contours_nn}
@@
 We consider both an adaptive $\chi^2_\text{lim}=\chi^2_\text{min}+12.6$
+(12.6 is the $\Delta\chi^2$ corresponding for 95\% confidence limit in the case
+of 6 parameters)
 and an absoulte $\chi^2_\text{lim}=1283.3$ (this is the 95\% confidence limit
-for a $\chi^2$ probability distribution with 1199 degrees of freedom).
+for a $\chi^2$ probability distribution with the 1199 degrees of freedom corresponding
+to the 1199 $C_\ell$s in the data set).
 These are the black points in Figures \ref{fig:contours_gauss}, \ref{fig:contours_nn},
 and \ref{fig:contours_br} respectively.  These confidence limits overlap very nicely
-with the credible limits discovered by MCMC after many more $\chi^2$ evaluations,
+with the true credible limits discovered by MCMC,
 comporting the the idea that, in the large data limit, Bayesiand credible limits
@@
 by MCMC after only 50,000 $\chi^2$ evaluations.  These are the black regions in
-Figure \ref{fig:mcmc_comparison}.  As can be seen, MCMC has not yet totally converged
+Figure \ref{fig:mcmc_comparison}.  As can be seen from Figures \ref{fig:mcmc_1_4},
+\ref{fig:mcmc_1_5}, and \ref{fig:mcmc_4_5}, MCMC has not yet totally converged
 in all of the sub-spaces of cosmological parameter space.  This should be compared
-with the reasonably complete coverage of the control credible limits seen in
+with the reasonably coverage of the confidence limits seen in
 Figures \ref{fig:contours_gauss}-\ref{fig:contours_br}.
@@
 by APS after 50,000 evalutations of $\chi^2$ and determined by the method outlined in
-Section \ref{sec:bayes}.  While we do not see the clean equivalence between the APS-derived credible limits and
-the true credible limits that we saw in Figures \ref{fig:toyBayes2}-\ref{fig:toyBayes4},
-we do find that the APS-derived credible limits roughly sketch the bounds of the 
-true credible limit.  This, combined with its robustness against multi-modal likelihood
-functions, suggests that APS may, at the very least be useful in cases where very little
-prior information is available and exploring the paramter space is at least as important as
-integrating it to find a rigorous bound.
+Section \ref{sec:bayes}.  Unfortunately, we do not see here the convergence to the true
+credible limits that we say in Figures \ref{fig:toyBayes2}-\ref{fig:toyBayes4}.  While
+it does appear that APS has correctly identified the locations and general shapes of the
+true credible limits, it has, in most cases, drawn credible limits that are larger than
+those found by MCMC.  This should not be surprising.  APS was designed with Frequentist
+confidence limits in mind, and Figures \ref{fig:contours_gauss}-\ref{fig:contours_br}
+seem much better converged than Figure \ref{fig:contours_bayes}.  However, it does seem
+that APS might be useful in cases where little prior knowledge is known about $\vec{\theta}$
+and researchers need to be confident that they have discovered all of the low-$\chi^2$
+regions in a particular parameter space.  Once APS has been used to identify rough
+Bayesian credible limits, more detailed algorithms (like MCMC) can be run using the prior
+knowledge gained by APS to refine the credible limits.
 
@@
 \includegraphics[scale=0.15]{mcmc_comparison_0_1_50k.eps}
+\label{fig:mcmc_0_1}
 }
@@
 \includegraphics[scale=0.15]{mcmc_comparison_0_2_50k.eps}
+\label{fig:mcmc_0_2}
 }
@@
 \includegraphics[scale=0.15]{mcmc_comparison_0_4_50k.eps}
+\label{fig:mcmc_0_4}
 }
@@
 \includegraphics[scale=0.15]{mcmc_comparison_0_5_50k.eps}
+\label{fig:mcmc_0_5}
 }
@@
 \includegraphics[scale=0.15]{mcmc_comparison_1_2_50k.eps}
+\label{fig:mcmc_1_2}
 }
@@
 \includegraphics[scale=0.15]{mcmc_comparison_1_4_50k.eps}
+\label{fig:mcmc_1_4}
 }
@@
 \includegraphics[scale=0.15]{mcmc_comparison_1_5_50k.eps}
+\label{fig:mcmc_1_5}
 }
@@
 \includegraphics[scale=0.15]{mcmc_comparison_2_4_50k.eps}
+\label{fig:mcmc_2_4}
 }
@@
 \includegraphics[scale=0.15]{mcmc_comparison_2_5_50k.eps}
+\label{fig:mcmc_2_5}
 }
@@
 \includegraphics[scale=0.15]{mcmc_comparison_4_5_50k.eps}
+\label{fig:mcmc_4_5}
 }
@@
 
-In Section \ref{sec:toy} we indicated that APS is more robust
-than MCMC against multi-modal likelihood functions.
-In Section \ref{sec:wmap} we showed that, on more well-behaved
-likelihood functions, APS discovers comparable parameter constraints
-as MCMC and that APS does not converge any more slowly than does MCMC.
-We also showed a great deal of qualitative similarity but quantitative
-difference between the Bayesian credible limits drawn by MCMC
-and the frequentist confidence limits drawn by APS.  The different
-convergence properties and different outputs of APS and MCMC lend themselves
-to different use cases.
-
-As was demonstrated in Figure \ref{fig:chimin}, MCMC is significantly
-more efficient than APS at finding the true value of $\chi^2_\text{min}$
-in a given parameter space.  This makes it ideally suited to problems where
-the uncertainty of the data set is poorly understood and must be fit
-for.  However, as we saw in Figure \ref{fig:toy}, this rapid convergence to
-$\chi^2_\text{min}$ can also result in a myopic disregard for other regions of
-interest in parameter space.  Once MCMC has been used to learn the uncertainty
-model of the data, it may be advisable to run APS as a guarantor that no
-alternative best-fit models have been ignored by MCMC.
-
-In cases where the uncertainty model is well-understood and the contraints in
-parameter space can safely be described by either the likelihood ratio's
-$\chi^2\le\chi^2_\text{min}+\Delta\chi^2$ or an imposed
-$\chi^2\le\chi^2_\text{lim}$, APS' robustness against multi-modality and
-comparatively rapid convergence in a confidence limit sense (recall Figures
-\ref{fig:contours_gauss} through \ref{fig:contours_br}) make it ideal for rapidly
-drawing constraints on parameter space.
-
-One advantage of MCMC that APS cannot overcome is its comparative ease to
+In Section \ref{sec:toy} we indicated that, consistent with the findings
+of Bryan {\it et al}. (2007), APS is more robust
+than MCMC against multi-modal likelihood functions (at least on the idealized
+cartoon $\vec{\theta}\rightarrow\chi^2$ functions considered).
+In Section \ref{sec:wmap} we showed that on realistic data, APS
+is able to yield Frequentist confidence limits that are comparable to
+Bayesian credible limits with significantly fewer calls to $\vec{\theta}\rightarrow\chi^2$
+than MCMC (recall that APS required only 50,000 $\chi^2$ evaluations to yield Figures
+\ref{fig:contours_gauss}-\ref{fig:contours_br}).  However, the Bayesian credible
+limits yielded by APS remain rough and imprecise.  We believe that this performance
+can still be of use to the community, especially when trying to evaluate proof-of-concept
+parameter constraints (perhaps for forecasting future parameter constraints)
+with very expensive $\vec{\theta}\rightarrow\chi^2$ functions on poorly-understood
+parameter spaces.  Certainly, for researchers interested in plotting Frequentist confidence limits,
+APS is a viable option.
+
+There remain, however, several concerns to be addressed in the use of APS.  Unlike MCMC, there is
+no obvious criterion for convergence of the APS algorithm.  One possibility is to interpret the Frequentist
+confidence limit as a hyperbox (i.e. keep track of the allowable maximum and minimum values of each parameter
+as $\chi^2=\chi^2_\text{lim}$ is explored) and watch the growth of the volume of that hyperbox as a function
+of the number of $\vec{\theta}\rightarrow\chi^2$ evaluations.  Figure \ref{fig:convergence} plots this
+convergence metric for the APS run in Figure \ref{fig:contours_gauss}.  While it appears that APS may have
+converged after $\sim$50,000 evaluations, the volume of the Frequentist confidence limit continues to grow,
+though Figure \ref{fig:contours_gauss} already shows good coverage of the control Bayesian credible limit. 
+It is, of course, possible that we are seeing the effect of the slight difference between the Bayesian
+credible limit at the Frequentist confidence limt, even in the large-data case.
+
+\begin{figure}
+\includegraphics[scale=0.3]{convergencePlot.eps}
+\caption{
+The parameter space volume of the Frequentist confidence limit
+in figure \ref{fig:contours_gauss} as a function of the number of
+$\vec{\theta}\rightarrow\chi^2$ evaluations.
+}
+\label{fig:convergence}
+\end{figure}
+
+Another concern is that, while APS was explicitly designed to rapidly converge
+to the Frequentist confidence limit (where rapidity is measured in the number of
+$\vec{\theta}\rightarrow\chi^2$ evaluations made), the searches in Section \ref{sec:algorithm}
+are complicated enough that they will add extra clock time to each $\vec{\theta}\rightarrow\chi^2$
+evaluation.  We have attempted to balance APS so that it does not rely to heavily on expensive calculations.
+However, some overhead is inevitably required.  Figure \ref{fig:time} shows the average overhead
+time in seconds added by APS to each $\vec{\theta}\rightarrow\chi^2$ evaluation as a function
+of the number of evaluations.  As the number of evaluations grow and the number of points APS must search
+when constructing its Gaussian Processes also grows, so does this extra time (the initial spike is due to the
+first expensive call to the hyperparameter optimization in Section \ref{sec:hyperparams}).
+Though the extra time stays within the bounds of a few 0.1 seconds per $\vec{\theta}\rightarrow\chi^2$
+evaluation, this may still be too expensive in the case of functions for whom a direct
+$\vec{\theta}\rightarrow\chi^2$ takes a vanishingly small amount of time.
+
+\begin{figure}
+\includegraphics[scale=0.3]{timeConvergencePlot.eps}
+\caption{
+The extra time added by APS to a $\vec{\theta}\rightarrow\chi^2$ 
+as a function of the number of
+$\vec{\theta}\rightarrow\chi^2$ evaluations.
+}
+\label{fig:time}
+\end{figure}
+
+Finally, one advantage of MCMC that APS cannot overcome is its comparative ease to
 implement.  To this end,
 we make our code available at \verb|https://github.com/uwssg/APS/|.  
-The code is presented as a series of C++ modules with directions
+The code is presented as a series of C++ classes with directions
 indicating where the user can interface with 

### 2014-08-04 aps4p5/chisq_aps4p5.cpp
wmap_likelihood function now only applies minimum and maximum cut if the mi and max are reasonable

@@ -86,7 +86,8 @@ double wmap_likelihood::operator()(array_1d<double> &v) const{
   FILE *output;
 
   for(i=0;i<dim;i++){
-      if(v.get_data(i)<mins.get_data(i) || v.get_data(i)>maxs.get_data(i)){
+      if((v.get_data(i)<mins.get_data(i) && mins.get_data(i)<chisq_exception) || 
+      (v.get_data(i)>maxs.get_data(i) && maxs.get_data(i)>-1.0*chisq_exception)){
           time_spent+=double(time(NULL))-before;
       
           return 2.0*chisq_exception;

### 2014-08-04 aps4p5/mcmc_test.cpp

@@ -53,14 +53,14 @@ for(i=0;i<dim;i++){
 
 
 printf("time to start declaring stuff\n");
-mcmc mcmc_test(dim,nchains,"chains/ellipse_140718_chains",min,max,sig,2.0,&chaos);
+mcmc mcmc_test(dim,nchains,"chains/ellipse_140804_chains",min,max,sig,2.0,&chaos);
 mcmc_test.set_chisq(&chifn,1);
 
 printf("done with constructor\n");
 
 //mcmc_test.set_statname("chains/integrable_test_mcmc_status.sav");
-mcmc_test.set_statname("chains/ellipse_140718_status.sav");
-mcmc_test.set_diagname("chains/ellipse_140718_diagnostic.sav");
+mcmc_test.set_statname("chains/ellipse_140804_status.sav");
+mcmc_test.set_diagname("chains/ellipse_140804_diagnostic.sav");
 mcmc_test.begin_update(1000);
 mcmc_test.step_update(500);
 //mcmc_test.cutoff_update(30000);

### 2014-08-04 aps4p5/aps.cpp
there was a typo in the order of statistics output by write_pts()

@@ -162,7 +162,7 @@ void aps::start_timingfile(){
     
     fprintf(output,"ct_aps time_aps -- ");
     fprintf(output,"ct_simplex time_simplex -- ");
-    fprintf(output,"time_refactoring -- time_optimizing_gp -- ");
+    fprintf(output,"time_optimizing_gp -- time_refactoring -- ");
     fprintf(output,"chisq_min target -- volumes -- called_wide ");
     fprintf(output,"called_focus");
     fprintf(output,"\n");

### 2014-08-04 aps4p5/mcmc_extraction_runner.cpp
some silly changes to mcmc_extraction_runner.cpp

@@ -12,7 +12,7 @@ nparams = 81;
 test.set_nchains(4);
 test.set_nparams(nparams);
 
-test.set_cutoff(12500);
+//test.set_cutoff(12500);
 //control has a total 453,051 samples
 
 test.set_chainname("/Users/noldor/physics/recreate_getdist/ieuchains_1304/wmap7_learn");
@@ -26,7 +26,7 @@ printf("thinby %d\nused %d\nkept %d\n",test.get_thinby(),test.get_total_used(),t
 printf("rows %d\n",test.get_total_rows());
 printf("best_covar %e\n",test.get_best_covar());
 
-test.print_samples("test_wmap7_samples50k.sav");
+test.print_samples("test_wmap7_samplescontrol.sav");
 
 kde kde_test;
 kde_test.set_data(test.get_samples());
@@ -51,8 +51,11 @@ ix.set(3,2);
 ix.set(4,17);
 ix.set(5,20);
 
+
 char outname[letters];
 int i,j;
+
+
 for(i=0;i<6;i++){
     for(j=i+1;j<6;j++){
         if(i!=3 && j!=3){
@@ -61,13 +64,13 @@ for(i=0;i<6;i++){
 
             //kde_test.plot_boundary(ix.get_data(i),dx.get_data(i),ix.get_data(j),dx.get_data(j),0.95,outname,3);
             
-            sprintf(outname,"mcmc_processed/wmap_scatter_%d_%d_50k.sav",i,j);
+            sprintf(outname,"mcmc_processed/wmap_scatter_%d_%d_control.sav",i,j);
             kde_test.plot_density(ix.get_data(i),dx.get_data(i),ix.get_data(j),dx.get_data(j),0.95,outname,3);
         }
     }
 }
 
-test.plot_delta("mcmc_good_pts_test50k.sav",0.5*12.61);
+test.plot_delta("mcmc_good_pts_testcontrol.sav",0.5*12.61);
 
 
 array_1d<double> RR,VV,WW,mean,var;
@@ -81,7 +84,7 @@ test.calculate_mean(mean,var);
 
 for(i=0;i<nparams;i++){
     if(fabs(var.get_data(i)/mean.get_data(i))>1.0e-20){
-        printf("%d %e -- %e %e\n",i,RR.get_data(i),mean.get_data(i),var.get_data(i));
+        printf("%d %e -- %e %e\n",i,RR.get_data(i),mean.get_data(i),sqrt(var.get_data(i)));
     }
 }
 

### 2014-08-04 aps4p5/drafts/draft_140804.tex
new draft added

--- 2014-07-21 aps4p5/drafts/draft_140719.tex
@@
 \title
-{An Efficient Parameter Space Search as an Alternative to Markov Chain Monte Carlo} 
+{Determining Frequentist Confidence Limits Using a Directed Parameter Space
+Search} 
 \author
@@
 
-\maketitle
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \begin{abstract} 
-We consider the problem of inferring constraints on a high-dimensional parameter
+We consider the problem of inferring constraints on a 
+high-dimensional parameter
 space with a computationally expensive likelihood function.
-Markov chain Monte Carlo (MCMC) methods offer significant improvements in
-efficiency over grid-based searches and are easy to implement in a wide range of
-cases.  However, MCMCs offer 
-few guarantees that all of the interesting regions of 
-parameter space are explored.
-We propose a
-machine learning algorithm that improves upon the performance of MCMC
-by intelligently targeting likelihood evaluations so as to quickly and
-accurately characterize the likelihood surface in both low- and high-likelihood
-regions.  
-We compare our algorithm to MCMC on toy examples and the 7-year WMAP
-cosmic microwave background data release.  
-Our algorithm finds comparable parameter
-constraints to MCMC comparably quickly and with greater
-certainty that all of the interesting regions of parameter space have been
-explored.
+We propose a machine learning algorithm that maps out the 
+Frequentist confidence limit on parameter space by intelligently targeting
+likelihood evaluations so as to quickly and accurately characterize the likelihood
+surface in both low- and high-likelihood regions.
+We compare our algorithm to Bayesian credible limits derived by 
+the well-tested Markov Chain Monte Carlo (MCMC)
+algorithms using both multi-modal toy likelihood functions and the 
+7-year WMAP cosmic microwave background likelihood function.  We find that our
+algorithm correctly identifies the location, general size, and general shape of
+high-likelihood regions in parameter space while being more robust against
+multi-modality than MCMC.
 \end{abstract}
+
+\maketitle
 
@@
 and defines the confidence limit as only those points which satisfy the rule.  
-Often, however, the calcuation $\vec{\theta}\rightarrow\chi^2$ is not rapid, or $N_p$
+Often, however, the calcuation $\vec{\theta}\rightarrow\chi^2$ is not rapid, 
+or $N_p$
 is large enough that such a grid-based method will take a prohibitively long
@@
 We propose an alternative solution, using methods from machine learning
-(to wit, Gaussian processes, though the algorithm admits many possible drivers; see
+(to wit, Gaussian processes, though the 
+algorithm admits many possible drivers; see
 Section \ref{sec:gp})
@@
 $\vec{\theta}$ (see Sections \ref{sec:bayes},\ref{sec:toy} and \ref{sec:wmap}).
-This rough approximation may principally be of use in cases where little 
-prior information about $\vec{\theta}$ exists and the researcher
-requires APS's robustness against multi-modality to be certain that she
-has discovered all of the best fits to her data.
+
+{\bf %marking new text
+Readers should be aware that, while MCMC can be analytically shown to
+asymptotically sample the Bayesian posterior, APS is designed to discover
+Frequentist confidence limits purely through exploration, i.e. there is no {\it a
+priori} motivation behind the APS algorithm.  Our only demonstration of APS'
+efficacy is empirical.  It has done well on the problems we have tested it on,
+though we have tried to choose those problems such that they are representative of
+problems likely to be encountered by the community.  Readers who require
+theoretically-motivated results will likely need to resort to a sampling routine
+like MCMC to derive their final constraints.  We hope to show, however, in Section
+\ref{sec:toy}, that even in these use cases, APS can provide a valuable service by
+identifying the general regions of parameter space over which the Bayesian
+posterior needs to be integrated, i.e. by guiding researchers in the process of setting
+priors for a sampling algorithm, a problem which has posed significant problems
+for MCMC in the case of multi-modal likelihood functions.
+} %marking new text
 
@@
 that point.  If
-$\chi^2 \neq \chi^2_\text{lim}$, bisection is again used and a new boundary point is found.
+$\chi^2 \neq \chi^2_\text{lim}$, use bisection to find a new boundary point.
 Store the resulting $\chi^2=\chi^2_\text{lim}$ point in $\{\hat{B}(\vec{M})\}$
@@
 perpendicularly to its ``radii'' (the vector $\vec{b}-\vec{M}$ would be a radius of
-$\chi^2=\chi^2_\text{lim}$ described a sphere), with the hopes of being able
+$\chi^2=\chi^2_\text{lim}$ described a sphere), with the hope of being able
 to extend the bounds of $\{\hat{B}_\text{MaxMin}(\vec{M})\}$.  In this way,
@@
 
-The final modification to the original APS algorithm which we propose in this
+The penultimate modification to the original APS algorithm which we propose in this
 work is to utilize function minimization to identify local minima in $\chi^2$.
@@
 that with our incorporation of bisection in sections \ref{sec:focus} and
-\ref{sec:unitSphere} above.  Our final innovation is to allow APS to use points
+\ref{sec:unitSphere} above.  An additional innovation is to allow APS to use points
 with 
@@
 
+\subsubsection{Simplex-driven APS search}
+\label{sec:simplex_strad}
+
+{\bf %marking new text
+
+Having introduced the concept of the Nelder-Mead (1965) simplex search in
+Section \ref{sec:simplex} above, we additionally use this tool to improve upon
+the original APS search algorith, steps (1A)-(3A).  Broadly speaking, the
+purpose of the original APS algorithm is to identify the point which maximizes
+the $S$ statistic in equation (\ref{eqn:sstat}) and evaluate
+$\vec{\theta}\rightarrow\chi^2$ at that point.  As more points are sampled, $S$
+will, presumably, come to be maximized at points that are near the
+$\chi^2=\chi^2_\text{lim}$ contour, but far from other sampled points, so APS
+will efficiently explore the parameter space.  Steps (1A)-(3A) approximately
+achieve this end by proposing a random sample of candidate points and evaluating
+$\vec{\theta}\rightarrow\chi^2$ at the point which maximizes $S$ within that
+sample.  We use the Nelder-Mead (1965) simplex to replace this process as follows:
+\begin{itemize}
+\item(1\~A) Perform step (1A) as above.
+\\
+\item(2\~A) Generate $N_p+1$ randomly chosen points $\{\hat{R}\}$ in parameter space.  
+For each point $\vec{r}$ in $\{\hat{R}\}$, use equation (\ref{eqn:sstat}) as in steps 
+(2A-3A) above to calculate $S$.  This set of $S$ values will be called
+$\{S\}$.
+\\
+\item(3\~A) Use $\{\hat{R}\}$ and $\{S\}$ as the seed for a
+Nelder-Mead (1965) simplex search.  At each subsequent point sampled by this
+search, calculate $S$ as above.  Run this simplex until $S$ converges to a local maximum.
+\\
+\item(4\~A) Evaluate $\vec{\theta}\rightarrow\chi^2$ at the point which
+maximized $S$.  Add this point to the list of points sampled by APS.  Return to
+step (2\~A).  Repeat until convergence.
+\end{itemize}
+By choosing the maximum $S$ point based on a simplex search rather than a
+finite random sample, we hope to find a true local maximum in $S$, rather than
+just the maximum relative to the limited number of sample points chosen.  Steps
+(1\~A)-(4\~A) replace steps (1A)-(3A) in our expanded APS algorithm.
+
+} %marking new text
+
 \subsection{Code implementing APS}
@@
 The code is written in C++.  It is designed for users to use
-with any $\vec{theta}\rightarrow\chi^2$ function.  Users need only
+with any $\vec{\theta}\rightarrow\chi^2$ function.  Users need only
 to write a sub-class of the \verb|chisquared| class defined in
@@
 methods (belonging to the class \verb|aps|) used to set them.
-This code includes several user-specified parameters which we
-outline below.
 
 \begin{itemize}
-\item The parameter $G$ from equation (\ref{eqn:threshold}) above can be set using the method
+\item The parameter $G$ from equation (\ref{eqn:threshold}) 
+above can be set using the method
 \verb|aps.set_grat|
 \\
-\item $\chi^2_\text{lim}$ can be manually set using the method \verb|aps.set_target|
+\item $\chi^2_\text{lim}$ can be manually set using the 
+method \verb|aps.set_target|
 \\
@@
 \\
-\item The number $N_g$ of nearest neighbors used in step 2A can be set in the constructor
+\item The number $N_g$ of nearest neighbors used in 
+steps (2\~A)-(3\~A) can be set in the constructor
 for the class \verb|aps|
-\\
-\item $N_c$ from step 2A can be set using the method \verb|aps.set_samples|
 \\
@@
 method \verb|aps.set_max| and \verb|aps.set_min|.  These bounds will constrain the range of
-points sampled in step 2A.  They will not contraint searches performed with bisection or the 
+points sampled in step (2\~A)-(3\~A).  
+They will not contrain searches performed with bisection or the 
 simplex minimizer.
@@
 \verb|aps.set_min|.  Setting a characteristic length scale will principally
-effect what points are considered nearest neighbors for the purposes of step 2A.
+effect what points are considered nearest neighbors for the purposes of steps
+(2\~A)-(3\~A).
 If the characteristic length scale of a parameter is large, that parameter will
@@
 \\
-\item(5D) $\vec{a}_\text{max}$ adn $\vec{a}_\text{min}$ should now describe an asymmetric,
+\item(5D) $\vec{a}_\text{max}$ and $\vec{a}_\text{min}$ should now describe an asymmetric,
 rectangular hyperbox about $\vec{a}$.  To symmetrize it, step through all $N_p$ dimensions
@@
 To integrate the posterior probability density, we will assume that all of the
-parameter space points in each hyperbox correspond have $\chi^2=\chi^2(\vec{a})$
-($\chi^2)\vec{a})$ being already known because $\vec{a}$ is one of the points for which APS evaluated
-$\vec{theta}\rightarrow\chi^2$).
+parameter space points in each hyperbox correspond have $\chi^2=\chi^2(\vec{a})$,
+{\bf %marking new text
+i.e. the $\chi^2$ discovered by APS for the point at the center of the hyperbox.
+} %marking new text
 The posterior probability associated with each box is then
@@
 \end{eqnarray}
-where $V(\vec{a})$ is the hypervolume of with the hyperbox surrounding $\vec{a}$.
+where $V(\vec{a})$ is the hypervolume of the hyperbox surrounding $\vec{a}$.
 To plot the $(1-\alpha)\%$ credible limit, one need only plot the points in hyperspace that
@@
 In  Section \ref{sec:toy}, we find that our results agree with the known credible limit.
-In Section \ref{sec:wmap}, we find that our results roughly approximate those derived from traditional
-Markov Chain Monte Carlo methods.
+{\bf %marking new text
+In Section \ref{sec:wmap} we find that APS correctly identifies the region and general
+shape of the Bayesian credible limits, however, with significant noise.  The Bayesian
+credible limits found by APS are systematically larger than the Bayesian credible
+limits found by MCMC.  This motivates our claim in the introduction that APS may be
+principally useful for identifying regions of high posterior density on largely unknown
+likelihood functions, allowing researchers to more shrewdly set thep priors for sampling
+algorithms like MCMC.
+}%marking nex text
 
@@
 
-Step (2A) of \APS requires that a prediction $\mu$ be made regarding the value
+Steps (2\~A)-(3\~A) of \APS requires that a prediction $\mu$ be made regarding the value
 of $\chi^2$ at points as yet unsampled by the algorithm.  An uncertainty $\sigma$
@@
 where $i$ indexes over all of the sampled points, to predict
-$\mu=\chi^{2,(q)}$ at the candidate point 
+$\mu=\chi^{2,(q)}$ at the unknown point 
 $\vec{\theta}^{(q)}$ by assuming that the function
@@
 $C_{ij}(\vec{\theta}^{(i)},\vec{\theta}^{(j)})$.
-If we have $N_g$ measurements of $\chi^2$, then we assume
+If we have $N_g$ measurements of $\chi^2$ 
+{\bf %marking nex text
+(recall from steps (2\~A-3\~A) above that, for
+each point where we must predict the value of $\chi^2$, we are only using the $N_g$
+nearest neighbor sampling points as data),
+}%marking new text
+ then we assume
 they are distributed according to
@@
 
-The normalization constant $K$ in equation (\ref{eqn:likelihood}) --known as the
+The normalization constant $K$ in equation (\ref{eqn:likelihood}) -- known as the
 ``Kriging parameter'' for the geophysicist who pioneered the overall method -- 
 also
-must be assumed.  Note that $K$ also multiplies the diagonal elements
-$C_{ii}$.  Determining the value of $K$ 
+must be assumed.  Determining the value of $K$ 
 is somewhat problematic because, examining equation
@@
 time elapsed since the algorithm began.  We divide this value by the number of calls made to
-the $\vec{\theta}\rightarrow\chi^2$ function and compare this amortized time to the average
-amount of time spent on one call to $\vec{\theta}\rightarrow\chi^2$.  APS us only allowed to
-reset its hyper-parameters if the difference between these two times is less than 0.1 second
+the $\vec{\theta}\rightarrow\chi^2$ function and 
+{\bf %marking new text
+compare this 
+average time per $\vec{\theta}\rightarrow\chi^2$ evaluation with all of the overhead
+imposed by the APS algorithm to the average
+amount of time spent on one call to $\vec{\theta}\rightarrow\chi^2$ without the overhead
+imposed by APS. 
+}%marking new text
+APS us only allowed to
+reset its hyper-parameters if the difference between these two times is less than 0.1 
+second per call to $\vec{\theta}\rightarrow\chi^2$
 (of course, we do require that APS optimize its hyper parameters at least once, so that
@@
 \chi^2(\vec{\theta})=
-\sum_i\left(\frac{\theta_i-c_{\text{nearest},i}}{w_{\text{nearest},i}}\right)^2
+\sum_i^{N_p=5}\left(\frac{\theta_i-c_{\text{nearest},i}}{w_{\text{nearest},i}}\right)^2
 \end{equation}
@@
 required 12,549 evaluations).  For comparison, we also ran 100 Markov Chain
-Monte Carlo searches on these $\chi^2$ function, each run consisting of five
+Monte Carlo searches on these $\chi^2$ functions, each run consisting of five
 independent chains.  We allowed the MCMCs to run
@@
 
+{\bf %marking new text
+This should not be a surprising statement.  In order to speed convergence,
+MCMC attempts to learn the size and shape of any high likelihood region it discovers.
+It uses what it learns to propose new random steps in parameter space so that it can
+efficiently converge to a state in which it is sampling the Bayesian posterior.  If
+an MCMC chain discovers one of several high likelihood regions, it will learn a proposal
+density that conforms to that high likelihood region, disregarding any other that may
+exist.  It will, in other words, become trapped in that high likelihood region.
+MCMC can overcome this difficulty by running several chains in parallel and hoping that,
+if multiple high likelihood regions exist, each chain will become trapped in different
+region and the whole MCMC will thus learn all of the modes in the parameter space.
+This, however, requires initializing a number of chains that is large relative to both
+the dimensionality of the parameter space and the number of discrete high likelihood
+regions it contains so that the odds of discovering all of the high likelihood regions
+are large.  If the user does not know how many high likelihood regions exist in
+the parameter space, this could be problematic.  Because steps (2\~A-4\~A) keep APS
+exploring widely through parameter space, even after a high likelihood region has been
+identified, APS can discover the presence of
+multiple high likelihood regions without requiring the user's prior knowledge of them.
+}%marking new text
+
 \begin{figure*}
@@
 with all of the noisy and ill-behaved properties that entails.
-To that end, we ill no consider for our $\vec{\theta}\rightarrow\chi^2$ 
+To that end, we will now consider for our $\vec{\theta}\rightarrow\chi^2$ 
 function the likelihood function on the
@@
 $\vec{\theta}\rightarrow\chi^2$ function in this case involves converting these
-paramters into the power spectrum of anisotropy $C_\ell$s on the sky and
+parameters into the power spectrum of anisotropy $C_\ell$s on the sky and
 comparing those $C_\ell$s to the actual measurements registered by the WMAP
@@
 with the true credible limits discovered by MCMC,
-comporting the the idea that, in the large data limit, Bayesiand credible limits
+comporting the the idea that, in the large data limit, Bayesian credible limits
 and Frequentist confidence limits ought to align.
@@
 by MCMC after only 50,000 $\chi^2$ evaluations.  These are the black regions in
-Figure \ref{fig:mcmc_comparison}.  As can be seen from Figures \ref{fig:mcmc_1_4},
-\ref{fig:mcmc_1_5}, and \ref{fig:mcmc_4_5}, MCMC has not yet totally converged
-in all of the sub-spaces of cosmological parameter space.  This should be compared
-with the reasonably coverage of the confidence limits seen in
-Figures \ref{fig:contours_gauss}-\ref{fig:contours_br}.
+Figure \ref{fig:mcmc_comparison}.  
+{\bf %marking new text
+While MCMC has clearly converged to the final
+confidence limits in some of the 2-dimensional sub-spaces shown,
+others (Figures \ref{fig:mcmc_1_4}, \ref{fig:mcmc_1_5}, and \ref{fig:mcmc_4_5})
+evince the presence of significant noise in the MCMC chains.
+Indeed, even the ``final'' credible limit contour (the red contour) in Figure
+\ref{fig:mcmc_1_4} shows significantly more noise along the edges than, for instance
+Figure \ref{fig:mcmc_0_1}.  This is meant, not necessarily to disparage the convergence
+properties of MCMC, but to show that any convergence advantage MCMC may have over APS is
+not absolute.  This should be compared with the clear advantage APS demonstrated over 
+MCMC in identifying multi-modal likelihood functions in Section \ref{sec:toy} above.
+} %marking new text
 
@@
 
+{\bf %marking new text
+We reiterate that the purpose of this section was to demonstrate that the presence of
+real-world noise in the data and likelihood function do not significantly impede APS's
+ability to determine confidence and credible limits.  The problem of drawing parameter
+constraints from the WMAP likelihood function is, by now, well understood, and shrewdly
+drawn priors and proposal densities are readily available.  
+Indeed, prior knowledge of early WMAP results allowed us to use the following 
+initial proposal densities for the MCMC runs displayed in this section:
+\begin{eqnarray}
+\Omega_\text{b}h^2&=&0.03\pm0.005\label{eqn:mcmcProposal}\\
+\Omega_\text{CDM}h^2&=&0.1\pm0.05\nonumber\\
+\theta&=&1.0\pm0.01\nonumber\\
+\tau&=&0.07\pm0.03\nonumber\\
+n_s&=&0.95\pm0.05\nonumber\\
+\ln\left[10^{10}A_s\right]&=&3.0\pm0.2\nonumber
+\end{eqnarray}
+where $\theta$ is 100 times the ratio of the sound horizon to the angular diameter
+distance and recombination (this is the input parameter CosmoMC uses in place of $h$).
+These distributions are fairly similar to the final, 
+marginalized, one-dimensional distributions of
+those same parameters
+\begin{eqnarray}
+\Omega_\text{b}h^2&=&0.0221\pm0.0006\nonumber\\
+\Omega_\text{CDM}h^2&=&0.112\pm0.006\nonumber\\
+\theta&=&1.039\pm0.003\nonumber\\
+\tau&=&0.05\pm0.03\nonumber\\
+n_s&=&0.96\pm0.02\nonumber\\
+\ln\left[10^{10}A_s\right]&=&3.12\pm0.07\nonumber\\
+h&=&0.7\pm0.03\nonumber
+\end{eqnarray}
+Conversely, for the purposes of our APS runs, we initialize the algorithm only with the
+gross assumptions that
+\begin{eqnarray}
+0.01\le&\Omega_\text{b}h^2&\le0.04\label{eqn:apsProposal}\\
+0.01\le&\Omega_\text{CDM}h^2&\le03\nonumber\\
+0.4\le&h&\le1.0\nonumber\\
+0.005\le&\tau&\le0.15\nonumber\\
+0.7\le&n_s&\le1.3\nonumber\\
+2.0\le&\ln\left[10^{10}A_s\right]&\le4.0\nonumber
+\end{eqnarray}
+In less well-studied problems, users may not understand their parameter spaces
+well enough to be able to supply shrewd proposal densities in the vein of equations 
+(\ref{eqn:mcmcProposal}).  In such a situation, the
+exploratory behavior of APS demonstrated in Section \ref{sec:toy} 
+may prove more important
+than the theoretical rigor offered by MCMC's sampling algorithm.
+}%marking new text
+
 \begin{figure*}
@@
 The 10 2-dimensional sub-spaces of our 6-dimensional parameter space
-($\tau$ is ignored because not real constraint can be gleaned from this data).
+($\tau$ is ignored because no real constraint can be gleaned from this data).
 The thick red contours are the 95\% Bayesian credible limits determined by MCMC after

### 2014-08-04 aps4p5/drafts/draft_current.tex
new draft up to date; just need to wait for the 1280.7 run to finish

--- 2014-08-04 aps4p5/drafts/draft_140804.tex
@@
 identifying the general regions of parameter space over which the Bayesian
-posterior needs to be integrated, i.e. by guiding researchers in the process of setting
-priors for a sampling algorithm, a problem which has posed significant problems
+posterior needs to be integrated, i.e. by guiding users in the process of setting
+priors and proposal densities for a sampling algorithm, 
+a problem which has posed significant problems
 for MCMC in the case of multi-modal likelihood functions.
@@
 principally useful for identifying regions of high posterior density on largely unknown
-likelihood functions, allowing researchers to more shrewdly set thep priors for sampling
+likelihood functions, allowing users to more shrewdly set priors for sampling
 algorithms like MCMC.
@@
 that APS might be useful in cases where little prior knowledge is known about $\vec{\theta}$
-and researchers need to be confident that they have discovered all of the low-$\chi^2$
+and users need to be confident that they have discovered all of the low-$\chi^2$
 regions in a particular parameter space.  Once APS has been used to identify rough
@@
 cartoon $\vec{\theta}\rightarrow\chi^2$ functions considered).
-In Section \ref{sec:wmap} we showed that on realistic data, APS
-is able to yield Frequentist confidence limits that are comparable to
-Bayesian credible limits with significantly fewer calls to $\vec{\theta}\rightarrow\chi^2$
-than MCMC (recall that APS required only 50,000 $\chi^2$ evaluations to yield Figures
-\ref{fig:contours_gauss}-\ref{fig:contours_br}).  However, the Bayesian credible
-limits yielded by APS remain rough and imprecise.  We believe that this performance
+{\bf %marking new text
+In Section \ref{sec:wmap}, we showed that the presence of realistic noise
+does not interfere with APS's ability to derive Frequentist confidence limits.
+}%marking new text
+However, the Bayesian credible
+limits yielded by APS remain noisy and imprecise.  We believe that this performance
 can still be of use to the community, especially when trying to evaluate proof-of-concept
@@
 with very expensive $\vec{\theta}\rightarrow\chi^2$ functions on poorly-understood
-parameter spaces.  Certainly, for researchers interested in plotting Frequentist confidence limits,
-APS is a viable option.
+parameter spaces.  Certainly, for users interested in plotting Frequentist confidence 
+limits, APS is a viable option.
 
